---
title: "gp3tools: An R Package for Reproducible Analysis and Reporting of Gazepoint GP3 Eye-Tracking Exports"
author:
  - name: "Stefanos Balaskas"
    affiliation: "University of Patras"
abstract: |
  Gazepoint GP3 and Gazepoint Analysis provide accessible eye-tracking workflows, but exported comma-separated value files often require substantial restructuring before they can support reproducible statistical analysis, quality auditing, manuscript reporting, and integration with the R ecosystem. `gp3tools` is an R package designed to import, validate, clean, summarise, model, visualise, and report Gazepoint GP3 and Gazepoint Analysis exports. The package supports all-gaze, fixation, pupil, area-of-interest, transition, time-course, quality-audit, and reporting workflows, with synthetic example data, vignettes, automated tests, continuous integration, and a reproducible software case study. This paper describes the package design, main functionality, validation strategy, and a synthetic demonstration workflow showing how `gp3tools` can convert raw Gazepoint-style exports into analysis-ready tables, diagnostic outputs, and reporting artefacts. The package is intended for researchers using Gazepoint systems in psychology, marketing, human-computer interaction, tourism, education, and related applied research contexts.
keywords: R package; eye tracking; Gazepoint; pupil data; area of interest; reproducible research; research software
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Introduction

Commercial and low-cost eye-tracking systems have increased access to gaze and pupil data in applied research. However, raw export files often require substantial preprocessing before they can support reproducible analysis. `gp3tools` addresses this gap for Gazepoint GP3 and Gazepoint Analysis exports by providing a structured R workflow for import, validation, preprocessing, summary, modelling, visualisation, and reporting.

The contribution of `gp3tools` is practical and reproducibility-oriented. It provides a bridge between Gazepoint export files and common R-based statistical workflows, while preserving transparent quality checks and reporting outputs. This paper introduces the package, describes its architecture, and demonstrates its use with synthetic Gazepoint-style data.

# Statement of need

Gazepoint exports contain rich gaze, fixation, pupil, media, participant, and area-of-interest information. In practice, researchers often need to combine multiple participant files, validate file-pairing, inspect sampling rates, audit tracking quality, handle missing pupil data, summarise area-of-interest behaviour, and prepare model-ready datasets. These steps are frequently repeated across projects and are vulnerable to inconsistent manual preprocessing.

`gp3tools` was developed to make these steps explicit, reusable, and inspectable. The package is especially useful when researchers need to move from raw Gazepoint-style CSV exports to analysis-ready tables, diagnostic summaries, reproducible reports, and manuscript-supporting outputs.

# Package design

`gp3tools` is organised around a sequential workflow:

1. import and file classification;
2. master-table construction and validation;
3. tracking-quality and sampling diagnostics;
4. pupil preprocessing and audit workflows;
5. area-of-interest summaries and model preparation;
6. transition, scanpath, and time-course preparation;
7. model-support helpers and sensitivity checks;
8. reporting, exclusion recommendation, and export utilities.

The package uses common R data structures and depends on widely used packages from the tidyverse ecosystem. Optional modelling and diagnostic extensions are kept in `Suggests` so that the core package remains installable without heavy dependencies.

# Core functionality

## Import and validation

The package imports Gazepoint-style all-gaze, fixation, and summary exports and provides checks for expected file-pairing, column availability, sampling behaviour, and tracking quality.

## Pupil workflows

Pupil workflows include baseline correction, interpolation, smoothing, artifact flagging, missingness auditing, reliability summaries, and model-ready window summaries.

## Area-of-interest workflows

Area-of-interest workflows include geometry checks, AOI coding audits, fixation/sample summaries, transition matrices, empirical-logit transformations, and GLMM/GAMM-oriented data preparation.

## Reporting and reproducibility

The package includes reporting helpers, readiness checks, exclusion recommendation tables, and export functions designed to support transparent manuscript reporting.

# Demonstration workflow

This section should be shortened from the full source RMD. The main manuscript should show only the essential workflow: import, validation, quality audit, pupil/AOI summaries, one model-ready output, and one reporting output. Full code and extended diagnostics should remain in the supplementary/source RMD.

```{r demo-installation, eval=FALSE}
# Development version
# remotes::install_github("stefanosbalaskas/gp3tools")

# After CRAN acceptance
# install.packages("gp3tools")
```

```{r demo-load, eval=FALSE}
library(gp3tools)
```

# Validation and software quality

`gp3tools` has been validated using automated tests, local CRAN-style checks, cross-platform GitHub Actions, and win-builder R-devel checks. The package includes synthetic example datasets and vignettes so that users can run workflows without private eye-tracking data.

For the released version described in this paper, the package passed local CRAN-style checks with no errors, warnings, or notes; GitHub Actions on Windows, macOS, Ubuntu release, and Ubuntu devel; and win-builder R-devel with only the expected first-submission/product-name note.

# Discussion

`gp3tools` supports reproducible Gazepoint GP3 analysis by converting repeated preprocessing and reporting tasks into explicit R functions. Its main value is not to replace general-purpose eye-tracking packages, but to provide a Gazepoint-specific bridge from vendor exports to validated, auditable, analysis-ready data structures.

# Limitations and future work

The package currently focuses on Gazepoint GP3 and Gazepoint Analysis export structures. Some interoperability helpers for other eye-tracking packages remain planned or optional. Future work may expand package adapters, documentation examples, and real-world case studies after broader public use.

# Availability

The package source code is available on GitHub, and the stable release has been archived through Zenodo. The current CRAN submission is under review.

# Supplementary material

The full source workflow is retained as supplementary material:

```{r supplement-path, echo=TRUE}
supplement_path <- "../supplement/gp3tools_full_showcase_supplement.Rmd"
supplement_path
```

# References

References will be added during journal-specific formatting. Required citations will include R, key package dependencies, Gazepoint-related materials where appropriate, and the archived software DOI.
