---
title: "gp3tools: Reproducible Analysis and Reporting of Gazepoint GP3 Eye-Tracking
  Exports"
author:
- name: Stefanos Balaskas
  affiliation: University of Patras, Department of Business Administration
abstract: "Eye-tracking studies increasingly require reproducible software pipelines
  that connect device exports with transparent preprocessing, quality control, statistical
  analysis, visualization, and reporting. Gazepoint GP3 and Gazepoint Analysis exports
  provide accessible gaze, fixation, pupil, timing, media, and area-of-interest data,
  but exported CSV folders usually require systematic validation before they can support
  defensible inference. `gp3tools` is an R package for importing, validating, preprocessing,
  modelling, visualising, and reporting Gazepoint GP3 / Gazepoint Analysis exports.
  The package supports file discovery, file-pair validation, sampling diagnostics,
  tracking-quality summaries, pupil preprocessing, area-of-interest summaries, fixation
  and transition workflows, model-ready data preparation, diagnostic plotting, exclusion
  recommendations, reporting checklists, and reproducible report generation. This
  paper presents the motivation, design, validation strategy, and synthetic demonstration
  workflow of `gp3tools`. The demonstration is used to evaluate workflow reproducibility
  and software behaviour; it is not used to draw substantive psychological, perceptual,
  or marketing conclusions. The package fills a practical gap between vendor-specific
  Gazepoint CSV exports and publication-ready R workflows for transparent eye-tracking
  analysis.\n"
output:
  word_document:
    toc: true
    toc_depth: '3'
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
bibliography: references.bib
link-citations: true
keywords: eye tracking; Gazepoint GP3; Gazepoint Analysis; R package; pupil preprocessing;
  area of interest; reproducible research; data quality; mixed models
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  dpi = 300
)
```

# Introduction

Eye-tracking research increasingly depends on reproducible software pipelines that connect raw device exports with transparent preprocessing, statistical analysis, visualization, and reporting. These pipelines matter because gaze, fixation, pupil, and area-of-interest (AOI) outcomes are not raw facts extracted directly from an eye tracker. They are constructed through device export formats, calibration quality, validity flags, timing fields, preprocessing rules, AOI definitions, trial windows, model specifications, and reporting decisions. Reproducibility therefore depends not only on the final statistical model but also on the traceability of the steps that precede it [@Wilson2014].

Gazepoint GP3 and Gazepoint Analysis are used in experimental and applied eye-tracking contexts because they provide accessible export workflows for gaze, fixation, pupil, timing, media, and AOI information [@GazepointGP3; @GazepointAPI]. However, exported CSV files are not automatically equivalent to analysis-ready data. A typical project may require checking whether all-gaze and fixation files are correctly paired, whether sampling rate is plausible, whether gaze and pupil validity are adequate, whether AOI labels and denominators are consistent, whether pupil artifacts and baseline decisions have been handled transparently, and whether the final outputs are sufficiently documented for reporting.

These requirements are not merely technical. Preprocessing choices can affect downstream interpretation. Pupil data are sensitive to blinks, artifacts, interpolation, and baseline correction [@KretSjakShie2019; @Mathot2018Baseline; @Mathot2018Pupillometry]. AOI analyses depend on how areas of interest are defined, interpreted, and reported [@Orquin2016]. More generally, research-software practices emphasize version control, testing, documentation, provenance, and citable software releases as part of reproducible computational work [@Wilson2014; @Smith2016; @Katz2021].

`gp3tools` was developed to make Gazepoint GP3 / Gazepoint Analysis workflows more explicit and reproducible in R [@RCoreTeam2026]. The package does not propose a new statistical estimator, eye-tracking theory, or universal preprocessing standard. Its contribution is a Gazepoint-specific workflow layer that converts vendor-style export folders into auditable R objects, diagnostic summaries, model-ready tables, plots, reports, and reporting-readiness outputs.

# Statement of need

Several tools already support eye-tracking and pupillometry workflows. `eyetrackingR` provides R-based tools for preparing, summarizing, visualizing, and analysing eye-tracking data [@DinkFerguson2015]. `PupillometryR` provides a pipeline for cleaning, preparing, plotting, and analysing pupillometry data [@Forbes2020]. General R packages support reproducible documents, statistical graphics, mixed models, generalized additive models, and generalized linear mixed models [@Wickham2016ggplot2; @Xie2015knitr; @XieAllaireGrolemund2018; @Bates2015; @Brooks2017; @Wood2017].

The need for `gp3tools` is therefore specific rather than generic. Existing tools are often platform-agnostic, signal-specific, paradigm-specific, or focused on later analysis stages. Gazepoint users need an export-aware workflow that starts from the concrete structure of Gazepoint GP3 / Gazepoint Analysis CSV folders and proceeds through validation, preprocessing, modelling, visualization, and reporting in the same R environment.

`gp3tools` addresses this gap by providing a reusable, documented, and testable workflow for Gazepoint exports. It is intended for researchers who need to process private Gazepoint recordings while sharing transparent code, synthetic demonstration data, diagnostic summaries, and reporting-ready outputs. Its intended users include applied eye-tracking researchers, behavioural scientists, marketing and consumer researchers, human-computer interaction researchers, usability laboratories, teaching laboratories, and methodologically oriented users who need auditable preprocessing before statistical modelling.

# Software contribution

`gp3tools` makes three software contributions. First, it provides a Gazepoint-specific import and validation layer for participant-level all-gaze, fixation, and summary exports. Second, it provides audit-oriented workflows for sampling, tracking quality, pupil preprocessing, AOI summaries, transition structures, modelling preparation, reporting readiness, and exclusion recommendations. Third, it provides public reproducibility material, including synthetic Gazepoint-style exports, R Markdown workflows, vignettes, tests, and a Zenodo-archived software release [@Balaskas2026gp3tools].

The package is best understood as workflow infrastructure. It supports the data-engineering, validation, and reporting stages that precede statistical inference. This distinction is important because many eye-tracking results depend on preprocessing and quality-control decisions that are often described only briefly in empirical manuscripts. By making these decisions explicit and reproducible, `gp3tools` helps researchers document how raw-style Gazepoint exports become analysis-ready datasets.

# Package design

`gp3tools` is organized around workflow families rather than isolated utilities. The first layer handles file discovery, import, and harmonization. It reads Gazepoint export folders, identifies all-gaze, fixation, and summary files, checks file pairing, and creates structured sample-level and fixation-level objects. This layer reduces the risk that later analyses are based on incomplete, duplicate, or mismatched files.

The second layer handles diagnostics and quality control. It estimates sampling rate, summarizes tracking quality, evaluates gaze and pupil validity, flags recordings requiring review, checks master-table structure, audits AOI coding, evaluates condition-level quality imbalance, and supports gaze recalibration when known target coordinates are available. These outputs are intended to be inspectable artifacts rather than hidden preprocessing steps.

The third layer prepares analysis-ready summaries. It supports AOI-window summaries, pupil-window summaries, fixation-level summaries, fixation-aligned data, event-aligned workflows, AOI transition tables, transition matrices, Markov-style objects, hidden-Markov-model-ready tables, semi-Markov-ready tables, time-course data, growth-curve-style data, and confirmatory model-ready objects for mixed-effects modelling.

The fourth layer supports modelling, robustness, reporting, and interoperability. It includes preparation functions or wrappers for linear mixed models, generalized linear mixed models, GAMM-style workflows, nested-model comparison, prediction plotting, residual-diagnostic workflows, leave-one-unit sensitivity, divergence-point estimation, multiverse-style preprocessing summaries, reporting checklists, exclusion recommendations, real-data readiness gates, manuscript-ready model tables, and adapters for related eye-tracking packages.

## Workflow map

Table 1 provides a compact map of the main workflow families. The full exported-function inventory remains in Supplementary Material S1.

```{r workflow-table, echo=FALSE}
workflow_families <- data.frame(
  `Workflow family` = c(
    "Import and file inspection",
    "Master-table validation",
    "Sampling and tracking quality",
    "Pupil preprocessing",
    "AOI workflows",
    "Model preparation and sensitivity",
    "Reporting and export"
  ),
  `Purpose` = c(
    "Read Gazepoint-style exports and identify expected file types.",
    "Create and inspect analysis-ready combined data.",
    "Check sampling behaviour, gaze validity, and review flags.",
    "Prepare pupil data for window and time-course analysis.",
    "Summarise and audit gaze behaviour by area of interest.",
    "Prepare model-ready datasets and evaluate robustness.",
    "Generate reporting checklists, exports, and reproducibility outputs."
  ),
  `Example outputs` = c(
    "classified files; imported all-gaze, fixation, and summary tables",
    "validated master tables; column diagnostics; duplicate-key checks",
    "sampling-rate summaries; tracking-quality tables; review flags",
    "baseline-corrected pupil series; interpolation flags; pupil-window summaries",
    "AOI sample and fixation summaries; transition matrices; denominator audits",
    "GLMM/GAMM input tables; leave-one-out summaries; divergence-point results",
    "reporting checklists; exclusion recommendations; exported CSV tables"
  ),
  check.names = FALSE
)
knitr::kable(workflow_families, caption = "Main workflow families supported by gp3tools.")
```

# Demonstration case study

The demonstration uses a synthetic Gazepoint-style export dataset. The purpose is to exercise the package under realistic but shareable conditions, not to test a substantive psychological or marketing hypothesis. The synthetic data include all-gaze files, fixation files, a summary export, media identifiers, AOI states, gaze coordinates, pupil values, validity indicators, missing pupil intervals, invalid samples, and participant-level quality variation.

The case study demonstrates that `gp3tools` can identify complete all-gaze/fixation file pairs, estimate sampling rate, flag review-required recordings, create analysis-ready all-gaze and fixation tables, summarize AOI viewing behaviour, audit AOI denominators, preprocess pupil data, run interpolation and baseline-quality checks, prepare confirmatory pupil-window and AOI-window model data, construct transition and sequence summaries, create exclusion recommendations, generate reporting-readiness objects, and produce diagnostic report outputs.

## Minimal reproducible workflow

The full executable workflow is provided in Supplementary Material S1. The main manuscript reports the minimal code structure used by a typical user.

```{r install-gp3tools, eval=FALSE}
# Development version
remotes::install_github("stefanosbalaskas/gp3tools")

# After CRAN acceptance
# install.packages("gp3tools")
```

```{r run-workflow, eval=FALSE}
library(gp3tools)

demo_dir <- system.file(
  "extdata",
  "gazepoint_realistic_demo_exports",
  package = "gp3tools"
)

workflow <- run_gazepoint_workflow(
  input_dir = demo_dir,
  output_dir = "gp3tools_demo_output",
  prefix = "gp3tools_demo",
  save_plots = TRUE,
  create_report = TRUE
)

summarise_gazepoint_workflow(workflow)
```

## Demonstration outputs

The larger paper-only synthetic showcase contains 48 synthetic all-gaze files, 48 synthetic fixation files, and one summary file. The end-to-end run produces 115,200 all-gaze rows, 5,691 fixation rows, 192 sampling-quality rows, 1,344 AOI summary rows, and 60 review-required rows. All 48 expected file pairs are complete. These values are software-demonstration outputs; they are not interpreted as empirical findings.

```{r demonstration-table, echo=FALSE}
demo_outputs <- data.frame(
  Output = c(
    "All-gaze rows",
    "Fixation rows",
    "Sampling-quality rows",
    "AOI summary rows",
    "Review-required rows",
    "Complete file pairs"
  ),
  Count = c(115200, 5691, 192, 1344, 60, 48),
  Interpretation = c(
    "sample-level gaze and pupil observations",
    "fixation-level observations",
    "participant-by-media sampling diagnostics",
    "area-of-interest summaries across trials/windows",
    "rows flagged for review rather than automatic deletion",
    "all expected all-gaze/fixation pairs found"
  )
)
knitr::kable(demo_outputs, caption = "Summary of synthetic demonstration outputs.")
```

```{r demonstration-plot, echo=FALSE, fig.cap="Scale of the main synthetic demonstration outputs. Counts are shown on a log10 scale because the outputs differ by several orders of magnitude."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot2::ggplot(demo_outputs, ggplot2::aes(x = reorder(Output, Count), y = Count)) +
    ggplot2::geom_col() +
    ggplot2::coord_flip() +
    ggplot2::scale_y_log10() +
    ggplot2::labs(
      x = NULL,
      y = "Count, log10 scale",
      title = "Synthetic demonstration output scale"
    ) +
    ggplot2::theme_minimal(base_size = 11)
}
```

# Validation and software quality

The validation strategy combines automated tests, reproducible examples, synthetic demonstration data, package checks, documentation, and release archiving. This strategy follows general principles of scientific computing, including readable code, version control, automated testing, scripted workflows, and preserved provenance [@Wilson2014]. It also aligns with software-citation and FAIR-oriented practices through public metadata, versioned releases, persistent citation information, and reusable documentation [@Smith2016; @Katz2021].

The package includes unit tests for core functions and workflow branches, following the standard R testing approach supported by `testthat` [@Wickham2011testthat]. The test suite checks import behaviour, validation outputs, diagnostic summaries, pupil preprocessing, AOI workflows, transition workflows, modelling preparation, reporting utilities, optional-dependency handling, and structured error/status reporting.

```{r validation-table, echo=FALSE}
validation_evidence <- data.frame(
  Evidence = c(
    "Automated tests",
    "Package checks",
    "Cross-platform CI",
    "Release archiving",
    "Documentation",
    "Synthetic demonstration"
  ),
  Role = c(
    "Checks core functions, edge cases, optional-dependency paths, and workflow objects.",
    "Confirms package installation, documentation, examples, vignettes, and R CMD check compliance.",
    "Evaluates package behaviour across major operating systems and R-devel/release settings.",
    "Provides a persistent versioned software citation through Zenodo.",
    "Provides README guidance, function help, vignettes, example data, and case-study workflows.",
    "Exercises the export-to-report workflow without exposing private participant data."
  )
)
knitr::kable(validation_evidence, caption = "Software-quality evidence reported for gp3tools.")
```

The software-quality claims in this paper are intentionally narrow. The demonstration shows that the package can execute a reproducible workflow on Gazepoint-like exports and produce expected diagnostic, analysis, and reporting objects. It does not prove that all possible Gazepoint exports will require no local inspection, nor does it validate any psychological theory. It also does not independently validate vendor hardware specifications such as nominal sampling rate or accuracy; those remain vendor specifications rather than package-level validation results.

# Data and code availability

The `gp3tools` source code is publicly available at `https://github.com/stefanosbalaskas/gp3tools`. Version 1.0.1 has been archived on Zenodo and should be cited using DOI `10.5281/zenodo.20731340` [@Balaskas2026gp3tools]. At the time of writing, the package has been submitted to CRAN and is awaiting review.

The repository includes source code, documentation, tests, example datasets, README workflow guidance, vignettes, citation metadata, and synthetic Gazepoint-style demonstration resources. The manuscript demonstration uses synthetic data so that the workflow can be reproduced publicly without redistributing private participant exports. Real empirical studies using private Gazepoint exports should report recruitment, ethics approval, calibration procedure, experimental design, preprocessing decisions, exclusion rules, and confirmatory model specifications.

# Limitations

The first limitation is scope. `gp3tools` is designed for Gazepoint GP3 / Gazepoint Analysis CSV exports and should not be presented as a universal solution for all eye-tracking platforms, export formats, or experimental paradigms. Other devices and software ecosystems may require different import layers, timing assumptions, coordinate systems, calibration diagnostics, or event structures.

The second limitation concerns synthetic demonstration data. Synthetic data make the workflow public and reproducible, but they cannot represent every real-world artifact pattern. Real Gazepoint projects may include calibration problems, participant movement, unusual timing structures, inconsistent AOI labels, missing files, duplicate files, different screen setups, stimulus-specific brightness differences, task interruptions, or nonstandard summary exports. The demonstration is therefore a reproducibility case study, not a comprehensive benchmark for all possible Gazepoint data problems.

The third limitation concerns statistical interpretation. `gp3tools` prepares model-ready data and provides representative workflows for LMMs, GLMMs, GAMMs, time-course models, AOI analyses, pupil-window analyses, transition summaries, and sensitivity checks. These functions do not determine the correct scientific model automatically. Researchers remain responsible for theoretical justification of AOIs and windows, model specification, assumption checking, multiplicity decisions, and the distinction between confirmatory and exploratory analyses.

Finally, some advanced workflows depend on optional packages or study-specific information. External-package checks require compatible installations and input schemas; stimulus-luminance auditing requires stimulus image files; gaze recalibration requires known target coordinates; and transition-count sensitivity requires appropriate count data and modelling dependencies. The package reports controlled statuses when optional components are unavailable, but users must decide which branches are appropriate for their study.

# Discussion

`gp3tools` addresses a practical reproducibility problem in Gazepoint-based eye-tracking research. Gazepoint exports provide rich CSV data, but researchers still need a transparent path from exported files to analysis-ready objects, diagnostic summaries, model-ready tables, figures, and reporting outputs. By making file validation, sampling diagnostics, tracking-quality summaries, AOI checks, pupil preprocessing, model preparation, and reporting readiness explicit, the package reduces reliance on undocumented manual steps.

The contribution is workflow integration. `gp3tools` does not introduce a new statistical model and does not claim to establish universal preprocessing standards for eye-tracking research. Instead, it provides Gazepoint-specific infrastructure that can be combined with existing R modelling packages and related eye-tracking tools. This design makes it useful for applied researchers who need to process private Gazepoint exports while sharing reproducible analysis code, synthetic demonstration data, and transparent diagnostic outputs.

The synthetic case study shows how a workflow can be reported without exposing private participant data. This is important because raw eye-tracking exports may contain sensitive participant-level quality patterns, proprietary stimulus information, or study-specific metadata. Public synthetic workflows allow package behaviour to be inspected while leaving substantive empirical inference to study-specific datasets.

Future development should focus on documentation, external validation, user-facing examples, and interoperability. Additional testing on independent Gazepoint datasets from different laboratories, tasks, stimulus types, screen configurations, and participant populations would strengthen confidence in edge-case handling. Further work could also expand export adapters, improve graphical summaries, and add examples for common applied research designs.

# Supplementary material

Supplementary Material S1 contains the full synthetic workflow, extended diagnostics, code, plots, model outputs, function inventory, and reproducibility information.

```{r supplement-path, echo=TRUE}
supplement_path <- "../supplement/gp3tools_full_showcase_supplement.Rmd"
supplement_path
```

# Conclusion

`gp3tools` provides a reproducible R workflow for Gazepoint GP3 / Gazepoint Analysis eye-tracking exports. It bridges the gap between vendor-specific CSV files and transparent statistical analysis by supporting import, validation, preprocessing, AOI summaries, pupil workflows, fixation and transition summaries, model-ready data preparation, diagnostic plotting, reporting checklists, and archived reproducibility materials. The package complements existing eye-tracking and pupillometry tools by focusing on Gazepoint-specific export structures and end-to-end workflow transparency. The synthetic demonstration validates package behaviour and workflow reproducibility, not substantive psychological effects. Used carefully, `gp3tools` can make Gazepoint-based eye-tracking analyses more auditable, reproducible, and publication-ready.

# References

