---
title: "Advanced static and dynamic AOI assignment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced static and dynamic AOI assignment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Scope

This article demonstrates base-R assignment of gaze coordinates to
non-rectangular and time-varying AOIs.

The helpers support:

- polygon AOIs defined by ordered vertices;
- rectangular or polygonal AOIs that change over time;
- participant- or trial-specific definitions;
- nearest, previous, or next definition-time matching;
- explicit overlap handling;
- coverage and definition-gap audits.

The resulting labels describe geometric membership only. They do not establish
attention, comprehension, or psychological interpretation without additional
design and measurement evidence.

```{r load-package}
library(gp3tools)
```

## Static polygon AOIs

Create two polygons: a triangular information region and a four-vertex action
region.

```{r static-definitions}
polygon_vertices <- rbind(
  data.frame(
    aoi_name = "information",
    vertex_order = 1:3,
    vertex_x = c(0.05, 0.45, 0.05),
    vertex_y = c(0.10, 0.30, 0.50)
  ),
  data.frame(
    aoi_name = "action",
    vertex_order = 1:4,
    vertex_x = c(0.60, 0.90, 0.90, 0.60),
    vertex_y = c(0.55, 0.55, 0.85, 0.85)
  )
)

polygon_vertices
```

Synthetic gaze samples span both AOIs and the surrounding display.

```{r static-samples}
set.seed(20260715)

static_gaze <- data.frame(
  sample = seq_len(120),
  x = stats::runif(120),
  y = stats::runif(120),
  stringsAsFactors = FALSE
)

static_assigned <- add_gazepoint_polygon_aoi(
  master_df = static_gaze,
  vertices = polygon_vertices,
  x_col = "x",
  y_col = "y",
  vertex_order_col = "vertex_order",
  output = "both",
  label_col = "aoi_label"
)

table(static_assigned$aoi_label)
```

```{r static-polygon-plot}
plot(
  static_gaze$x,
  static_gaze$y,
  asp = 1,
  xlim = c(0, 1),
  ylim = c(1, 0),
  xlab = "Normalized x",
  ylab = "Normalized y",
  main = "Synthetic gaze and polygon AOIs",
  pch = 19,
  cex = 0.7
)

polygon(
  polygon_vertices$vertex_x[
    polygon_vertices$aoi_name == "information"
  ],
  polygon_vertices$vertex_y[
    polygon_vertices$aoi_name == "information"
  ],
  border = 2,
  lwd = 2
)

polygon(
  polygon_vertices$vertex_x[
    polygon_vertices$aoi_name == "action"
  ],
  polygon_vertices$vertex_y[
    polygon_vertices$aoi_name == "action"
  ],
  border = 3,
  lwd = 2
)

legend(
  "topright",
  legend = c("Information polygon", "Action polygon"),
  lty = 1,
  lwd = 2,
  col = c(2, 3),
  bty = "n"
)
```

Boundary points can be counted as inside or outside. When polygons overlap,
`overlap = "first"`, `"last"`, or `"error"` makes the assignment rule explicit.

## Time-varying rectangular AOIs

The following target moves horizontally across a synthetic display. Definitions
are available every 250 ms.

```{r dynamic-definitions}
definition_times <- seq(0, 2000, by = 250)
target_left <- 0.10 + 0.00025 * definition_times

dynamic_rectangles <- data.frame(
  trial = "trial_01",
  aoi_time = definition_times,
  aoi_name = "moving_target",
  left = target_left,
  right = target_left + 0.20,
  top = 0.35,
  bottom = 0.65,
  stringsAsFactors = FALSE
)

utils::head(dynamic_rectangles)
```

Synthetic gaze follows the target with small coordinate noise and includes
several deliberately displaced samples.

```{r dynamic-samples}
sample_times <- seq(0, 2000, by = 20)
true_left <- 0.10 + 0.00025 * sample_times

dynamic_gaze <- data.frame(
  trial = "trial_01",
  time = sample_times,
  x = true_left + 0.10 + stats::rnorm(
    length(sample_times),
    sd = 0.035
  ),
  y = 0.50 + stats::rnorm(
    length(sample_times),
    sd = 0.035
  ),
  stringsAsFactors = FALSE
)

dynamic_gaze$x[c(20, 60, 90)] <-
  dynamic_gaze$x[c(20, 60, 90)] + 0.35
```

Match each gaze sample to the nearest available AOI definition.

```{r dynamic-assignment}
dynamic_assigned <- add_gazepoint_dynamic_aoi(
  master_df = dynamic_gaze,
  aoi_defs = dynamic_rectangles,
  x_col = "x",
  y_col = "y",
  time_col = "time",
  group_cols = "trial",
  match = "nearest",
  max_time_gap = 150,
  output = "both",
  label_col = "aoi_label"
)

table(dynamic_assigned$aoi_label, useNA = "ifany")
```

```{r dynamic-membership-plot}
plot(
  dynamic_assigned$time,
  dynamic_assigned$x,
  type = "l",
  xlab = "Time (ms)",
  ylab = "Normalized x",
  main = "Gaze relative to a moving rectangular AOI"
)

lines(
  dynamic_rectangles$aoi_time,
  dynamic_rectangles$left,
  lty = 2
)

lines(
  dynamic_rectangles$aoi_time,
  dynamic_rectangles$right,
  lty = 2
)

points(
  dynamic_assigned$time[
    dynamic_assigned$aoi_label == "moving_target"
  ],
  dynamic_assigned$x[
    dynamic_assigned$aoi_label == "moving_target"
  ],
  pch = 19,
  cex = 0.55
)

legend(
  "topleft",
  legend = c("Gaze x", "AOI bounds", "Inside matched AOI"),
  lty = c(1, 2, NA),
  pch = c(NA, NA, 19),
  bty = "n"
)
```

`match = "previous"` is useful when each definition remains valid until the next
update. `match = "next"` supports designs where definitions describe the next
known frame. `max_time_gap` prevents distant definitions from being applied
silently.

## Dynamic polygon AOIs

Dynamic polygons use repeated vertex rows at each definition time.

```{r dynamic-polygon}
dynamic_polygons <- do.call(
  rbind,
  lapply(
    c(0, 1000),
    function(definition_time) {
      offset <- definition_time / 2000

      data.frame(
        trial = "trial_01",
        aoi_time = definition_time,
        aoi_name = "moving_polygon",
        vertex_order = 1:4,
        vertex_x = offset + c(0, 0.25, 0.25, 0),
        vertex_y = c(0.10, 0.10, 0.35, 0.35),
        stringsAsFactors = FALSE
      )
    }
  )
)

dynamic_polygons
```

The same `add_gazepoint_dynamic_aoi()` interface is used with
`shape = "polygon"` and the vertex columns.

## Coverage audit

Audit whether samples had a usable definition, how large matching gaps were,
and whether coordinates fell inside or outside all AOIs.

```{r dynamic-audit}
coverage_audit <- audit_gazepoint_dynamic_aoi_coverage(
  dynamic_assigned,
  label_col = "aoi_label",
  group_cols = "trial",
  max_time_gap = 100,
  x_col = "x",
  y_col = "y"
)

coverage_audit$overview
coverage_audit$group_summary
coverage_audit$aoi_summary
utils::head(coverage_audit$flagged_rows)
```

A `"review"` status identifies missing gaze, unavailable definitions, excessive
definition-time gaps, or samples outside all AOIs. Outside-AOI samples are not
automatically invalid; their relevance depends on the stimulus and analysis
plan.

## Recommended reporting

Report:

- coordinate system and screen normalization;
- polygon vertex ordering or rectangle bounds;
- AOI-definition update frequency;
- participant/trial grouping;
- definition-time matching rule;
- maximum permitted matching gap;
- boundary and overlap rules;
- percentage of samples with definitions;
- inside-, outside-, and missing-gaze proportions.

These details make dynamic AOI assignment auditable and reproducible.
