---
title: "AOI Labelling, Sliding Windows, and Fixation Simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{AOI Labelling, Sliding Windows, and Fixation Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4.5
)
library(gp3tools)
library(ggplot2)
```

## Purpose

This article combines three reusable feature-engineering helpers:

1. simulate Gazepoint-like fixation events;
2. add rectangular AOI membership to sample-level coordinates;
3. summarise gaze and pupil measures in sliding time windows.

## Simulate fixation events

```{r fixation-simulation}
fixations <- simulate_gazepoint_fixations(
  n_subjects = 4,
  n_fix = 30,
  coordinate_system = "pixels",
  screen_width = 1280,
  screen_height = 720,
  sd = 45,
  seed = 2026
)

fixations
```

```{r fixation-plot}
ggplot(fixations, aes(x, y, group = USER_ID)) +
  geom_path(alpha = 0.5) +
  geom_point(aes(size = duration_ms), alpha = 0.7) +
  scale_y_reverse() +
  facet_wrap(~ USER_ID) +
  coord_fixed() +
  labs(
    x = "Screen x (pixels)",
    y = "Screen y (pixels)",
    size = "Duration (ms)",
    title = "Simulated fixation paths"
  ) +
  theme_minimal()
```

## Create a sample-level trace

```{r sample-trace}
set.seed(2026)

n <- 500
samples <- data.frame(
  USER_ID = rep(c("P01", "P02"), each = n / 2),
  trial = rep(c("T01", "T02"), each = n / 2),
  TIME = rep(seq(0, 2.49, by = 0.01), 2),
  FPOGX = c(
    seq(0.15, 0.85, length.out = n / 2),
    seq(0.85, 0.15, length.out = n / 2)
  ) + rnorm(n, 0, 0.015),
  FPOGY = 0.50 + rnorm(n, 0, 0.06),
  mean_pupil = 3.3 + rnorm(n, 0, 0.05)
)

aoi_defs <- data.frame(
  name = c("left", "right"),
  L = c(0.00, 0.55),
  R = c(0.45, 1.00),
  T = c(0.20, 0.20),
  B = c(0.80, 0.80)
)
```

## Add AOI membership

```{r aoi}
labelled <- add_gazepoint_aoi(
  samples,
  aoi_defs,
  output = "both",
  overlap = "error"
)

table(labelled$aoi_current, useNA = "ifany")
```

```{r aoi-plot}
ggplot(labelled, aes(FPOGX, FPOGY, shape = aoi_current)) +
  geom_point(alpha = 0.45) +
  scale_y_reverse() +
  facet_wrap(~ USER_ID) +
  coord_fixed() +
  labs(
    x = "Normalized x",
    y = "Normalized y",
    shape = "AOI",
    title = "Rectangular AOI classification"
  ) +
  theme_minimal()
```

## Sliding-window summaries

```{r windows}
labelled$right_aoi_numeric <- as.numeric(labelled$aoi_right)

windows <- analyze_gazepoint_window(
  labelled,
  window_size = 250,
  step = 100,
  summary_stats = c("mean", "sd", "valid_prop"),
  by = c("USER_ID", "trial"),
  value_cols = c("mean_pupil", "right_aoi_numeric")
)

windows
```

```{r window-plot}
ggplot(
  windows,
  aes(window_mid, right_aoi_numeric_mean, group = USER_ID)
) +
  geom_line() +
  geom_point() +
  facet_wrap(~ USER_ID) +
  labs(
    x = "Window midpoint (s)",
    y = "Proportion in right AOI",
    title = "Sliding-window AOI summaries"
  ) +
  theme_minimal()
```

## Recommended reporting

Report AOI coordinate units and boundaries, boundary inclusion rules, overlap
handling, the window width and step, whether windows overlap, the timestamp
unit, and the summary statistics calculated within each window.
