---
title: "Gazepoint AOI Perturbation and Uncertainty"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gazepoint AOI Perturbation and Uncertainty}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(gp3tools)
has_aoi_core <- is.function(tryCatch(getExportedValue("eyeprocess", "run_aoi_sensitivity_analysis"), error = function(e) NULL))
```

## Architecture

This workflow is deliberately thin. `gp3tools` handles Gazepoint-specific column names and coordinate units; `eyeprocess` owns AOI perturbation, assignment stability, feature recomputation, statistical propagation, and visualization. The wrapper does not create a second scientific implementation.

## When to use

Use the adapter when Gazepoint samples or fixations must be tested against small, defensible AOI-boundary changes. Do not use it to search for a favorable AOI or to replace calibration, screen-coordinate, or tracking-quality checks.

## Synthetic normalized Gazepoint-style data

```{r synthetic}
set.seed(20260918)
aois <- data.frame(
  aoi_id = c("headline", "image", "claim", "disclosure", "cta"),
  xmin = c(.10, .30, .30, .30, .68), xmax = c(.40, .70, .70, .70, .92),
  ymin = c(.08, .19, .36, .50, .60), ymax = c(.18, .34, .48, .59, .71)
)
gaze <- data.frame(
  sample_id = 1:80, participant = rep(paste0("p", 1:8), each = 10),
  trial = rep(rep(1:2, each = 5), 8),
  FPOGX = pmin(1, pmax(0, runif(80, .08, .94))),
  FPOGY = pmin(1, pmax(0, runif(80, .06, .74))),
  duration = runif(80, .06, .18)
)
```

## Audit the adapter boundary

Normalized coordinates are not treated as pixels implicitly. Screen dimensions must be supplied before conversion.

```{r audit, eval=has_aoi_core}
audit <- audit_gazepoint_aoi_uncertainty(
  gaze, aois, gaze_x_col = "FPOGX", gaze_y_col = "FPOGY",
  coordinate_unit = "normalized", screen_width_px = 1920, screen_height_px = 1080
)
audit$overview
```

## Run the vendor-neutral sensitivity workflow

```{r run, eval=has_aoi_core}
result <- run_gazepoint_aoi_sensitivity(
  gaze, aois, gaze_x_col = "FPOGX", gaze_y_col = "FPOGY",
  coordinate_unit = "normalized", perturbation_unit = "deg",
  dilations = c(.25, .50, 1.00), erosions = .25,
  translations_x = .50, translations_y = .50,
  translations_xy = list(c(.25, -.25)),
  screen_width_px = 1920, screen_height_px = 1080,
  viewing_distance = 60, physical_screen_size = c(53.1, 29.9),
  observation_id_col = "sample_id", participant_col = "participant",
  trial_col = "trial", duration_col = "duration",
  observation_level = "sample",
  overlap_policy = "ambiguous", quality_rules = list(missing = "preserve")
)
result$core_result$stability$overall
```

`translations_xy` is forwarded unchanged to the vendor-neutral core; `gp3tools` does not implement a separate translation engine. The complete `eyeprocess` result remains unchanged inside `result$core_result`; the outer object records Gazepoint source columns and unit-conversion settings.

## Visual output

```{r plot, eval=has_aoi_core}
plot_gazepoint_aoi_sensitivity(result, "assignment")
```

Other delegated plot families are `perturbations`, `coefficient`, and `surface`.

## Gazepoint AOI pre-run checklist

For a fillable preregistration-style template that records source columns, coordinate conversion, perturbation choices, observation level, failure handling, and reporting fields, see the companion **Gazepoint AOI Sensitivity Analysis Plan** article.


Before calling the adapter, confirm:

1. the exact Gazepoint x/y source columns;
2. whether coordinates are normalized or pixels;
3. screen width/height for normalized-to-pixel conversion;
4. physical screen size and viewing distance for degree-based perturbations;
5. whether rows are samples or fixations;
6. the perturbation grid, overlap policy, and boundary policy;
7. the fixed downstream model specification in the delegated `eyeprocess` callback.

The adapter never infers these scientific choices from the desired result.

## Gazepoint AOI troubleshooting

Use `audit_gazepoint_aoi_uncertainty()` first when units or source columns are uncertain. Once the adapter boundary is valid, troubleshoot the delegated core result in this order: geometry, assignments, feature denominators, then model callback.

| Symptom | First check | Do not do |
|---|---|---|
| Normalized coordinates rejected | screen dimensions and `coordinate_unit` | Treat 0-1 values as pixels |
| Geometry branch fails | `result$core_result$grid_result$audit` | Fall back silently to the older margin-sensitivity path |
| Many ambiguous assignments | perturbed layout and core reassignment matrix | Resolve overlap by AOI order |
| Sample count is zero | valid sample denominator | Recode missing gaze as zero |
| Sample count is missing | missing source coordinates | Treat the trial as observed non-inspection |
| Model callback fails/non-converges | core failures/models tables | Report it as a valid null effect |
| Model `N` changes | core inference stability | Report only coefficient direction |

A reporting sentence should identify the original Gazepoint columns and coordinate unit before describing AOI robustness, because otherwise the perturbation scale is not reproducible.

### Adapter/core API links

- `audit_gazepoint_aoi_uncertainty()` validates the Gazepoint-to-core boundary.
- `run_gazepoint_aoi_sensitivity()` delegates the complete workflow to `eyeprocess`.
- `plot_gazepoint_aoi_sensitivity()` delegates all four plot families.
- Use the `eyeprocess` AOI troubleshooting guidance for geometry, denominator, convergence, and model-N interpretation.

## Gazepoint interpretation clinic

Use the adapter only after source coordinates and screen geometry are explicit. For normalized Gazepoint samples, declare `observation_level = "sample"`; for fixation exports or fixation centroids, declare `"fixation"`.

| Pattern | Interpretation | Action |
|---|---|---|
| Assignments and model stable | Stable within the declared AOI envelope | Report both layers and the screen/viewing geometry |
| Assignments unstable, model stable | Geometry changes mapping without materially changing the model result | Inspect reassignment matrices and AOI-level stability |
| Model unstable | Substantive inference depends on geometry and/or branch-specific case loss | Inspect coefficient intervals, convergence, and model `N` |
| Adapter/core failure | The branch is non-evaluable | Keep the failure visible; do not fall back to the older margin-sensitivity implementation |

At sample level, a zero count means valid gaze samples existed but none landed in that AOI. If all coordinates for a participant/trial are missing, the count remains missing rather than becoming zero.

### Reporting example

> Gazepoint normalized coordinates were converted using the declared screen dimensions, and AOI sensitivity was evaluated in degrees of visual angle. Sample-level assignments were recomputed across the nominal and perturbed geometries using explicit ambiguity handling. We report assignment stability, coefficient ranges/intervals, convergence, and model-N variation; failed branches were retained in the audit trail.

## Interpretation and limitations

A high unchanged-assignment proportion means the result is stable only with respect to the declared perturbations. It is not a probability that an AOI or scientific conclusion is true. Overlap remains explicit ambiguity by default; missing coordinates remain missing; failed branches and non-converged models remain in the core audit trail. When a model callback is supplied, the core also reports the range of model `N` across converged perturbations so case loss is not hidden.

## Reporting

Report the original Gazepoint coordinate columns, source coordinate unit, observation level, screen dimensions used for normalized conversion, viewing geometry for degree-based margins, perturbation values, overlap/screen policies, zero-versus-missing denominator rules, recomputed features, fixed statistical model specification, coefficient/interval ranges, convergence, model-`N` variation, and failed/non-converged branches.

If `eyeprocess` is unavailable, the adapter fails directly rather than silently falling back to the older gp3tools margin-sensitivity implementation.

## Adapter API map

Use `audit_gazepoint_aoi_uncertainty()` to validate the Gazepoint-to-core boundary, `run_gazepoint_aoi_sensitivity()` to delegate the full analysis to `eyeprocess`, and `plot_gazepoint_aoi_sensitivity()` for the delegated geometry, assignment, coefficient, and robustness-surface diagnostics.
