---
title: "Auditable Binocular Pupil Reconstruction and Sensitivity Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Auditable Binocular Pupil Reconstruction and Sensitivity Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4.5
)
library(gp3tools)
library(ggplot2)
```

## Why a separate reconstruction workflow?

Binocular pupil channels are related but not interchangeable. A tracker may temporarily
retain one eye while the other eye is unavailable, and a simple sample-wise mean can
therefore change its measurement basis from two eyes to one without making that change
visible in the analysis table. Conversely, filling a missing eye with a prediction can
increase continuity while introducing model error.

`gp3tools` already provides lightweight binocular fusion with
`combine_gazepoint_eyes()`, a regression-oriented fusion helper with
`regress_gazepoint_pupils()`, pupil artifact detection, PCHIP interpolation, baseline
correction, preprocessing audits, and sensitivity workflows. The functions in this
article add a stricter layer for **model-based missing-eye reconstruction**. Their design
principle is simple: a predicted pupil value must never become indistinguishable from an
observed pupil value.

The workflow was motivated in part by Ong et al. (2025), who reconstructed a missing
left or right pupil value from the other eye with linear regression when estimating a
mean pupil series. Their paper also makes clear that regression can introduce bias when
pupil variation is rapid or irregular. `gp3tools` therefore treats cross-eye
reconstruction as a declared policy that should be validated and sensitivity-checked,
not as a universally correct replacement for missing measurements.

## What reconstruction does and does not mean

The workflow can estimate a temporarily unavailable pupil channel from the
simultaneously observed contralateral channel. It can answer questions such as:

- how much bilateral information is actually available;
- whether a participant/session has enough paired samples for calibration;
- how accurately an eye can be predicted when its true value is deliberately hidden;
- how much of a final analysis series depends on predicted values;
- whether reconstruction is concentrated in one experimental condition;
- whether conclusions are sensitive to using complete cases, available-eye values, or
  regression-supported reconstruction.

It does **not** establish that a predicted value is the unobserved biological truth, that
naturally missing values are missing at random, or that a high cross-eye correlation is
sufficient evidence for unrestricted imputation.

## Synthetic example

The example is generated entirely in R and is not an empirical Gazepoint recording.
We start with `simulate_gazepoint_pupil_data()` and introduce short monocular losses
while retaining the generator's bilateral blink losses.

```{r data}
pupil <- simulate_gazepoint_pupil_data(
  n_subjects = 6,
  n_trials = 4,
  n_time_bins = 120,
  seed = 20260817
)

# Add a small deterministic between-eye calibration offset.
pupil$pupil_right <- 0.08 + 0.98 * pupil$pupil_right

# Introduce synthetic monocular losses. These are deliberately not labelled as blinks.
left_loss <- c(80:84, 510:515, 1320:1324, 2110:2116)
right_loss <- c(260:264, 890:894, 1690:1694, 2480:2485)
pupil$pupil_left[left_loss] <- NA_real_
pupil$pupil_right[right_loss] <- NA_real_

pupil[75:90, c(
  "subject", "trial", "condition", "timestamp_ms",
  "pupil_left", "pupil_right", "blink"
)]
```

## Recommended position in the preprocessing workflow

Cross-eye reconstruction and temporal interpolation address different missing-data
situations. A defensible order for a study that chooses to use reconstruction is:

1. preserve raw eye-specific channels;
2. identify invalid measurements, blinks, track loss, and other artifacts;
3. create eye-specific cleaned channels;
4. diagnose bilateral availability and agreement;
5. fit and validate cross-eye calibration on directly observed bilateral samples;
6. reconstruct eligible monocular gaps, if the study has declared that policy;
7. construct the combined pupil signal under an explicit combination policy;
8. if required, interpolate remaining *temporal* gaps under a separately declared rule;
9. baseline-correct, smooth, summarise, and model;
10. report reconstruction burden and run sensitivity analyses.

This order is not a claim that every study should reconstruct. In many datasets the
appropriate result of the diagnostic stage will be to retain available-eye or
complete-case processing instead.

## 1. Diagnose before reconstructing

```{r diagnose}
diagnostics <- diagnose_gazepoint_binocular_pupil(
  pupil,
  left_col = "pupil_left",
  right_col = "pupil_right",
  time_col = "timestamp_ms",
  group_cols = c("subject", "trial"),
  time_unit = "milliseconds",
  valid_min = 1,
  valid_max = 9,
  min_pairs = 40
)

diagnostics$summary[c(
  "subject", "trial", "n", "n_bilateral", "n_left_only",
  "n_right_only", "prop_bilateral", "correlation",
  "rmse_between_eyes", "calibration_eligible"
)]
```

The diagnostic object also retains missing-eye run information.

```{r diagnose-gaps}
head(diagnostics$gaps)
```

Agreement is best inspected rather than reduced to one correlation coefficient.

```{r agreement-plot, fig.cap="Observed left/right pupil agreement. The dashed identity line is distinct from the fitted descriptive relationship."}
plot_gazepoint_binocular_diagnostics(
  pupil,
  type = "agreement",
  left_col = "pupil_left",
  right_col = "pupil_right"
)
```

```{r bland-altman, fig.cap="Descriptive left-minus-right agreement across the observed pupil range."}
plot_gazepoint_binocular_diagnostics(
  pupil,
  type = "bland_altman",
  left_col = "pupil_left",
  right_col = "pupil_right"
)
```

The difference plot is descriptive. Its mean difference and +/-1.96 SD lines should not
be interpreted as proof that the two eye channels are interchangeable.

## 2. Fit auditable bidirectional calibration

The calibration stage fits separate left-from-right and right-from-left models. Here the
primary models are participant-specific and a pooled fallback is retained. Fallback is
explicit in the returned model table.

```{r calibration}
calibration <- fit_gazepoint_binocular_calibration(
  pupil,
  left_col = "pupil_left",
  right_col = "pupil_right",
  group_cols = "subject",
  # Default fallback is pooled when primary groups are supplied.
  min_pairs = 80,
  min_unique = 10,
  min_r2 = NULL,
  valid_min = 1,
  valid_max = 9
)

calibration$models[c(
  "model_id", "direction", "calibration_level", "n_pairs",
  "intercept", "slope", "r_squared", "rmse", "mae",
  "eligible", "reason"
)]
```

`min_r2 = NULL` is intentional. R-squared is reported, but no universal R-squared
threshold is presented as a validity rule. A study can declare a threshold if it has a
substantive reason to do so.

## 3. Validate by hiding values that are actually known

Bilateral observations provide an unusually useful validation opportunity: one eye can
be deliberately hidden and reconstructed from the other while its true observed value is
retained outside the training data for error calculation.

```{r validation}
validation <- validate_gazepoint_binocular_reconstruction(
  pupil,
  left_col = "pupil_left",
  right_col = "pupil_right",
  time_col = "timestamp_ms",
  group_cols = "subject",
  gap_group_cols = c("subject", "trial"),
  direction = "both",
  mask_prop = 0.10,
  mask_mode = "contiguous",
  block_size = 5,
  repeats = 3,
  seed = 1729,
  min_pairs = 80,
  valid_min = 1,
  valid_max = 9
)

validation$summary
```

The result reports the fraction of requested held-out samples that could actually be
reconstructed, together with RMSE, MAE, bias, and correlation.

```{r validation-plot, fig.cap="Observed versus reconstructed pupil values under artificial monocular loss."}
plot_gazepoint_binocular_diagnostics(validation, type = "validation")
```

```{r validation-residuals, fig.cap="Prediction error as a function of reconstructed pupil diameter."}
plot_gazepoint_binocular_diagnostics(validation, type = "residuals")
```

A good artificial-missingness result is evidence that the declared model predicts held-out
bilateral samples well. It cannot show that naturally missing samples follow the same
missingness process.

## 4. Reconstruct under explicit gates

The following policy allows regression reconstruction only for missing-eye runs no longer
than 150 ms, disallows predictor extrapolation, and excludes any row that is already
marked as a blink.

```{r reconstruct}
reconstructed <- reconstruct_gazepoint_binocular_pupil(
  pupil,
  left_col = "pupil_left",
  right_col = "pupil_right",
  time_col = "timestamp_ms",
  group_cols = "subject",
  gap_group_cols = c("subject", "trial"),
  method = "linear_regression",
  calibration = calibration,
  time_unit = "milliseconds",
  max_gap_ms = 150,
  allow_edge_gaps = TRUE,
  allow_extrapolation = FALSE,
  valid_min = 1,
  valid_max = 9,
  exclude_flag_cols = "blink"
)

table(reconstructed$gp3_binocular_status)
```

The original columns are untouched. The added columns distinguish the observed and final
channels, prediction direction, model identity, calibration level, model R-squared,
extrapolation status, gap duration, and row-level outcome.

```{r provenance-columns}
reconstructed[75:90, c(
  "pupil_left", "pupil_right",
  "gp3_binocular_left_observed", "gp3_binocular_right_observed",
  "gp3_binocular_left_final", "gp3_binocular_right_final",
  "gp3_binocular_reconstructed", "gp3_binocular_model_id",
  "gp3_binocular_gap_ms", "gp3_binocular_status"
)]
```

## 5. Construct a combined signal as a separate decision

Reconstruction does not silently create the outcome variable. The analyst chooses how a
combined pupil signal should be constructed.

```{r construct}
combined <- construct_gazepoint_combined_pupil(
  reconstructed,
  left_col = "pupil_left",
  right_col = "pupil_right",
  policy = "reconstructed_mean",
  output_col = "pupil_reconstruction_policy",
  status_col = "pupil_reconstruction_source"
)

table(combined$pupil_reconstruction_source)
```

Alternative policies include `complete_case`, `available_eye`, `left_only`, and
`right_only`.

## 6. Inspect temporal provenance

```{r trace, fig.cap="Observed binocular traces with reconstructed final values marked explicitly."}
plot_gazepoint_binocular_diagnostics(reconstructed, type = "trace")
```

```{r timeline, fig.cap="Row-level binocular information status over time."}
plot_gazepoint_binocular_diagnostics(reconstructed, type = "timeline")
```

```{r gaps, fig.cap="Duration of missing-eye runs that were reconstructed."}
plot_gazepoint_binocular_diagnostics(reconstructed, type = "gaps", bins = 15)
```

The dashboard option returns a list of these ggplot objects rather than adding a layout
dependency.

```{r dashboard}
dashboard <- plot_gazepoint_binocular_diagnostics(
  reconstructed,
  type = "dashboard"
)
names(dashboard)
```

## 7. Audit reconstruction burden and condition imbalance

```{r audit}
audit <- audit_gazepoint_binocular_reconstruction(
  reconstructed,
  by = "condition",
  max_reconstruction_prop = 0.20,
  max_group_rate_difference = 0.10
)

audit$overall
audit$by_group
audit$reconstruction_shift
audit$audit
```

The thresholds above are governance flags. They are not universal statistical or
physiological validity cutoffs.

```{r burden-plot, fig.cap="Reconstruction burden by experimental condition."}
plot_gazepoint_binocular_diagnostics(audit, type = "burden")
```

## 8. Compare alternative pupil-construction policies

```{r sensitivity}
sensitivity <- analyse_gazepoint_binocular_sensitivity(
  reconstructed,
  left_col = "pupil_left",
  right_col = "pupil_right",
  policies = c(
    "complete_case",
    "available_eye",
    "reconstructed_mean",
    "left_only",
    "right_only"
  ),
  condition_col = "condition"
)

sensitivity$summary
sensitivity$correlations
sensitivity$condition_contrasts
```

The condition contrasts are descriptive mean differences only. The function deliberately
does not generate a new p-value for every preprocessing policy.

```{r sensitivity-plot, fig.cap="Mean pupil diameter under alternative binocular construction policies."}
plot_gazepoint_binocular_diagnostics(sensitivity, type = "sensitivity")
```

## 9. Reconstruction and PCHIP are different decisions

Ong et al. (2025) used PCHIP to create a common high-frequency time base, removed
interpolated segments corresponding to original gaps longer than 250 ms, and then used
cross-eye regression when one eye remained unavailable. `gp3tools` does not hard-code
that study-specific sequence or cutoff.

For a workflow that uses cross-eye reconstruction first, PCHIP can be applied afterwards
to **remaining** short gaps in the explicitly constructed signal. This is useful when the
analyst wants to keep model-based cross-eye support separate from temporal interpolation.

```{r pchip, eval=TRUE}
if (requireNamespace("pracma", quietly = TRUE)) {
  pchip <- interpolate_gazepoint_pupil_pchip(
    combined,
    pupil_col = "pupil_reconstruction_policy",
    time_col = "timestamp_ms",
    grouping_cols = c("subject", "trial"),
    max_gap_ms = 150,
    output_col = "pupil_after_reconstruction_pchip",
    flag_col = "pupil_pchip_after_reconstruction",
    status_col = "pupil_pchip_status"
  )

  table(pchip$pupil_pchip_after_reconstruction)
}
```

A study can reverse the order if its measurement and synchronization design requires it,
but it should report the choice. The two operations should not be collapsed into an
undocumented generic "imputation" step.

## 10. Baseline correction follows the declared combined signal

After choosing the pupil-construction policy, the resulting signal can enter the existing
baseline and smoothing workflow. Baseline windows should be designed prospectively when
possible. Ong et al. selected the first period with at least 400 ms of consecutive valid
bilateral data because a controlled baseline had not been collected; the authors identify
this as a limitation rather than a universal rule.

A study with an experimental baseline can use the existing
`baseline_correct_gazepoint_pupil()` interface on the declared combined column.

```{r baseline-example, eval=FALSE}
baseline_corrected <- baseline_correct_gazepoint_pupil(
  combined,
  pupil_col = "pupil_reconstruction_policy",
  time_col = "timestamp_ms",
  group_cols = c("subject", "trial"),
  baseline_window = c(0, 200),
  min_baseline_samples = 5
)
```

The example is not executed because the synthetic trials are not designed as a real
baseline experiment; the code illustrates where the existing function belongs in the
workflow.

## 11. Manuscript-ready reporting

```{r reporting}
reporting <- summarise_gazepoint_binocular_reporting(
  reconstructed,
  audit = audit,
  validation = validation
)

reporting$summary
cat(reporting$text)
```

The returned limitations can be reused when preparing a methods supplement or analysis
decision log.

```{r reporting-limitations}
reporting$limitations
```

## 12. When reconstruction should not be used

A conservative workflow should retain an explicit `reconstruction_ineligible` or
`reconstruction_blocked_*` status when:

- too few bilateral observations are available for calibration;
- one eye has effectively no variance;
- the fitted relationship is unstable or biologically implausible;
- a prediction requires prohibited extrapolation;
- the predicted diameter violates declared bounds;
- the missing-eye run exceeds the declared duration limit;
- a row is explicitly excluded because it is blink/artifact contaminated;
- pseudo-missing validation shows unacceptable prediction error for the intended use;
- reconstruction burden is so large or imbalanced that the downstream result is
  materially policy-dependent.

The framework therefore permits a scientifically meaningful outcome of **no
reconstruction**.

## 13. Relationship to the existing lightweight binocular helpers

Use `combine_gazepoint_eyes()` when a transparent sample-wise fusion rule is all that is
required. Use the existing `regress_gazepoint_pupils()` workflow when its documented
regression-fusion behavior matches the analysis plan. Use the functions in this article
when the goal is specifically to reconstruct a temporarily missing eye with explicit
calibration eligibility, gap/extrapolation gates, pseudo-missing validation, row-level
provenance, reconstruction burden, and sensitivity analysis.

No existing function changes its default behavior.

## References

Ong, J., He, W., Maglanque, P., Jiang, X., Gillman, L. M., Vergis, A., & Hardy, K. (2025). A Preprocessing Pipeline for Pupillometry Signal from Multimodal iMotion Data. *Sensors*, 25(15), 4737. doi:10.3390/s25154737.
