---
title: "Bayesian planning workflow"
description: "Choose model families, check Bayesian readiness, extract pupil and scanpath features, and prepare external Bayesian analyses."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.retina = 2
)
```


This article demonstrates how the Bayesian-planning helpers in `gp3tools`
can be used before committing to a full Bayesian model. The workflow covers
model-family selection, structural readiness checks, pupil-response feature
extraction, scanpath geometry, HDDM export preparation, and `brms` template
generation.

The helpers support analysis planning and reproducible preparation. They do
not require Stan, `brms`, HDDM, or Python unless the user subsequently chooses
to fit an external model.

```{r load-package}
library(gp3tools)

set.seed(42)
```

## Choose an outcome family before fitting

The statistical family should follow the scale and distribution of the
outcome rather than being selected after inspecting significance.

```{r model-family-table}
families <- recommend_gazepoint_model_family()

selected_metrics <- c(
  "fixation_duration",
  "dwell_time",
  "fixation_count",
  "aoi_proportion",
  "pupil_timecourse",
  "scanpath_length",
  "binary_choice"
)

families[
  families$metric %in% selected_metrics,
  ,
  drop = FALSE
]
```

The recommendations are planning guidance. Distributional shape, zero values,
boundary values, overdispersion, hierarchical structure, and study design
still need to be inspected in the actual dataset.

## Simulate a compact pupil time course

The following synthetic dataset represents eight participants completing six
trials each. Negative times form a prestimulus baseline, and positive times
form the response window.

```{r simulate-pupil}
pupil <- expand.grid(
  subject = paste0(
    "S",
    sprintf("%02d", 1:8)
  ),
  trial = 1:6,
  time_ms = seq(
    -200,
    1000,
    by = 50
  )
)

pupil$condition <- ifelse(
  pupil$trial %% 2 == 0,
  "specific_claim",
  "vague_claim"
)

response_shape <- exp(
  -((pupil$time_ms - 450) / 260)^2
)

condition_shift <- ifelse(
  pupil$condition == "vague_claim",
  0.12,
  0.04
)

subject_effects <- setNames(
  rnorm(
    8,
    mean = 0,
    sd = 0.025
  ),
  paste0(
    "S",
    sprintf("%02d", 1:8)
  )
)

pupil$pupil_mm <- 3 +
  unname(
    subject_effects[pupil$subject]
  ) +
  condition_shift * response_shape +
  rnorm(
    nrow(pupil),
    mean = 0,
    sd = 0.025
  )

set.seed(123)

missing_rows <- sample(
  seq_len(nrow(pupil)),
  30
)

pupil$pupil_mm[missing_rows] <- NA_real_

head(pupil)
```

## Plot the synthetic pupil time course

```{r pupil-timecourse-plot}
tc <- aggregate(
  pupil_mm ~ time_ms + condition,
  data = pupil,
  FUN = function(x) {
    mean(
      x,
      na.rm = TRUE
    )
  }
)

specific_tc <- tc[
  tc$condition == "specific_claim",
  ,
  drop = FALSE
]

vague_tc <- tc[
  tc$condition == "vague_claim",
  ,
  drop = FALSE
]

plot(
  specific_tc$time_ms,
  specific_tc$pupil_mm,
  type = "l",
  lwd = 2,
  ylim = range(
    tc$pupil_mm,
    na.rm = TRUE
  ),
  xlab = "Time from stimulus onset (ms)",
  ylab = "Mean pupil size (mm)",
  main = "Synthetic pupil time course"
)

lines(
  vague_tc$time_ms,
  vague_tc$pupil_mm,
  lty = 2,
  lwd = 2
)

abline(
  v = 0,
  lty = 3
)

legend(
  "topright",
  legend = c(
    "Specific claim",
    "Vague claim",
    "Stimulus onset"
  ),
  lty = c(
    1,
    2,
    3
  ),
  lwd = c(
    2,
    2,
    1
  ),
  bty = "n"
)
```

## Extract trial-level pupil-response features

`summarize_gazepoint_pupil_response_features()` calculates transparent
trial-level summaries from sample-level pupil data. The baseline and response
windows are specified explicitly.

```{r pupil-feature-extraction}
pupil_features <- summarize_gazepoint_pupil_response_features(
  data = pupil,
  pupil = "pupil_mm",
  time = "time_ms",
  subject = "subject",
  trial = "trial",
  baseline_window = c(
    -200,
    0
  ),
  response_window = c(
    0,
    1000
  ),
  condition = "condition"
)

head(pupil_features)
```

The returned table includes the baseline mean, peak dilation, latency to peak,
area under the baseline-corrected curve, and response-window missingness.

```{r pupil-feature-plot}
boxplot(
  peak_dilation ~ condition,
  data = pupil_features,
  xlab = "Condition",
  ylab = "Peak dilation from baseline",
  main = "Trial-level pupil-response feature"
)

stripchart(
  peak_dilation ~ condition,
  data = pupil_features,
  vertical = TRUE,
  method = "jitter",
  pch = 16,
  add = TRUE
)
```

## Check readiness before modelling

`check_gazepoint_bayesian_readiness()` performs lightweight structural checks.
It can identify missing columns, insufficient within-participant observations,
missing condition variation, and problematic outcome values.

```{r bayesian-readiness}
readiness <- check_gazepoint_bayesian_readiness(
  data = pupil_features,
  outcome = "peak_dilation",
  subject = "subject",
  trial = "trial",
  condition = "condition",
  metric_type = "continuous",
  min_observations_per_subject = 4
)

readiness
```

A warning is a prompt for review rather than an automatic exclusion decision.
The appropriate threshold depends on the study design, outcome, planned model,
and sensitivity analyses.

## Create a statistical analysis plan checklist

```{r bayesian-sap}
sap <- create_gazepoint_bayesian_sap(
  outcome = "Peak pupil dilation",
  design = paste(
    "Within-participant contrast between",
    "specific-claim and vague-claim trials"
  ),
  primary_model = "Bayesian hierarchical Gaussian model",
  baseline_window = c(
    -200,
    0
  ),
  analysis_window = c(
    0,
    1000
  ),
  missingness_threshold = 0.20,
  blink_padding_ms = 50,
  output = "data.frame"
)

sap
```

The checklist makes preprocessing, exclusion, prior, hierarchical-structure,
convergence, and reporting decisions explicit before the main analysis.

## Prepare an HDDM export

HDDM requires one row per decision trial with a participant identifier,
positive response time, and binary response. Continuous predictors can be
standardized within participant.

```{r hddm-export}
choice_data <- pupil_features

set.seed(456)

choice_data$rt <- runif(
  nrow(choice_data),
  min = 0.45,
  max = 1.40
)

choice_data$response <- sample(
  c(
    0,
    1
  ),
  nrow(choice_data),
  replace = TRUE
)

hddm_ready <- prepare_gazepoint_hddm_export(
  data = choice_data,
  subject = "subject",
  rt = "rt",
  response = "response",
  predictors = c(
    "peak_dilation",
    "auc"
  )
)

head(hddm_ready)
```

The helper does not fit a drift-diffusion model. The returned table can be
written to CSV and analysed in a separately documented Python/HDDM
environment.

## Create a `brms` model template without fitting

```{r brms-template}
pupil_template <- create_gazepoint_brms_template(
  metric_type = "pupil_timecourse",
  outcome = "pupil_bc",
  time = "time_ms",
  condition = "condition",
  subject = "subject"
)

pupil_template
```

The output contains a candidate formula, family, prior text, and reporting
notes. It is deliberately a template rather than a fitted model.

## Extract scanpath geometry

The next example creates synthetic sequential gaze coordinates and extracts
trial-level scanpath geometry.

```{r simulate-scanpath}
set.seed(789)

scanpath <- do.call(
  rbind,
  lapply(
    1:10,
    function(trial_id) {
      n <- 25

      data.frame(
        subject = paste0(
          "S",
          sprintf(
            "%02d",
            ceiling(trial_id / 2)
          )
        ),
        trial = trial_id,
        condition = ifelse(
          trial_id %% 2 == 0,
          "specific_claim",
          "vague_claim"
        ),
        time_ms = seq_len(n) * 40,
        x = cumsum(
          rnorm(
            n,
            mean = ifelse(
              trial_id %% 2 == 0,
              9,
              13
            ),
            sd = 8
          )
        ) + 400,
        y = cumsum(
          rnorm(
            n,
            mean = 4,
            sd = 6
          )
        ) + 300
      )
    }
  )
)

head(scanpath)
```

```{r scanpath-features}
scan_features <- compute_gazepoint_scanpath_geometry(
  data = scanpath,
  x = "x",
  y = "y",
  subject = "subject",
  trial = "trial",
  time = "time_ms",
  condition = "condition"
)

head(scan_features)
```

The output includes scanpath length, straight-line distance, path efficiency,
convex-hull area, spatial dispersion, and the number of valid points.

```{r scanpath-plot}
one_trial <- scanpath[
  scanpath$trial == 1,
  ,
  drop = FALSE
]

plot(
  one_trial$x,
  one_trial$y,
  type = "b",
  pch = 16,
  xlab = "Screen x-coordinate",
  ylab = "Screen y-coordinate",
  main = "Example synthetic scanpath"
)

text(
  one_trial$x[
    c(
      1,
      nrow(one_trial)
    )
  ],
  one_trial$y[
    c(
      1,
      nrow(one_trial)
    )
  ],
  labels = c(
    "Start",
    "End"
  ),
  pos = c(
    2,
    4
  )
)
```

```{r scanpath-feature-plot}
boxplot(
  scanpath_length ~ condition,
  data = scan_features,
  xlab = "Condition",
  ylab = "Scanpath length",
  main = "Extracted scanpath-length feature"
)

stripchart(
  scanpath_length ~ condition,
  data = scan_features,
  vertical = TRUE,
  method = "jitter",
  pch = 16,
  add = TRUE
)
```

## Interpretation boundary

The helpers above produce model-ready quantities and reproducible planning
objects. They do not independently establish deeper cognition, critical
evaluation, comprehension, credibility assessment, or a causal psychological
mechanism.

Eye-movement measures can support claims about visual allocation, timing,
sequence, and movement geometry. Pupil measures can support claims about
pupil dynamics under documented preprocessing and luminance control.
Stronger psychological interpretations require an explicit experimental
design, preregistered contrasts, validated complementary measures, suitable
models, and appropriately cautious language.
