---
title: "Binocular Pupil Fusion and Downsampling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Binocular Pupil Fusion and Downsampling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4.5
)
library(gp3tools)
library(ggplot2)
```

## Purpose

This article demonstrates three lightweight binocular pupil operations:

- sample-wise left/right averaging;
- cross-eye regression as a diagnostic smoothing and fusion method;
- integer-factor aggregation for lower-frequency analysis tables.

Regression-based fusion is a preprocessing device, not a causal model. It
should be compared with the simple binocular mean and inspected visually.

## Simulate binocular pupil measurements

```{r simulate}
set.seed(2026)

n <- 400
pupil <- data.frame(
  USER_ID = "P01",
  trial = rep(c("T01", "T02"), each = n / 2),
  TIME = rep(seq(0, 1.99, by = 0.01), 2)
)

latent <- 3.4 +
  0.20 * sin(2 * pi * pupil$TIME / 2) +
  ifelse(pupil$trial == "T02", 0.12, 0)

pupil$LPupil <- latent + rnorm(n, 0, 0.035)
pupil$RPupil <- 0.10 + 0.97 * latent + rnorm(n, 0, 0.045)

pupil$LPupil[85:90] <- NA_real_
pupil$RPupil[250:255] <- NA_real_
```

## Mean binocular pupil

```{r mean}
pupil_mean <- mean_gazepoint_pupil(
  pupil,
  lp_col = "LPupil",
  rp_col = "RPupil"
)

head(pupil_mean)
```

## Regression-smoothed binocular trace

```{r regression}
pupil_regressed <- regress_gazepoint_pupils(
  pupil_mean,
  lp_col = "LPupil",
  rp_col = "RPupil",
  group_cols = "trial",
  direction = "bidirectional",
  min_complete = 20
)

ggplot(pupil_regressed, aes(TIME)) +
  geom_line(aes(y = mean_pupil, linetype = "binocular mean")) +
  geom_line(
    aes(y = pupil_regressed, linetype = "regression-smoothed")
  ) +
  facet_wrap(~ trial) +
  labs(
    x = "Time (s)",
    y = "Pupil size",
    linetype = NULL,
    title = "Simple mean and bidirectional regression fusion"
  ) +
  theme_minimal()
```

## Downsample by aggregation

```{r downsample}
downsampled <- downsample_gazepoint_pupil(
  pupil_regressed,
  factor = 5,
  pupil_cols = c("mean_pupil", "pupil_regressed"),
  group_cols = "trial",
  method = "mean"
)

downsampled[c(
  "USER_ID", "trial", "TIME", "mean_pupil",
  "pupil_regressed", "n_samples_aggregated"
)]
```

```{r downsample-plot}
ggplot() +
  geom_line(
    data = pupil_regressed,
    aes(TIME, pupil_regressed, linetype = "original"),
    alpha = 0.55
  ) +
  geom_point(
    data = downsampled,
    aes(TIME, pupil_regressed, shape = "aggregated"),
    size = 1.5
  ) +
  facet_wrap(~ trial) +
  labs(
    x = "Time (s)",
    y = "Regression-smoothed pupil",
    linetype = NULL,
    shape = NULL,
    title = "Original and factor-five aggregated pupil traces"
  ) +
  theme_minimal()
```

## Recommended reporting

Report which eye columns were used, how missing monocular values were handled,
the regression direction, the minimum complete binocular samples required,
the fallback rule, the downsampling factor, and whether samples were averaged
or decimated.
