---
title: "Cluster-permutation time-course workflow"
---

```{r setup, include = FALSE}
library(gp3tools)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = '#>',
  fig.width = 7,
  fig.height = 4.5,
  warning = FALSE,
  message = FALSE
)
```

This article gives a compact example of the conservative cluster-permutation workflow in `gp3tools`. The workflow is intended for two-condition, within-subject, one-dimensional time-course data. It is suitable for cautious screening of time-contiguous effects, not for exact onset or offset claims.

## Simulated example data

```{r}
set.seed(123)

raw <- simulate_gazepoint_cluster_timecourse_data(
  n_subjects = 12,
  n_time_bins = 30,
  conditions = c("control", "treatment"),
  effect_start = 12,
  effect_end = 20,
  effect_size = 0.7,
  seed = 123
)

head(raw)
```

## Prepare the time-course grid

`prepare_gazepoint_timecourse_test_data()` converts a long-format data set into the internal column contract expected by `run_gazepoint_cluster_permutation()`.

```{r}
prepared <- prepare_gazepoint_timecourse_test_data(
  raw,
  subject_col = "subject",
  condition_col = "condition",
  time_col = "time_bin",
  outcome_col = "outcome",
  condition_order = c("control", "treatment")
)

head(prepared)
```

## Audit the grid

Before running a cluster-permutation workflow, check whether the data grid is complete, paired, and compatible with the validated two-condition workflow.

```{r}
grid_audit <- audit_gazepoint_timecourse_grid(prepared)

grid_audit$grid_summary
grid_audit$readiness
```

```{r}
design_diagnostic <- diagnose_gazepoint_cluster_design(prepared)

design_diagnostic
```

## Run the cluster-permutation workflow

For a fast documentation example, this article uses a small number of permutations. Real analyses should use a larger number and report the settings.

```{r}
cluster_result <- run_gazepoint_cluster_permutation(
  prepared,
  condition_order = c("control", "treatment"),
  n_permutations = 99,
  cluster_threshold = 1.5,
  min_time_bins = 2,
  seed = 123
)

names(cluster_result)
```

## Summarise clusters

```{r}
cluster_summary <- summarize_gazepoint_time_clusters(cluster_result)

cluster_summary
```

## Plot the cluster result

```{r}
plot_gazepoint_cluster_permutation(
  cluster_result,
  plot_type = "difference",
  significant_only = FALSE
)
```

## Cautious report text

`report_gazepoint_cluster_permutation()` returns a compact report object. The generated wording is deliberately cautious: detected ranges are treated as cluster-level time intervals, not as exact effect-onset or effect-offset estimates.

```{r}
cluster_report <- report_gazepoint_cluster_permutation(cluster_result)

cluster_report$report_text
```

## Threshold sensitivity

Cluster-level results can depend on the cluster-forming threshold. A compact sensitivity table helps document whether the broad pattern is stable across reasonable threshold choices.

```{r}
threshold_sensitivity <- run_gazepoint_cluster_threshold_sensitivity(
  prepared,
  thresholds = c(1.25, 1.5, 1.75),
  condition_order = c("control", "treatment"),
  n_permutations = 49,
  min_time_bins = 2,
  seed = 123
)

threshold_sensitivity$summary
```

## External export helpers

The external export helpers prepare long-format files and README notes for possible continuation in specialist tools. They do not run or validate the external analyses.

```{r}
external_dir <- tempfile("gp3_cluster_export_")

external_files <- export_gazepoint_mne_cluster_input(
  prepared,
  outdir = external_dir
)

external_files
```

## Guardrails for unsupported extensions

Some advanced method names are available as explicit guardrails. They intentionally stop with explanatory messages rather than silently supporting methods outside the validated scope.

```{r}
try(run_gazepoint_tfce(), silent = TRUE)
```
