---
title: "Exporting to other ecosystems"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exporting to other ecosystems}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)

library(gp3tools)
```

This article summarises package bridges for preparing outputs for other R, Python, and physiological-analysis ecosystems.

The goal is not to replace specialised tools, but to create clean and documented handoff tables.

## Eye-tracking ecosystem adapters

```{r, eval=FALSE}
eyetrackingr_data <- prepare_gazepoint_eyetrackingr_input(all_gaze)
pupillometryr_data <- prepare_gazepoint_pupillometryr_input(all_gaze)
gazer_data <- run_gazepoint_gazer_crosscheck(all_gaze)
eyetools_data <- run_gazepoint_eyetools_fixation_detection(all_gaze)
```

## Physiological and EDA bridges

```{r, eval=FALSE}
rhrv_input <- prepare_gazepoint_rhrv_input(biometric_data)
pyppg_input <- prepare_gazepoint_pyppg_input(biometric_data)

ledalab_input <- prepare_gazepoint_ledalab_input(eda_data)
pspm_input <- prepare_gazepoint_pspm_input(eda_data)
cvxeda_input <- prepare_gazepoint_cvxeda_input(eda_data)
```

## Recommended metadata

Exported handoff tables should retain participant IDs, trial IDs, time columns, sampling-rate information, event markers, preprocessing flags, and package-generated diagnostics.

## Reporting note

When using external tools, report both the gp3tools preparation step and the downstream software version/settings.
