---
title: "Event-detector comparison workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Event-detector comparison workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5
)
```

## Scope

Different event detectors implement different definitions and thresholds.
`compare_gazepoint_event_detectors()` places native velocity-threshold,
lightweight HMM, and optional eyetools outputs into a common event table.

Agreement indicates methodological convergence between detector definitions.
It does not establish that one detector recovers a uniquely true cognitive
event sequence.

```{r load-package}
library(gp3tools)
```

## Synthetic gaze trace

```{r synthetic-data}
n <- 180L

gaze <- data.frame(
  USER_ID = rep("P01", n),
  trial = rep(c("T01", "T02"), each = n / 2),
  TIME = seq(0, by = 1 / 60, length.out = n),
  FPOGX = c(
    rep(0.20, 55),
    seq(0.20, 0.80, length.out = 10),
    rep(0.80, 55),
    seq(0.80, 0.35, length.out = 10),
    rep(0.35, 50)
  ),
  FPOGY = 0.50,
  stringsAsFactors = FALSE
)

utils::head(gaze)
```

## Compare velocity thresholds and the HMM branch

```{r compare-detectors}
comparison <- compare_gazepoint_event_detectors(
  data = gaze,
  id_col = "USER_ID",
  trial_col = "trial",
  x_col = "FPOGX",
  y_col = "FPOGY",
  time_col = "TIME",
  methods = c("velocity", "hmm"),
  velocity_thresholds = c(5, 10, 20),
  min_duration = 60,
  hmm_states = 3,
  min_overlap = 0.5
)

comparison$runs
```

A detector branch that cannot be fitted is recorded in `runs` while successful
branches remain available for review.

## Detector-level summaries

```{r detector-summary}
comparison$detector_summary
```

The summary reports fixation counts and duration distributions. Large
differences indicate sensitivity to event-definition choices.

```{r count-plot}
plot_gazepoint_event_detector_agreement(
  comparison,
  plot = "counts"
)
```

```{r duration-plot}
plot_gazepoint_event_detector_agreement(
  comparison,
  plot = "durations"
)
```

## Pairwise interval agreement

```{r agreement-summary}
comparison$pairwise_agreement
```

Events are compared by interval intersection-over-union. The directional
agreement columns distinguish the proportion of events from detector A that
match detector B from the reverse comparison.

```{r agreement-plot}
if (nrow(comparison$pairwise_agreement)) {
  plot_gazepoint_event_detector_agreement(
    comparison,
    plot = "agreement"
  )
}
```

## Unmatched events

```{r unmatched-events}
utils::head(comparison$unmatched_events)
```

Unmatched events support targeted review of threshold sensitivity, boundary
differences, short events, and detector-specific segmentation.

## Optional eyetools branch

The external branch remains optional and is never a required package
dependency.

```{r eyetools-example, eval=FALSE}
comparison_external <- compare_gazepoint_event_detectors(
  data = gaze,
  id_col = "USER_ID",
  trial_col = "trial",
  x_col = "FPOGX",
  y_col = "FPOGY",
  time_col = "TIME",
  methods = c("velocity", "hmm", "eyetools"),
  run_optional_eyetools = TRUE,
  eyetools_method = "vti"
)
```

When `eyetools` is unavailable or the optional branch is disabled, the run
table records an explicit skip status.

## Reporting

Report:

- coordinate units and any conversion to visual degrees;
- sampling rate and timestamp units;
- velocity thresholds;
- minimum event durations;
- HMM state count;
- eyetools method and parameters when used;
- interval-overlap criterion;
- detector-specific event counts and durations;
- pairwise agreement and unmatched-event rates.

Treat detector comparison as a sensitivity and validation exercise rather than
as evidence that one algorithm captures the only valid event structure.
