---
title: "Benchmarking gaze-event detectors against reviewed intervals"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 7
    fig_height: 4.5
vignette: >
  %\VignetteIndexEntry{Benchmarking gaze-event detectors against reviewed intervals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(gp3tools)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center"
)
```

## Purpose

Event detectors can disagree because they use different definitions, thresholds,
and model assumptions. `gp3tools` therefore separates two questions:

1. How do detector outputs agree with one another?
2. How closely does each detector reproduce a specified set of reviewed or
   synthetic reference intervals?

The first question is addressed by `compare_gazepoint_event_detectors()`. This
article addresses the second with:

- `create_gazepoint_event_review_template()`;
- `benchmark_gazepoint_event_detectors()`;
- `summarise_gazepoint_event_detector_benchmark()`;
- `plot_gazepoint_event_detector_benchmark()`.

The reviewed intervals are a reference standard for a particular audit, not a
claim that one event definition is universally correct.

## Synthetic trace with known fixation intervals

The example below creates three stable gaze periods separated by rapid
transitions. The reference table records the intervals used to generate the
trace.

```{r synthetic-data}
set.seed(20260716)

sample_rate <- 100
sample_time <- seq(0, 2.39, by = 1 / sample_rate)

x <- c(
  rep(0.20, 80),
  seq(0.20, 0.50, length.out = 10),
  rep(0.50, 70),
  seq(0.50, 0.80, length.out = 10),
  rep(0.80, 70)
)

gaze <- data.frame(
  USER_ID = "P01",
  trial = "T01",
  TIME = sample_time,
  FPOGX = x + rnorm(length(x), sd = 0.0015),
  FPOGY = 0.50 + rnorm(length(x), sd = 0.0015)
)

reviewed <- data.frame(
  USER_ID = "P01",
  trial = "T01",
  review_event_id = 1:3,
  start_time = c(0.00, 0.90, 1.70),
  end_time = c(0.79, 1.59, 2.39),
  event_type = "fixation",
  review_status = "accepted"
)

head(gaze)
reviewed
```

A simple trace plot helps verify that the reviewed intervals correspond to the
intended stable periods.

```{r synthetic-trace, fig.cap="Synthetic gaze trace and reviewed fixation intervals."}
plot(
  gaze$TIME,
  gaze$FPOGX,
  type = "l",
  xlab = "Time (s)",
  ylab = "Horizontal gaze coordinate",
  main = "Synthetic fixation and transition structure"
)

for (i in seq_len(nrow(reviewed))) {
  rect(
    reviewed$start_time[i],
    par("usr")[3],
    reviewed$end_time[i],
    par("usr")[4],
    border = NA,
    density = 12,
    angle = 45
  )
}

lines(gaze$TIME, gaze$FPOGX)
```

## Run native velocity, HMM, and optional external branches

The comparison layer runs all requested methods on the same sample-level data.
The optional external branch is recorded as disabled unless explicitly enabled
and `{eyetools}` is installed.

```{r detector-comparison}
comparison <- compare_gazepoint_event_detectors(
  gaze,
  trial_col = "trial",
  methods = c("velocity", "hmm", "eyetools"),
  velocity_thresholds = c(5, 10, 20),
  min_duration = 50,
  hmm_states = 3,
  run_optional_eyetools = FALSE
)

comparison$runs
comparison$detector_summary
```

Detector failures do not discard successful branches. The benchmark uses only
successful detector outputs retained in the comparison object.

## Benchmark against the synthetic reference

```{r benchmark}
benchmark <- benchmark_gazepoint_event_detectors(
  comparison,
  reviewed,
  min_overlap = 0.50,
  time_unit = "seconds"
)

summarise_gazepoint_event_detector_benchmark(
  benchmark,
  level = "detector"
)
```

The detector-level table reports:

- reviewed and detected event counts;
- true positives, false positives, and false negatives;
- precision, recall, and F1;
- matched-event intersection-over-union;
- onset, offset, and duration error in milliseconds;
- detected-minus-reviewed count bias.

Matching is one-to-one within each participant/trial sequence. Candidate pairs
are ordered by decreasing interval intersection-over-union and retained only
when neither event has already been matched.

## Inspect sequence-level and event-level diagnostics

```{r detailed-results}
head(
  summarise_gazepoint_event_detector_benchmark(
    benchmark,
    level = "sequence"
  )
)

head(
  summarise_gazepoint_event_detector_benchmark(
    benchmark,
    level = "matches"
  )
)

summarise_gazepoint_event_detector_benchmark(
  benchmark,
  level = "errors"
)
```

The unmatched-event table distinguishes false positives from false negatives,
which is useful when reviewing threshold sensitivity or detector-specific
failure modes.

## Plot benchmark diagnostics

```{r benchmark-f1, fig.cap="Event-level F1 scores against the reviewed intervals."}
plot_gazepoint_event_detector_benchmark(
  benchmark,
  plot = "f1"
)
```

```{r benchmark-pr, fig.cap="Precision and recall against the reviewed intervals."}
plot_gazepoint_event_detector_benchmark(
  benchmark,
  plot = "precision_recall"
)
```

```{r benchmark-overlap, fig.cap="Mean interval overlap among matched events."}
plot_gazepoint_event_detector_benchmark(
  benchmark,
  plot = "overlap"
)
```

```{r benchmark-timing, fig.cap="Absolute onset, offset, and duration errors."}
plot_gazepoint_event_detector_benchmark(
  benchmark,
  plot = "timing_error"
)
```

## Prepare manually reviewed data

Create a CSV-ready template from the sample-level data. One placeholder row is
created per sequence by default; reviewers can duplicate rows for additional
fixations.

```{r review-template}
review_template <- create_gazepoint_event_review_template(
  gaze,
  trial_col = "trial",
  rows_per_sequence = 1,
  reviewer = "reviewer_1"
)

review_template
```

A real review workflow can keep the annotation file outside the package
repository:

```{r manual-review-files, eval=FALSE}
write.csv(
  review_template,
  "C:/path/outside/repository/event_review.csv",
  row.names = FALSE
)

reviewed_real <- read.csv(
  "C:/path/outside/repository/event_review.csv",
  stringsAsFactors = FALSE
)

benchmark_real <- benchmark_gazepoint_event_detectors(
  comparison_real,
  reviewed_real,
  min_overlap = 0.50,
  time_unit = "seconds"
)
```

For each accepted event, fill `start_time` and `end_time`, preserve the sequence
identifiers, and set `review_status` to `"accepted"`. Pending or rejected rows
are excluded from the benchmark.

## Recommended reporting

A transparent methods supplement should report:

- how reviewed intervals were produced;
- the number of reviewers and any adjudication procedure;
- the overlap threshold used for matching;
- detector settings and sampling rate;
- precision, recall, F1, overlap, and timing error;
- the number and type of failed or skipped detector runs;
- whether conclusions changed across plausible detector definitions.

Benchmark results support sensitivity assessment. They should not be used to
claim that a detector uniquely identifies cognitive states or exact mental-event
onsets.
