---
title: "Experimental Bayesian bridge helpers"
description: "Optional workflows for brms, HDDM scripts, adaptive trial selection, HMM event classification, Gaussian-process pupil imputation, and CNN uncertainty filtering."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.retina = 2
)
```


The experimental bridge helpers connect `gp3tools` outputs with optional
Bayesian and machine-learning workflows while keeping heavy external
dependencies outside the core package.

The helpers support reproducible preparation and diagnostics. They do not
make an external model automatically valid, and they do not replace
domain-specific validation.

```{r load-package}
library(gp3tools)

set.seed(101)
```

## Optional `brms` fitting

`fit_gazepoint_brms_model()` calls `brms::brm()` only when `brms` is installed.
A complete fit can also require a working Stan backend, appropriate priors,
sufficient sampling, and convergence assessment.

The following is a template and is not evaluated while the article is built:

```{r optional-brms-fit, eval=FALSE}
fit <- fit_gazepoint_brms_model(
  data = model_data,
  formula = pupil_peak ~ condition + (1 | subject),
  family = "gaussian",
  chains = 4,
  iter = 2000,
  warmup = 1000,
  cores = 4,
  backend = "cmdstanr"
)

summary(fit)
```

For documentation or preregistration, users can first create a dependency-free
template:

```{r dwell-brms-template}
dwell_template <- create_gazepoint_brms_template(
  metric_type = "dwell_time",
  outcome = "claim_dwell_ms",
  condition = "condition",
  subject = "subject",
  item = "stimulus"
)

dwell_template
```

## Generate an HDDM fitting script

The R package does not import HDDM or execute Python. Instead, it can create a
reproducible Python script from an HDDM-ready CSV export.

```{r create-hddm-data}
hddm_data <- data.frame(
  subj_idx = rep(
    1:4,
    each = 8
  ),
  rt = runif(
    32,
    min = 0.45,
    max = 1.40
  ),
  response = sample(
    c(
      0,
      1
    ),
    32,
    replace = TRUE
  ),
  target_dwell_ms_z = rnorm(32),
  pupil_peak_z = rnorm(32)
)

head(hddm_data)
```

```{r generate-hddm-script}
tmp_csv <- tempfile(
  fileext = ".csv"
)

tmp_py <- tempfile(
  fileext = ".py"
)

utils::write.csv(
  hddm_data,
  tmp_csv,
  row.names = FALSE
)

create_gazepoint_hddm_fit_script(
  data_file = tmp_csv,
  output_file = tmp_py,
  regressions = c(
    v = "target_dwell_ms_z",
    a = "pupil_peak_z"
  ),
  draws = 1000,
  burn = 500
)

readLines(
  tmp_py,
  n = 18
)
```

The generated script should be archived with the analysis and executed in a
documented HDDM environment. The HDDM version, Python environment, model
parameterization, sampling settings, convergence checks, and posterior
predictive assessment should be reported.

## Select an adaptive trial

`select_gazepoint_adaptive_trial()` implements lightweight acquisition rules
when candidate-level posterior means and uncertainties have already been
estimated.

```{r adaptive-candidates}
candidates <- data.frame(
  stimulus = paste0(
    "stimulus_",
    1:8
  ),
  posterior_mean = c(
    0.20,
    0.32,
    0.28,
    0.45,
    0.38,
    0.31,
    0.41,
    0.35
  ),
  posterior_sd = c(
    0.10,
    0.16,
    0.28,
    0.08,
    0.20,
    0.14,
    0.12,
    0.24
  )
)

selected <- select_gazepoint_adaptive_trial(
  candidates = candidates,
  mean = "posterior_mean",
  sd = "posterior_sd",
  acquisition = "ucb",
  kappa = 2
)

selected
```

```{r adaptive-selection-plot}
scored <- candidates

scored$acquisition_score <-
  scored$posterior_mean +
  2 * scored$posterior_sd

bar_positions <- barplot(
  scored$acquisition_score,
  names.arg = scored$stimulus,
  las = 2,
  ylab = "Upper-confidence-bound score",
  main = "Adaptive candidate selection"
)

selected_index <- match(
  selected$stimulus,
  scored$stimulus
)

points(
  bar_positions[selected_index],
  selected$acquisition_score,
  pch = 19,
  cex = 1.4
)

legend(
  "topright",
  legend = "Selected next trial",
  pch = 19,
  bty = "n"
)
```

Adaptive selection requires a prespecified acquisition rule and appropriate
guardrails for stimulus balance, participant burden, stopping rules, and
confirmatory inference.

## Lightweight HMM event classification

`classify_gazepoint_events_hmm()` estimates gaze velocity and applies a
lightweight unsupervised hidden-state classifier.

It is suitable for exploratory diagnostics and method-development workflows.
It is not a replacement for a validated fixation/saccade detector.

```{r simulate-hmm-gaze}
n <- 90

gaze <- data.frame(
  subject = "S01",
  time_ms = seq_len(n) * 16,
  x = cumsum(
    c(
      rnorm(
        30,
        mean = 0.05,
        sd = 0.02
      ),
      rnorm(
        25,
        mean = 4.00,
        sd = 0.80
      ),
      rnorm(
        35,
        mean = 0.40,
        sd = 0.10
      )
    )
  ) + 500,
  y = cumsum(
    c(
      rnorm(
        30,
        mean = 0.04,
        sd = 0.02
      ),
      rnorm(
        25,
        mean = 3.50,
        sd = 0.70
      ),
      rnorm(
        35,
        mean = 0.30,
        sd = 0.10
      )
    )
  ) + 300
)

head(gaze)
```

```{r hmm-classification}
events <- classify_gazepoint_events_hmm(
  data = gaze,
  x = "x",
  y = "y",
  time = "time_ms",
  subject = "subject",
  n_states = 3,
  state_labels = c(
    "slow",
    "medium",
    "fast"
  )
)

table(
  events$hmm_event,
  useNA = "ifany"
)
```

```{r hmm-path-plot}
event_symbol <- match(
  events$hmm_event,
  c(
    "slow",
    "medium",
    "fast"
  )
)

plot(
  events$x,
  events$y,
  type = "l",
  xlab = "Screen x-coordinate",
  ylab = "Screen y-coordinate",
  main = "Synthetic gaze path and HMM states"
)

valid_event <- !is.na(event_symbol)

points(
  events$x[valid_event],
  events$y[valid_event],
  pch = event_symbol[valid_event]
)

legend(
  "topleft",
  legend = c(
    "Slow",
    "Medium",
    "Fast"
  ),
  pch = 1:3,
  bty = "n"
)
```

```{r hmm-velocity-plot}
plot(
  events$time_ms,
  events$gaze_velocity,
  type = "h",
  xlab = "Time (ms)",
  ylab = "Coordinate velocity per millisecond",
  main = "Velocity used by the HMM classifier"
)
```

The state labels are relative to the observed sequence. They should not be
treated as universal fixation, pursuit, and saccade labels without external
validation.

## Gaussian-process pupil imputation

`impute_gazepoint_pupil_gp()` performs within-sequence interpolation with a
Gaussian-process smoother.

The intended use is limited reconstruction of short missing intervals after
artifact and blink detection. Long unusable intervals should ordinarily
remain missing or trigger trial-level review.

```{r simulate-missing-pupil}
pupil <- data.frame(
  subject = "S01",
  trial = 1,
  time_ms = seq(
    0,
    1500,
    by = 25
  )
)

pupil$pupil_mm <-
  3 +
  0.18 *
    exp(
      -((pupil$time_ms - 650) / 280)^2
    ) +
  rnorm(
    nrow(pupil),
    sd = 0.015
  )

pupil$pupil_mm[
  pupil$time_ms >= 425 &
    pupil$time_ms <= 550
] <- NA_real_

pupil$pupil_mm[
  pupil$time_ms >= 900 &
    pupil$time_ms <= 975
] <- NA_real_

head(pupil)
```

```{r gp-imputation}
imputed <- impute_gazepoint_pupil_gp(
  data = pupil,
  pupil = "pupil_mm",
  time = "time_ms",
  subject = "subject",
  trial = "trial",
  max_train = 80
)

table(
  imputed$pupil_was_gp_imputed
)
```

```{r gp-imputation-plot}
plot(
  imputed$time_ms,
  imputed$pupil_gp_imputed,
  type = "l",
  lwd = 2,
  xlab = "Time (ms)",
  ylab = "Pupil size (mm)",
  main = "Gaussian-process pupil imputation"
)

observed <- !is.na(
  imputed$pupil_mm
)

was_imputed <-
  imputed$pupil_was_gp_imputed

points(
  imputed$time_ms[observed],
  imputed$pupil_mm[observed],
  pch = 16
)

points(
  imputed$time_ms[was_imputed],
  imputed$pupil_gp_imputed[was_imputed],
  pch = 1,
  cex = 1.3
)

legend(
  "topright",
  legend = c(
    "GP-completed series",
    "Observed samples",
    "Imputed samples"
  ),
  lty = c(
    1,
    NA,
    NA
  ),
  pch = c(
    NA,
    16,
    1
  ),
  bty = "n"
)
```

The imputation flag should be retained so the percentage and location of
reconstructed samples can be audited.

## CNN and webcam uncertainty filtering

`filter_gazepoint_cnn_uncertainty()` post-processes externally generated gaze
coordinates. It does not train or validate a webcam or convolutional neural
network.

```{r simulate-cnn-predictions}
cnn <- data.frame(
  frame = 1:120,
  x = 500 + cumsum(
    rnorm(
      120,
      sd = 3
    )
  ),
  y = 300 + cumsum(
    rnorm(
      120,
      sd = 2
    )
  ),
  uncertainty = c(
    runif(
      100,
      0.05,
      0.40
    ),
    runif(
      20,
      1.20,
      3.00
    )
  )
)

cnn$x[
  sample(
    seq_len(nrow(cnn)),
    5
  )
] <- NA_real_
```

```{r cnn-filtering}
filtered <- filter_gazepoint_cnn_uncertainty(
  data = cnn,
  x = "x",
  y = "y",
  uncertainty = "uncertainty",
  max_uncertainty = 1
)

table(
  filtered$cnn_valid_frame
)
```

```{r cnn-coordinate-plot}
plot(
  filtered$x,
  filtered$y,
  pch = 16,
  cex = 0.5 +
    2 * filtered$cnn_uncertainty_weight,
  xlab = "Predicted x-coordinate",
  ylab = "Predicted y-coordinate",
  main = "Webcam/CNN predictions weighted by uncertainty"
)
```

```{r cnn-weight-plot}
plot(
  filtered$frame,
  filtered$cnn_uncertainty_weight,
  type = "h",
  xlab = "Frame",
  ylab = "Uncertainty weight",
  main = "Frame-level uncertainty weights"
)

abline(
  h = 0,
  lty = 3
)
```

Invalid frames receive zero weight. Nevertheless, an uncertainty threshold is
not a substitute for calibration, accuracy, precision, latency, demographic
bias, head-pose, illumination, and out-of-distribution validation.

## Practical reporting boundary

For an external Bayesian or machine-learning analysis, report:

- external software and version;
- computational environment;
- model formula and parameterization;
- prior distributions;
- chain, iteration, and warmup settings;
- convergence and posterior predictive diagnostics;
- adaptive-selection or stopping rule;
- event-classification validation;
- missingness and imputation rules;
- uncertainty definition and threshold;
- sensitivity analyses;
- the distinction between exploratory and confirmatory outputs.

These bridge helpers improve preparation and auditability. They do not make
advanced models automatically robust, causal, confirmatory, or
psychologically interpretable.
