---
title: "External face-data synchronisation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{External face-data synchronisation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  warning = FALSE,
  message = FALSE
)

library(gp3tools)
```

This article demonstrates how to align externally generated facial-behaviour data with Gazepoint rows.

The helpers shown here do **not** infer facial expressions from Gazepoint CSV files. They also do **not** interpret facial behaviour as emotion. Their purpose is narrower: to align already imported and standardised external face-analysis data with Gazepoint timing or frame indices, and to audit the quality of that alignment.

The current scope is:

1. nearest-time synchronisation;
2. exact-frame synchronisation;
3. synchronisation-status columns;
4. matched/unmatched/outside-tolerance summaries;
5. absolute time-difference summaries for nearest-time matching.

Window summaries, trial-level aggregation, multimodal modelling, and emotion interpretation are later workflow stages.

## Example Gazepoint and face-analysis data

The Gazepoint table may be a sample-level table, trial-level table, or another table with timing or frame information.

```{r}
gaze <- data.frame(
  subject_id = c("P001", "P001", "P001", "P002", "P002"),
  trial_id = c(1, 1, 1, 1, 1),
  time_sec = c(0.000, 0.033, 0.066, 0.000, 0.050),
  VID_FRAME = c(1, 2, 3, 1, 2),
  AOI = c("claim", "claim", "logo", "claim", "evidence"),
  stringsAsFactors = FALSE
)
```

The facial-behaviour table is assumed to come from an external face-analysis pipeline. It can be standardised before synchronisation, or standardised automatically by `sync_gazepoint_face_data()`.

```{r}
face <- data.frame(
  participant_id = c("P001", "P001", "P001", "P002", "P002"),
  frame = c(1, 2, 3, 1, 2),
  timestamp = c(0.000, 0.033, 0.066, 0.000, 0.060),
  confidence = c(0.98, 0.96, 0.94, 0.95, 0.93),
  success = c(1, 1, 1, 1, 1),
  AU04_r = c(0.05, 0.06, 0.05, 0.10, 0.12),
  AU12_r = c(0.20, 0.22, 0.25, 0.10, 0.08),
  stringsAsFactors = FALSE
)
```

## Nearest-time synchronisation

`sync_gazepoint_face_data()` can match each Gazepoint row to the nearest face-analysis row within a grouping structure. The `by` argument maps Gazepoint grouping columns to facial-data grouping columns.

```{r}
synced_time <- sync_gazepoint_face_data(
  gazepoint_data = gaze,
  face_data = face,
  method = "nearest_time",
  by = c(subject_id = "participant_id"),
  gaze_time_col = "time_sec",
  tolerance_sec = 0.050
)

synced_time
```

The output keeps the Gazepoint columns and appends matched face-analysis columns. It also adds synchronisation metadata:

```{r}
synced_time[, c(
  "subject_id",
  "time_sec",
  "face_time_sec",
  "face_confidence",
  "face_valid",
  "face_AU04_r",
  "face_AU12_r",
  "face_sync_status",
  "face_sync_diff_sec",
  "face_sync_abs_diff_sec",
  "face_sync_within_tolerance"
)]
```

The main status values are:

- `"matched"`: nearest facial row was within the tolerance;
- `"outside_tolerance"`: nearest facial row was found but exceeded the tolerance;
- `"unmatched"`: no facial row was available in the group;
- `"missing_gaze_time"`: the Gazepoint row had no usable time value.

## Exact-frame synchronisation

When Gazepoint and external face-analysis tables share frame indices, exact-frame matching can be used.

```{r}
synced_frame <- sync_gazepoint_face_data(
  gazepoint_data = gaze,
  face_data = face,
  method = "frame_exact",
  by = c(subject_id = "participant_id"),
  gaze_frame_col = "VID_FRAME"
)

synced_frame[, c(
  "subject_id",
  "VID_FRAME",
  "face_frame",
  "face_confidence",
  "face_AU04_r",
  "face_AU12_r",
  "face_sync_status",
  "face_sync_within_tolerance"
)]
```

Exact-frame matching does not estimate time differences. It is most appropriate when both data sources are known to share the same video-frame basis.

## Audit synchronisation quality

`audit_gazepoint_face_sync()` summarises synchronisation quality across all rows or within user-defined groups.

```{r}
sync_audit <- audit_gazepoint_face_sync(
  synced_time,
  group_cols = "subject_id",
  min_matched_percent = 70,
  warning_matched_percent = 85,
  max_abs_diff_sec = 0.050
)

sync_audit$overview
```

```{r}
sync_audit$group_summary
```

```{r}
sync_audit$issue_summary
```

The audit reports matched rows, unmatched rows, rows outside tolerance, missing timing or frame indicators, and absolute time-difference summaries when available.

## Example with an outside-tolerance row

The following example uses a stricter tolerance to illustrate how rows can be flagged for review.

```{r}
synced_strict <- sync_gazepoint_face_data(
  gazepoint_data = gaze,
  face_data = face,
  method = "nearest_time",
  by = c(subject_id = "participant_id"),
  gaze_time_col = "time_sec",
  tolerance_sec = 0.005
)

synced_strict[, c(
  "subject_id",
  "time_sec",
  "face_time_sec",
  "face_sync_status",
  "face_sync_abs_diff_sec",
  "face_sync_within_tolerance"
)]
```

```{r}
audit_gazepoint_face_sync(
  synced_strict,
  group_cols = "subject_id",
  min_matched_percent = 70,
  warning_matched_percent = 85,
  max_abs_diff_sec = 0.005
)$overview
```

## Recommended interpretation

Synchronisation status is a technical alignment diagnostic. It should not be interpreted as evidence of facial expression validity or emotional state.

Prefer cautious language such as:

- face-data alignment;
- matched facial-behaviour rows;
- synchronisation tolerance;
- timing difference;
- frame-level match;
- synchronisation coverage;
- rows requiring alignment review.

Avoid unsupported language such as:

- emotion detection from Gazepoint;
- hidden affect;
- psychological diagnosis;
- real-time emotional response inferred directly from classifier labels;
- micro-expression evidence.

## Suggested workflow position

A transparent workflow is:

1. import external face-analysis CSVs with `read_gazepoint_face_export()`;
2. standardise face-analysis columns with `standardize_gazepoint_face_columns()`;
3. audit face-data quality with `audit_gazepoint_face_quality()`;
4. synchronise face data with Gazepoint rows using `sync_gazepoint_face_data()`;
5. audit synchronisation quality with `audit_gazepoint_face_sync()`;
6. only then proceed to trial-window summaries, AOI-window summaries, or multimodal modelling.
