---
title: "External face-data reporting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{External face-data reporting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)

library(gp3tools)
```

This article demonstrates reporting helpers for external facial-behaviour workflows used alongside Gazepoint data.

These helpers do **not** infer facial expressions from Gazepoint CSV files. They also do **not** interpret facial behaviour as emotion. Their purpose is narrower: to create reviewer-facing reporting checklists and compact QC summaries for externally generated face-analysis data.

The current scope is:

1. document whether face-analysis input data are available;
2. check whether standardised face-analysis columns are present;
3. summarise face-data quality-audit readiness;
4. summarise synchronisation-audit readiness;
5. summarise window-summary and reactivity readiness;
6. summarise model-reporting readiness when models are used;
7. preserve explicit interpretation cautions.

## Example external face-analysis data

```{r}
face <- data.frame(
  participant_id = "P001",
  frame = 1:4,
  timestamp = c(0.000, 0.033, 0.066, 0.099),
  confidence = c(0.98, 0.96, 0.94, 0.92),
  success = c(1, 1, 1, 1),
  AU04_r = c(0.05, 0.06, 0.08, 0.09),
  AU12_r = c(0.20, 0.22, 0.25, 0.27),
  stringsAsFactors = FALSE
)

face_std <- standardize_gazepoint_face_columns(face)
```

## Quality audit

```{r}
quality_audit <- audit_gazepoint_face_quality(face_std)

quality_audit$overview
```

## Synchronisation audit

```{r}
gaze <- data.frame(
  participant_id = "P001",
  time_sec = c(0.000, 0.033, 0.066, 0.099),
  AOI = c("claim", "claim", "logo", "evidence"),
  stringsAsFactors = FALSE
)

synced <- sync_gazepoint_face_data(
  gazepoint_data = gaze,
  face_data = face,
  by = c(participant_id = "participant_id"),
  gaze_time_col = "time_sec",
  tolerance_sec = 0.050
)

sync_audit <- audit_gazepoint_face_sync(synced)

sync_audit$overview
```

## Window summaries

```{r}
windows <- data.frame(
  participant_id = "P001",
  window = c("baseline", "response"),
  window_start_sec = c(0.000, 0.066),
  window_end_sec = c(0.033, 0.120),
  stringsAsFactors = FALSE
)

face_windows <- summarize_gazepoint_face_windows(
  face_std,
  windows = windows,
  group_cols = "participant_id",
  window_label_col = "window",
  measure_cols = c("AU04_r", "AU12_r")
)

face_windows
```

## Reactivity summary

```{r}
face_reactivity <- summarize_gazepoint_face_reactivity(
  face_windows,
  baseline_window = "baseline",
  response_window = "response",
  group_cols = "participant_id",
  measure_cols = c("AU04_r", "AU12_r")
)

face_reactivity
```

## Optional model object

```{r}
model_data <- data.frame(
  AU12_r_mean = c(0.10, 0.20, 0.30, 0.40),
  rating = c(3, 4, 5, 6)
)

face_model <- fit_gazepoint_face_window_lmm(
  model_data,
  outcome = "rating",
  predictors = "AU12_r_mean"
)

face_model$formula
```

## Reporting checklist

`create_gazepoint_face_reporting_checklist()` creates a compact checklist for methods, supplements, or internal review.

```{r}
checklist <- create_gazepoint_face_reporting_checklist(
  face_data = face_std,
  quality_audit = quality_audit,
  sync_audit = sync_audit,
  window_summary = face_windows,
  reactivity_summary = face_reactivity,
  multimodal_model = face_model
)

checklist
```

The checklist records whether key workflow objects are present, what evidence is available, and what should be reported.

## Markdown QC report

`report_gazepoint_face_qc()` returns a compact markdown report by default.

```{r}
face_report <- report_gazepoint_face_qc(
  face_data = face_std,
  quality_audit = quality_audit,
  sync_audit = sync_audit,
  window_summary = face_windows,
  reactivity_summary = face_reactivity,
  multimodal_model = face_model
)

cat(paste(face_report, collapse = "\n"))
```

## Structured report list

The same information can be returned as a structured list.

```{r}
face_report_list <- report_gazepoint_face_qc(
  face_data = face_std,
  quality_audit = quality_audit,
  sync_audit = sync_audit,
  window_summary = face_windows,
  reactivity_summary = face_reactivity,
  multimodal_model = face_model,
  output = "list"
)

names(face_report_list)
```

```{r}
face_report_list$model_summary
```

## Recommended reporting language

Prefer cautious language such as:

- external facial-behaviour data;
- face-analysis confidence;
- valid face-analysis rows;
- synchronisation coverage;
- timing difference;
- window-level facial-behaviour summary;
- baseline-to-response facial-behaviour change;
- multimodal association.

Avoid unsupported language such as:

- true emotion detection;
- hidden affect;
- psychological diagnosis;
- micro-expression evidence;
- emotional state inferred directly from an algorithmic label;
- causal mechanism without appropriate design support.

## Suggested workflow position

A transparent workflow is:

1. import external face-analysis CSVs with `read_gazepoint_face_export()`;
2. standardise face-analysis columns with `standardize_gazepoint_face_columns()`;
3. audit face-data quality with `audit_gazepoint_face_quality()`;
4. synchronise face data with Gazepoint rows using `sync_gazepoint_face_data()`;
5. audit synchronisation quality with `audit_gazepoint_face_sync()`;
6. summarise facial-behaviour variables within analysis windows with `summarize_gazepoint_face_windows()`;
7. compute descriptive baseline-to-response changes with `summarize_gazepoint_face_reactivity()`;
8. prepare multimodal analysis tables with `prepare_gazepoint_multimodal_data()`;
9. fit explicit models where appropriate;
10. create checklist/reporting summaries with `create_gazepoint_face_reporting_checklist()` and `report_gazepoint_face_qc()`.
