---
title: "External face-data quality audit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{External face-data quality audit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  warning = FALSE,
  message = FALSE
)

library(gp3tools)
```

This article demonstrates quality-control helpers for external facial-behaviour exports.

These helpers do **not** infer facial expressions from Gazepoint CSV files. They audit externally generated face-analysis tables after import and standardisation. The intended inputs are CSV outputs from tools such as OpenFace-style, py-feat-style, MediaPipe-style, FaceReader-style, or generic frame-level facial-behaviour pipelines.

The current scope is:

1. audit face-detection validity;
2. summarise confidence and success indicators;
3. detect duplicate frame indices;
4. inspect basic timing continuity;
5. produce cautious status summaries and descriptive QC plots.

Synchronisation with Gazepoint timing, AOIs, trials, pupil, GSR, HR/IBI, or task phases is a later workflow stage.

## Example face-analysis data

```{r}
face <- data.frame(
  participant_id = rep(c("P001", "P002"), each = 4),
  frame = c(1, 2, 3, 4, 1, 2, 2, 4),
  timestamp = c(0.000, 0.033, 0.066, 0.099, 0.000, 0.033, 0.066, 0.500),
  confidence = c(0.98, 0.96, 0.94, 0.92, 0.95, 0.70, 0.40, 0.88),
  success = c(1, 1, 1, 1, 1, 1, 1, 1),
  AU04_r = c(0.05, 0.06, 0.05, 0.04, 0.10, 0.15, 0.20, 0.18),
  AU12_r = c(0.20, 0.22, 0.25, 0.24, 0.10, 0.12, 0.08, 0.09),
  stringsAsFactors = FALSE
)
```

## Audit quality

`audit_gazepoint_face_quality()` standardises unstandardised input when needed and returns overview, group-level, and issue-level summaries.

```{r}
face_audit <- audit_gazepoint_face_quality(
  face,
  group_cols = "participant_id",
  confidence_threshold = 0.80,
  min_valid_percent = 70,
  warning_valid_percent = 85,
  max_time_gap_sec = 0.20,
  max_duplicate_frame_percent = 0
)

face_audit$overview
```

```{r}
face_audit$group_summary
```

```{r}
face_audit$issue_summary
```

The object keeps the standardised data used for the audit.

```{r}
head(face_audit$data)
```

## Compact summary

`summarize_gazepoint_face_quality()` returns the overview table. The British spelling alias is also available.

```{r}
summarize_gazepoint_face_quality(face_audit)
```

```{r}
summarise_gazepoint_face_quality(face)
```

## Plot quality status

`plot_gazepoint_face_quality()` can visualise status counts across groups.

```{r}
plot_gazepoint_face_quality(
  face_audit,
  plot_type = "status",
  title = "Face-data quality status"
)
```

## Plot valid-row percentages

```{r}
plot_gazepoint_face_quality(
  face_audit,
  plot_type = "validity",
  title = "Valid face-analysis rows"
)
```

## Plot confidence

```{r}
plot_gazepoint_face_quality(
  face_audit,
  plot_type = "confidence",
  title = "Median face-detection confidence"
)
```

## Plot timing gaps

```{r}
plot_gazepoint_face_quality(
  face_audit,
  plot_type = "time_gaps",
  title = "Maximum face-data timing gap"
)
```

## Recommended interpretation

The audit reports data quality, not emotion validity. A group with high valid-row percentage and high confidence has better technical coverage for downstream facial-behaviour analysis. It does not prove that algorithmic emotion labels are psychologically valid.

Prefer cautious language such as:

- face-detection confidence;
- face-analysis coverage;
- valid facial-behaviour frames;
- action-unit availability;
- head-pose availability;
- timing continuity;
- facial-behaviour data quality.

Avoid unsupported language such as:

- true emotion detection;
- micro-expression evidence;
- hidden affect;
- psychological diagnosis;
- participant emotional state inferred directly from a classifier.

## Suggested workflow position

A transparent workflow is:

1. import external face-analysis CSVs with `read_gazepoint_face_export()`;
2. standardise timing, frame, confidence, and validity columns with `standardize_gazepoint_face_columns()`;
3. audit quality with `audit_gazepoint_face_quality()`;
4. review quality summaries and plots;
5. only then proceed to synchronisation with Gazepoint timing, trial windows, AOIs, pupil, GSR, HR/IBI, or task phases.
