---
title: "External face-export import"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{External face-export import}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)

library(gp3tools)
```

This article demonstrates the first-stage helpers for importing external facial-analysis exports into a `gp3tools` workflow.

These helpers do **not** infer facial expressions from Gazepoint CSV files. They are designed for CSV outputs produced by external tools such as OpenFace-style, py-feat-style, MediaPipe-style, FaceReader-style, or generic frame-level facial-behaviour pipelines.

The current intended scope is:

1. read external face-analysis CSV files;
2. detect or record the likely source format;
3. add file/session/participant metadata;
4. standardise common timing, frame, confidence, and validity columns.

Synchronisation with Gazepoint timing, event windows, AOIs, pupil, GSR, or HR/IBI should be handled in later workflow stages.

## Example external face-analysis export

Here we create a small OpenFace-style example. Real outputs may contain many more columns, including landmarks, head pose, gaze vectors, action-unit intensities, action-unit classifications, or software-specific quality indicators.

```{r}
face_csv <- tempfile(fileext = ".csv")

writeLines(
  c(
    "frame,timestamp,confidence,success,gaze_0_x,pose_Tx,pose_Rx,AU04_r,AU12_r,AU12_c",
    "1,0.000,0.98,1,0.10,1.5,0.10,0.05,0.20,1",
    "2,0.033,0.97,1,0.12,1.6,0.11,0.04,0.25,1",
    "3,0.066,0.45,1,0.15,1.7,0.12,0.07,0.30,1",
    "4,0.099,0.90,0,0.14,1.8,0.13,0.06,0.15,0"
  ),
  face_csv
)
```

## Read the external export

`read_gazepoint_face_export()` reads one file, several files, or a directory of CSV files. It adds metadata columns and attempts to detect the likely source format when `source = "auto"`.

```{r}
face_raw <- read_gazepoint_face_export(
  face_csv,
  participant_id = "P001",
  session_id = "S001"
)

face_raw
```

The returned object keeps the original external columns.

```{r}
class(face_raw)
```

```{r}
attr(face_raw, "gp3_face_settings")
```

## Standardise common columns

`standardize_gazepoint_face_columns()` adds common columns for later auditing and synchronisation.

```{r}
face_std <- standardize_gazepoint_face_columns(face_raw)

face_std[, c(
  "face_source",
  "face_file",
  "participant_id",
  "face_frame",
  "face_time_sec",
  "face_time_ms",
  "face_confidence",
  "face_success",
  "face_valid",
  "face_pose_tx",
  "face_pose_rx",
  "AU04_r",
  "AU12_r",
  "AU12_c"
)]
```

By default, `face_valid` uses a cautious rule based on available confidence and success fields. If both are present, a row is valid only when the face was successfully detected and confidence is at least the selected threshold.

```{r}
attr(face_std, "gp3_face_standardization")
```

## Generic CSVs

For non-standard exports, users can explicitly map the relevant columns.

```{r}
generic_face <- data.frame(
  subject = c("P001", "P001", "P001"),
  video_frame = c(10, 11, 12),
  seconds = c(1.00, 1.03, 1.06),
  score = c(0.95, 0.70, 0.88),
  detected = c("yes", "yes", "no"),
  smile = c(0.10, 0.20, 0.05),
  brow_raise = c(0.30, 0.25, 0.10),
  stringsAsFactors = FALSE
)

generic_std <- standardize_gazepoint_face_columns(
  generic_face,
  source = "generic",
  participant_id_col = "subject",
  frame_col = "video_frame",
  time_col = "seconds",
  confidence_col = "score",
  success_col = "detected",
  confidence_threshold = 0.80
)

generic_std
```

## Recommended interpretation

The standardised table should be interpreted as a **facial-behaviour data table**, not as direct access to internal emotional states.

Prefer cautious terms such as:

- facial action-unit intensity;
- face-detection confidence;
- head-pose estimate;
- smile-related movement;
- brow movement;
- facial-behaviour reactivity;
- algorithmic valence/arousal score, when supplied by an external tool.

Avoid unsupported claims such as:

- true emotion detection;
- micro-expression evidence;
- psychological diagnosis;
- participant emotional state inferred directly from a black-box label.

## Next workflow stages

The next stages, not covered in this article, are:

1. quality auditing of face-detection success, confidence, occlusion, and missing frames;
2. synchronisation with Gazepoint timing using participant ID, timestamp, frame, `CNT`, `TIMETICK`, TTL, or event logs;
3. aggregation into stimulus, AOI, trial, or task-phase windows;
4. cautious multimodal modelling with gaze, pupil, GSR, HR/IBI, and facial-behaviour features.
