---
title: "Fixation, transitions, and scanpaths"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fixation, transitions, and scanpaths}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)

library(gp3tools)
```

This article describes fixation-level workflows, AOI transitions, and scanpath-style summaries.

## Fixation alignment

```{r, eval=FALSE}
fixation_aligned <- prepare_gazepoint_fixation_aligned_data(
  fixation_data = fixations,
  trial_data = trial_summary,
  by = c('participant_id', 'trial_id')
)
```

## Transition matrices

```{r, eval=FALSE}
transition_matrix <- compute_gazepoint_transition_matrix(
  fixation_aligned,
  aoi_col = 'AOI',
  group_cols = c('participant_id', 'trial_id')
)

time_varying <- compute_gazepoint_time_varying_transition_matrix(
  fixation_aligned,
  aoi_col = 'AOI',
  time_col = 'time_sec',
  bin_width = 0.500
)
```

## Advanced modelling preparation

```{r, eval=FALSE}
semi_markov_data <- prepare_gazepoint_semimarkov_data(fixation_aligned)
hmm_data <- prepare_gazepoint_hmm_data(fixation_aligned)
transition_nb <- fit_gazepoint_transition_count_nb_sensitivity(time_varying)
```

## Interpretation note

Transitions describe observed sequences between AOIs. They should be interpreted as scanpath or visual-attention dynamics, not as direct proof of decision strategy.
