---
title: "Velocity Fixations, Blink Detection, and Blink-Aware Interpolation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Velocity Fixations, Blink Detection, and Blink-Aware Interpolation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4.5
)
library(gp3tools)
library(ggplot2)
```

## Purpose

This article demonstrates a transparent event-preprocessing branch for
sample-level gaze and pupil data:

1. smooth noisy gaze coordinates;
2. derive fixation events with a velocity threshold;
3. detect blink-like pupil intervals;
4. interpolate only short, internally bounded blink gaps.

The velocity threshold is expressed in scaled coordinate units per second.
When gaze coordinates are already expressed in visual degrees, `vmax` is in
degrees per second. For normalized or pixel coordinates, use `x_scale` and
`y_scale` when an appropriate conversion is available.

## Simulated sample trace

```{r simulate}
set.seed(2026)

time <- seq(0, 2.99, by = 0.01)
n <- length(time)

trace <- data.frame(
  USER_ID = "P01",
  trial = "T01",
  TIME = time,
  FPOGX = c(
    rep(0.25, 100),
    seq(0.25, 0.75, length.out = 10),
    rep(0.75, n - 110)
  ) + rnorm(n, 0, 0.003),
  FPOGY = 0.50 + rnorm(n, 0, 0.003),
  mean_pupil = 3.2 + 0.15 * sin(2 * pi * time / 3)
)

trace$mean_pupil[121:135] <- NA_real_
```

## Smooth gaze coordinates

```{r smooth}
smoothed <- smooth_gazepoint_coordinate(
  trace,
  method = "median",
  window = 5,
  group_cols = "trial"
)

ggplot(smoothed, aes(TIME)) +
  geom_line(aes(y = FPOGX, linewidth = "raw"), alpha = 0.45) +
  geom_line(aes(y = FPOGX_smooth, linewidth = "smoothed")) +
  scale_linewidth_manual(values = c(raw = 0.4, smoothed = 0.9)) +
  labs(
    x = "Time (s)",
    y = "Horizontal gaze coordinate",
    linewidth = NULL,
    title = "Raw and median-smoothed gaze coordinates"
  ) +
  theme_minimal()
```

## Detect velocity-defined fixations

```{r fixations}
fixation_result <- detect_gazepoint_fixations_velocity(
  smoothed,
  x_col = "FPOGX_smooth",
  y_col = "FPOGY_smooth",
  group_cols = "trial",
  vmax = 5,
  min_duration = 80,
  return = "both"
)

fixation_result$events
```

```{r fixation-plot}
fixation_samples <- fixation_result$samples

ggplot(fixation_samples, aes(TIME, FPOGX_smooth)) +
  geom_line() +
  geom_point(
    aes(shape = velocity_fixation),
    size = 1.4,
    alpha = 0.8
  ) +
  labs(
    x = "Time (s)",
    y = "Smoothed horizontal coordinate",
    shape = "Fixation",
    title = "Velocity-threshold sample labels"
  ) +
  theme_minimal()
```

## Detect blink intervals

```{r blinks}
blinks <- detect_gazepoint_blinks(
  trace,
  pupil_col = "mean_pupil",
  group_cols = "trial",
  min_duration = 50,
  merge_gap_ms = 20
)

blinks
```

## Interpolate eligible blink gaps

```{r interpolate}
interpolated <- interpolate_gazepoint_blinks(
  trace,
  blinks,
  pupil_cols = "mean_pupil",
  group_cols = "trial",
  method = "linear",
  max_gap_ms = 500
)

ggplot(interpolated, aes(TIME)) +
  geom_line(aes(y = mean_pupil, linetype = "observed")) +
  geom_line(
    aes(y = mean_pupil_blink_interp, linetype = "blink-interpolated")
  ) +
  labs(
    x = "Time (s)",
    y = "Pupil size",
    linetype = NULL,
    title = "Blink-aware pupil interpolation"
  ) +
  theme_minimal()
```

## Recommended reporting

Report the coordinate unit, conversion to visual angle when used, velocity
threshold, minimum fixation duration, blink definition, interpolation method,
maximum interpolated gap, and whether leading or trailing gaps remained
missing.
