---
title: "Gazepoint data quality: accuracy, precision, sampling, and loss"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gazepoint data quality: accuracy, precision, sampling, and loss}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = requireNamespace("eyeprocess", quietly = TRUE)
)
```

```{r}
library(gp3tools)
```

## Why this workflow exists

Gazepoint exports contain useful native validity and coordinate fields, but a
vendor summary is not a substitute for a reproducible study-level quality
analysis. The gp3tools wrappers in this article resolve Gazepoint columns and
metadata and then delegate every scientific formula to the vendor-neutral
eyeprocess core.

The workflow keeps four questions separate:

1. **Accuracy:** how far was gaze from a known target?
2. **Precision:** how stable was gaze while the intended target was stable?
3. **Sampling:** what timing stream was actually recorded?
4. **Availability:** how much usable gaze was available and how was it lost?

No wrapper automatically excludes a trial or participant.

## Native GP3 coordinates and units

Gazepoint BPOG/FPOG/LPOG/RPOG screen coordinates are treated as normalized
coordinates. TIME is interpreted as seconds and MSTIMER as milliseconds. For a
generic column such as gaze_x, the wrapper does not inspect the numeric range and
guess a unit: set coordinate_unit explicitly.

If you want angular results from normalized or pixel coordinates, provide
physical display geometry and viewing distance and request output_unit =
"degrees". Merely supplying geometry does not change units.

## Reproducible 9-point example

The following synthetic data mimic a validation/check-target stream with native
BPOG coordinates. The four profiles deliberately separate calibration offset,
spatial noise, irregular timing, and missingness.

```{r synthetic}
set.seed(20260918)

targets <- expand.grid(
  target_x = c(0.20, 0.50, 0.80),
  target_y = c(0.20, 0.50, 0.80)
)

profiles <- data.frame(
  profile = c(
    "good",
    "offset",
    "noisy",
    "irregular_missing"
  ),
  bias_x = c(0, 0.035, 0, 0),
  bias_y = c(0, -0.025, 0, 0),
  noise = c(0.003, 0.003, 0.018, 0.006)
)

rows <- list()
k <- 0L

for (p in seq_len(nrow(profiles))) {
  for (target_id in seq_len(nrow(targets))) {
    ms <- 0
    for (sample_id in seq_len(12)) {
      interval <- 1000 / 60
      if (profiles$profile[p] == "irregular_missing" &&
          sample_id %in% c(5, 10)) {
        interval <- interval * 2.5
      }
      ms <- ms + interval

      x <- targets$target_x[target_id] +
        profiles$bias_x[p] +
        rnorm(1, 0, profiles$noise[p])
      y <- targets$target_y[target_id] +
        profiles$bias_y[p] +
        rnorm(1, 0, profiles$noise[p])

      valid <- 1
      reason <- NA_character_
      if (profiles$profile[p] == "irregular_missing" &&
          sample_id %in% c(6, 7)) {
        x <- NA_real_
        y <- NA_real_
        valid <- 0
        reason <- if (sample_id == 6) "blink" else "tracker_invalidity"
      }

      k <- k + 1L
      rows[[k]] <- data.frame(
        USER = "P001",
        SESSION_ID = "S001",
        MEDIA_ID = paste0(profiles$profile[p], "_", target_id),
        profile = profiles$profile[p],
        target_id = target_id,
        MSTIMER = ms,
        BPOGX = x,
        BPOGY = y,
        BPOGV = valid,
        target_x = targets$target_x[target_id],
        target_y = targets$target_y[target_id],
        missing_reason = reason
      )
    }
  }
}

gp_validation <- do.call(rbind, rows)
```

## Build the quality report

For normalized coordinates, the report can stay normalized or be converted to
degrees when geometry is available. This example uses a 24-inch-class
1920-by-1080 display and an explicitly declared viewing distance.

```{r report}
geometry <- list(
  screen_width_px = 1920,
  screen_height_px = 1080,
  screen_width_cm = 53.0,
  screen_height_cm = 29.8,
  viewing_distance_cm = 60
)

quality <- create_gazepoint_quality_report(
  gp_validation,
  missing_reason_col = "missing_reason",
  level = "trial",
  output_unit = "degrees",
  geometry = geometry,
  nominal_sampling_hz = 60,
  bcea_probability = 0.68,
  preprocessing_spec = "synthetic raw validation samples; no interpolation",
  quality_rules = "descriptive QC with study-defined review flags"
)

quality[c(
  ".gp3_quality_participant",
  ".gp3_quality_trial",
  "accuracy_mean",
  "precision_rms_s2s",
  "precision_sd",
  "bcea",
  "effective_sampling_hz",
  "valid_sample_fraction",
  "data_loss_fraction"
)]
```

## Accuracy and precision are different

The offset profile should show worse target-referenced accuracy while retaining
relatively good short-term precision. The noisy profile is centered on the
target on average but has poorer precision. This is why a single "calibration
quality" label cannot replace explicit accuracy and precision metrics.

```{r accuracy-precision}
aggregate(
  cbind(accuracy_mean, precision_rms_s2s, precision_sd) ~ profile,
  transform(
    quality,
    profile = sub("_[0-9]+$", "", .gp3_quality_trial)
  ),
  mean,
  na.rm = TRUE
)
```

### Aggregation levels

The convenience layer supports dataset, sample, participant, session, trial, and
file grouping. Sample level uses a deterministic source-row identifier. It is
useful for tracing validity and target error, but a single sample does not
provide a meaningful stable-target estimate of RMS-S2S, SD precision, or BCEA.

## Review thresholds are not exclusions

Thresholds are study-specific and should normally be defined before inspecting
outcomes. They create quality_flags and review_required only.

```{r thresholds}
reviewed <- create_gazepoint_quality_report(
  gp_validation,
  missing_reason_col = "missing_reason",
  level = "trial",
  output_unit = "degrees",
  geometry = geometry,
  nominal_sampling_hz = 60,
  thresholds = list(
    accuracy_mean = list(max = 1.0),
    valid_sample_fraction = list(min = 0.80)
  )
)

reviewed[c(
  ".gp3_quality_trial",
  "quality_flags",
  "review_required"
)]

attr(reviewed, "gazepoint_quality_adapter")$automatic_exclusion
```

## Visual dashboard

```{r dashboard, fig.width=9, fig.height=7}
plot_gazepoint_quality_dashboard(quality)
```

The dashboard is descriptive. It should be paired with the row-level quality
table and with the study's pre-specified review/exclusion logic.

## Focused plot gallery

Use a single validation target for compact visual comparisons so spatial
precision is not contaminated by intentional target-to-target gaze movement.

```{r focused-plot-data}
quality_center <- quality[grepl("_5$", quality$.gp3_quality_trial), ]
```

### Target-referenced accuracy

```{r focused-accuracy, fig.width=8, fig.height=4.8}
eyeprocess::plot_gaze_accuracy(quality_center)
```

Higher values mean larger error from the known target. A stable calibration
offset can therefore be precise but inaccurate.

### RMS sample-to-sample precision

```{r focused-precision, fig.width=8, fig.height=4.8}
eyeprocess::plot_gaze_precision(quality_center)
```

RMS-S2S describes short-term fluctuation during stable gaze. It is not a
substitute for accuracy, and it should not be interpreted across target
transitions or intended saccades.

### BCEA

```{r focused-bcea, fig.width=8, fig.height=4.8}
eyeprocess::plot_bcea(quality_center)
```

BCEA is an area-based dispersion summary. Report its probability level and area
unit; it is not target-referenced accuracy.

### Sampling intervals

```{r focused-sampling, fig.width=8, fig.height=4.8}
irregular_center <- gp_validation[
  gp_validation$profile == "irregular_missing" &
    gp_validation$target_id == 5,
]
eyeprocess::plot_sampling_intervals(
  irregular_center,
  time = "MSTIMER",
  time_unit = "ms"
)
```

The focused plots are provided by the vendor-neutral core. gp3tools keeps the
Gazepoint-specific convenience dashboard as a thin adapter rather than
duplicating plotting/statistical logic.

For timing diagnostics, distinguish `long_interval_count` from
`dropped_interval_count`: the first counts unusually long observed intervals,
whereas the second estimates how many nominal samples those long gaps represent.
Neither should be described as a direct hardware-level packet-loss measurement.

## A failure case worth keeping

Generic standardized columns require an explicit unit. This deliberately fails:

```{r generic-unit-failure, error=TRUE}
generic <- data.frame(
  timestamp_ms = c(0, 16.7, 33.4),
  gaze_x = c(0.50, 0.51, 0.50),
  gaze_y = c(0.50, 0.50, 0.50)
)

create_gazepoint_quality_report(generic)
```

The correct call declares the unit:

```{r generic-unit-correct}
create_gazepoint_quality_report(
  generic,
  coordinate_unit = "normalized",
  target_x_col = NULL,
  target_y_col = NULL,
  nominal_sampling_hz = 60
)
```

## Quality decisions as sensitivity analysis

Keep the full Gazepoint quality report intact and separate the measurement
diagnostic from the eventual inclusion decision.

```{r sensitivity}
keep <- reviewed[!reviewed$review_required, ]
flagged <- reviewed[reviewed$review_required, ]
```

For downstream modeling, join the report back to the trial or participant key
and compare the primary specification with one or more pre-specified sensitivity
specifications. This can include a stricter exclusion rule or stratification by
quality status. The wrapper never removes observations for you.

## Reporting guideline

For a manuscript, report the display/coordinate geometry needed to understand
the unit, the validation/check-target procedure, the accuracy statistic, the
precision definition, the BCEA probability when used, nominal and observed
sampling behavior, the operational definition of data loss, and the actual
review/exclusion rule.

A compact starting point is:

```{r reporting}
report_gazepoint_quality(quality)
```

A manuscript-ready sentence can then add the acquisition context:

> Gaze quality was evaluated from known validation targets using target-referenced
> Euclidean accuracy, RMS sample-to-sample and spatial-SD precision, and 68% BCEA
> after explicit conversion from normalized screen coordinates to degrees of
> visual angle. The nominal 60-Hz device rate was supplemented by timestamp-based
> effective sampling and interval diagnostics. Missing or invalid gaze was
> summarized by loss fraction and loss-run duration. Pre-specified thresholds
> generated manual-review flags rather than automatic exclusions.

## Interpretation boundaries

These measures characterize measurement quality. They do not establish
attention, engagement, cognitive load, motivation, competence, or any clinical
property. A review flag is an analysis-governance signal, not a diagnosis or a
participant label.

## API links

- summarise_gazepoint_spatial_quality()
- create_gazepoint_quality_report()
- plot_gazepoint_quality_dashboard()
- report_gazepoint_quality()

The underlying formulas and scientific contracts are maintained in eyeprocess;
gp3tools remains the Gazepoint-specific adapter layer.
