---
title: "Gazepoint worked example: evidence verification latency"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gazepoint worked example: evidence verification latency}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Purpose

This article shows the thin `gp3tools` route for **time to first source/evidence inspection**. It does not implement a second survival engine: all event/censor construction, model fitting, diagnostics, and reporting are delegated to `eyeprocess`.

## Build the synthetic verification dataset

```{r raw, eval=requireNamespace("eyeprocess", quietly=TRUE)}
raw <- eyeprocess::simulate_gaze_survival_inputs(
  "verification",
  seed = 20260918,
  n_participants = 36,
  trials_per_participant = 3
)
```

## Run the explicit adapter workflow

```{r analysis, eval=requireNamespace("eyeprocess", quietly=TRUE)}
verification <- run_gazepoint_latency_analysis(
  raw$trials,
  raw$events,
  target_aoi = "source_evidence",
  formula = "condition",
  participant_col = "participant_id",
  cox_structure = "cluster_robust",
  aft_distribution = "weibull",
  km_group = "condition",
  event_type = "first_aoi_entry",
  condition_col = "condition_id",
  event_detector = "synthetic_truth",
  aoi_specification = "fixed synthetic source/evidence AOI",
  quality_rules = list(valid_fraction_min = .90)
)

verification$censoring
verification$report$effects
verification$ph_diagnostics
```

`cox_structure` and `aft_distribution` are required deliberately. `gp3tools` never chooses a repeated-participant estimator or AFT family for the analyst.

## Visualize time to evidence inspection

```{r plot, eval=requireNamespace("eyeprocess", quietly=TRUE), fig.cap="Kaplan-Meier curves for time to first source/evidence AOI entry."}
eyeprocess::plot_gaze_survival_curve(
  verification$data,
  group = "condition"
)
```

A lower survival curve indicates that a larger share of trials has already entered the source/evidence AOI by that time.

## Complementary verification plots

```{r incidence, eval=requireNamespace("eyeprocess", quietly=TRUE), fig.cap="Single-event 1-KM view for evidence inspection."}
eyeprocess::plot_gaze_cumulative_incidence(
  verification$data,
  group = "condition"
)
```

```{r hazard, eval=requireNamespace("eyeprocess", quietly=TRUE), fig.cap="Empirical event/risk increments for evidence inspection."}
eyeprocess::plot_gaze_hazard(
  verification$data,
  group = "condition"
)
```

Use the 1-KM plot only for the single target event represented here. It is not a competing-risks cumulative-incidence estimator when mutually exclusive event types compete. The hazard panel is an empirical event/risk diagnostic, not a smooth latent hazard estimate.

## Interpretation

The clustered Cox hazard ratio describes the relative instantaneous evidence-inspection rate among trials still at risk. It is not a mean-latency ratio. The delegated Weibull AFT output instead expresses a multiplicative event-time contrast under the named distribution.

## Troubleshooting clinic

Do not silently convert any of the following into ordinary right censoring:

- an incomplete trial observation window;
- missing or ambiguous trial/event identity;
- gaze quality below the declared rule;
- event time after the censoring limit.

If event counts are sparse or censoring is extreme, report that limitation even if the backend converges. If the scientific model requires latent participant frailty, request `cox_structure = "frailty"` explicitly and report the corresponding marginal Cox PH diagnostic separately.

## Reporting example

> Gazepoint evidence-inspection latency was analysed through the vendor-neutral `eyeprocess` survival engine using `gp3tools`. Complete usable trials without a source/evidence entry were retained as right-censored observations; unresolved or poor-quality trials were not converted to censoring. We fitted a participant-clustered Cox model and a pre-specified Weibull AFT sensitivity model, reporting effect estimates with 95% confidence intervals together with censoring summaries and proportional-hazards diagnostics.

Report both package versions, the target event/AOI, time origin, observation-window rule, quality criterion, Cox structure, AFT family, event/censor counts, diagnostics, and sensitivity branches.

## API links

- `prepare_gazepoint_survival_data()` — thin preparation adapter.
- `run_gazepoint_latency_analysis()` — thin end-to-end adapter.
- `eyeprocess::plot_gaze_survival_curve()` — vendor-neutral survival visualization.
- `eyeprocess::report_gaze_survival_model()` — vendor-neutral reporting bundle.
