---
title: "gp3tools to gpbiometrics workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gp3tools to gpbiometrics workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(gp3tools)
```

## Workflow

The integration contract is:

```text
gp3tools gaze import and QC
  -> standardized gaze/master output
  -> gpbiometrics biometric import and preprocessing
  -> timestamp alignment
  -> AOI/event-contingent biometric summaries
  -> combined audit and report
```

The bridge does not infer emotion, stress, preference, cognition,
comprehension, health status, or diagnosis.

## Prepare gp3tools gaze output

```{r}
gaze <- data.frame(
  USER_ID = rep("P01", 6),
  MEDIA_ID = rep("T01", 6),
  MSTIMER = seq(0, 100, by = 20),
  BPOGX = c(0.2, 0.3, 0.4, 0.7, 0.8, 0.9),
  BPOGY = c(0.4, 0.4, 0.5, 0.5, 0.6, 0.6),
  BPOGV = 1,
  AOI = c("claim", "claim", "claim", "evidence", "evidence", "evidence"),
  LPD = c(3.10, 3.12, 3.11, 3.15, 3.16, 3.17)
)

gaze_bridge <- prepare_gazepoint_gpbiometrics_bridge(gaze)
gaze_bridge
```

In a real project, `gaze` would normally be produced by
`read_gazepoint_folder()` and `create_gazepoint_master()` after gp3tools QC.

## Connect biometric output

The following data frame represents the stable table returned after
gpbiometrics import and preprocessing:

```{r}
biometrics <- data.frame(
  participant_id = rep("P01", 6),
  trial_id = rep("T01", 6),
  time_s = seq(0.001, 0.101, by = 0.02),
  GSR = c(1.00, 1.02, 1.03, 1.06, 1.05, 1.07),
  HR = c(70, 70, 71, 72, 72, 73),
  event = c("trial_start", "", "", "cta", "", "trial_end")
)
```

A real gpbiometrics import can be substituted directly:

```{r, eval=FALSE}
biometrics <- getExportedValue(
  "gpbiometrics",
  "import_gazepoint_biometrics"
)("path/to/biometric_export.csv")
```

## Synchronize and summarise

```{r}
workflow <- run_gazepoint_gpbiometrics_workflow(
  gaze_bridge,
  biometrics,
  signal_cols = c("GSR", "HR"),
  event_col = "event",
  tolerance_s = 0.01
)

workflow
workflow$audit
workflow$signal_summary
```

The native engine performs an explicit nearest-timestamp match within
participant and trial. A study-specific or gpbiometrics adapter can be supplied
through the `adapter` argument and is tested through the same return contract.

## Combined report

```{r}
report <- create_gazepoint_cross_package_report(workflow)
cat(report, sep = "\n")
```

The report records match coverage and timing error before any substantive
interpretation.
