---
title: "Heatmaps and spatial visualisation"
---

```{r setup, include=FALSE}
library(gp3tools)
```

This article gives compact examples for creating gaze and fixation heatmaps from
Gazepoint-style coordinate data. The examples use synthetic data and are intended
to demonstrate plotting behaviour rather than to evaluate visual attention.

## Example data

```{r}
set.seed(42)

n <- 250

gaze <- data.frame(
  x = pmin(pmax(c(
    stats::rnorm(n / 2, mean = 0.35, sd = 0.08),
    stats::rnorm(n / 2, mean = 0.68, sd = 0.07)
  ), 0), 1),
  y = pmin(pmax(c(
    stats::rnorm(n / 2, mean = 0.42, sd = 0.08),
    stats::rnorm(n / 2, mean = 0.58, sd = 0.07)
  ), 0), 1),
  duration = stats::runif(n, min = 60, max = 350)
)

head(gaze)
```

## Preparing heatmap data

```{r}
prepared <- prepare_gazepoint_heatmap_data(
  gaze,
  x_col = "x",
  y_col = "y",
  weight_col = "duration",
  display_width = 1280,
  display_height = 720
)

head(prepared[, c(".gp3_x_px", ".gp3_y_px", ".gp3_weight")])
```

## Native heatmap

```{r}
plot_gazepoint_heatmap(
  prepared,
  bins = 45,
  alpha = 0.85
)
```

## Heatmap with raw points

```{r}
plot_gazepoint_heatmap(
  prepared,
  bins = 45,
  alpha = 0.80,
  show_points = TRUE
)
```

## Background-image overlay

`plot_gazepoint_heatmap_overlay()` can place the heatmap over a PNG stimulus
image. The background-image helper uses the optional `png` package.

```{r}
if (requireNamespace("png", quietly = TRUE)) {
  bg <- tempfile(fileext = ".png")

  img <- array(1, dim = c(720, 1280, 3))
  img[150:570, 120:520, ] <- 0.92
  img[220:520, 720:1120, ] <- 0.86

  png::writePNG(img, bg)

  plot_gazepoint_heatmap_overlay(
    gaze,
    background_image = bg,
    x_col = "x",
    y_col = "y",
    weight_col = "duration",
    display_width = 1280,
    display_height = 720,
    bins = 45,
    heatmap_alpha = 0.70
  )
}
```

## Exporting a heatmap

```{r eval=FALSE}
p <- plot_gazepoint_heatmap(
  prepared,
  bins = 45
)

export_gazepoint_heatmap_png(
  p,
  filename = "gazepoint_heatmap.png",
  width = 8,
  height = 5,
  dpi = 300
)
```
