---
title: "Large-export performance benchmarking"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Large-export performance benchmarking}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(gp3tools)
```

## Purpose

This workflow measures elapsed time, approximate R-heap growth, output size,
and empirical scaling across increasing Gazepoint export sizes. It is a
regression framework rather than a hardware-independent speed claim.

```{r}
gp3tools_performance_limits()
```

The default limits are deliberately conservative. Projects can supply a
machine-specific baseline and stricter ratio limits after collecting stable
measurements on the same system.

## Fast contract run

A small run verifies the complete benchmark contract without slowing package
checks.

```{r}
small <- benchmark_gazepoint_export_performance(
  scales = data.frame(
    total_rows = c(1000L, 5000L),
    n_files = c(1L, 2L)
  ),
  operations = "generate",
  trials = 1L
)

small$summary
small$regression$overall
```

## Large-export profile

The installed script uses 60,000, 240,000, and 960,000 rows distributed across
1, 4, and 16 files, respectively. Run it outside ordinary R CMD checks:

```{r, eval=FALSE}
source(
  system.file(
    "benchmarks",
    "run-large-export-performance.R",
    package = "gp3tools"
  )
)
```

On Windows CMD, set `GP3TOOLS_BENCHMARK_OUTPUT` to an external folder before
running the script. Benchmark outputs should not be committed when they contain
machine-specific paths or private export data.

## Baseline-relative regression checks

A saved result can be used as a baseline:

```{r, eval=FALSE}
current_audit <- check_gazepoint_performance_regression(
  current_benchmark,
  baseline = previous_benchmark,
  elapsed_ratio_limit = 1.25,
  memory_ratio_limit = 1.25
)
```

Absolute and baseline-relative failures remain visible in the audit table.
