---
title: "Metric extraction and model-boundary guidance"
description: "Extract eye-tracking and pupil metrics while keeping modelling and interpretation aligned with what the measures support."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.retina = 2
)
```


This article presents a conservative metric-to-model workflow:

1. define the raw observation unit;
2. extract a transparent metric;
3. inspect its scale and distribution;
4. select a compatible statistical family;
5. account for the study's hierarchical structure;
6. keep interpretation within the limits of the measure.

```{r load-package}
library(gp3tools)

set.seed(202)
```

## Start from the metric rather than the preferred model

```{r metric-family-table}
families <- recommend_gazepoint_model_family()

families[
  families$metric %in% c(
    "fixation_duration",
    "dwell_time",
    "fixation_count",
    "aoi_proportion",
    "scanpath_length",
    "convex_hull_area",
    "pupil_timecourse"
  ),
  ,
  drop = FALSE
]
```

The recommendation table does not remove the need to inspect the observed
distribution. A duration outcome can contain structural zeros, truncation,
censoring, or participant-specific scales that require additional decisions.

## Extract simple AOI dwell summaries

The following example simulates sample-level AOI assignments. Each row
represents a 16-ms sample.

```{r simulate-aoi-samples}
samples <- expand.grid(
  subject = paste0(
    "S",
    sprintf("%02d", 1:10)
  ),
  trial = 1:4,
  sample = 1:80
)

samples$condition <- ifelse(
  samples$trial %% 2 == 0,
  "specific_claim",
  "vague_claim"
)

samples$time_ms <-
  samples$sample * 16

samples$aoi <- sample(
  c(
    "source_label",
    "claim_text",
    "evidence_panel",
    "other"
  ),
  size = nrow(samples),
  replace = TRUE,
  prob = c(
    0.20,
    0.38,
    0.22,
    0.20
  )
)

head(samples)
```

AOI dwell is calculated as the number of assigned samples multiplied by the
sample interval.

```{r calculate-aoi-dwell}
samples$sample_duration_ms <- 16

aoi_dwell <- aggregate(
  sample_duration_ms ~
    subject +
    trial +
    condition +
    aoi,
  data = samples,
  FUN = sum
)

names(aoi_dwell)[
  names(aoi_dwell) == "sample_duration_ms"
] <- "dwell_ms"

head(aoi_dwell)
```

```{r aoi-dwell-plot}
claim_dwell <- aoi_dwell[
  aoi_dwell$aoi == "claim_text",
  ,
  drop = FALSE
]

boxplot(
  dwell_ms ~ condition,
  data = claim_dwell,
  xlab = "Condition",
  ylab = "Claim-AOI dwell time (ms)",
  main = "Example extracted AOI dwell feature"
)

stripchart(
  dwell_ms ~ condition,
  data = claim_dwell,
  vertical = TRUE,
  method = "jitter",
  pch = 16,
  add = TRUE
)
```

A difference in claim-AOI dwell supports a statement about time allocated to
that AOI under the specified denominator and inclusion rules. It does not, by
itself, establish deeper scrutiny, comprehension, skepticism, memory, or
critical evaluation.

## Scanpath geometry is a descriptive movement feature

```{r simulate-scanpaths}
set.seed(303)

scanpath <- do.call(
  rbind,
  lapply(
    1:12,
    function(trial_id) {
      n <- 30

      data.frame(
        subject = paste0(
          "S",
          sprintf(
            "%02d",
            ceiling(trial_id / 3)
          )
        ),
        trial = trial_id,
        condition = ifelse(
          trial_id %% 2 == 0,
          "specific_claim",
          "vague_claim"
        ),
        time_ms = seq_len(n) * 40,
        x = 450 +
          cumsum(
            rnorm(
              n,
              mean = ifelse(
                trial_id %% 2 == 0,
                8,
                12
              ),
              sd = 10
            )
          ),
        y = 320 +
          cumsum(
            rnorm(
              n,
              mean = 2,
              sd = 8
            )
          )
      )
    }
  )
)
```

```{r extract-scanpath-geometry}
scan_features <- compute_gazepoint_scanpath_geometry(
  data = scanpath,
  x = "x",
  y = "y",
  subject = "subject",
  trial = "trial",
  time = "time_ms",
  condition = "condition"
)

head(scan_features)
```

```{r example-scanpath-plot}
trial_one <- scanpath[
  scanpath$trial == 1,
  ,
  drop = FALSE
]

plot(
  trial_one$x,
  trial_one$y,
  type = "b",
  pch = 16,
  xlab = "Screen x-coordinate",
  ylab = "Screen y-coordinate",
  main = "Example scanpath geometry"
)

text(
  trial_one$x[
    c(
      1,
      nrow(trial_one)
    )
  ],
  trial_one$y[
    c(
      1,
      nrow(trial_one)
    )
  ],
  labels = c(
    "Start",
    "End"
  ),
  pos = c(
    2,
    4
  )
)
```

```{r scanpath-length-plot}
boxplot(
  scanpath_length ~ condition,
  data = scan_features,
  xlab = "Condition",
  ylab = "Scanpath length",
  main = "Example scanpath-length feature"
)

stripchart(
  scanpath_length ~ condition,
  data = scan_features,
  vertical = TRUE,
  method = "jitter",
  pch = 16,
  add = TRUE
)
```

Scanpath length, convex-hull area, dispersion, and efficiency describe spatial
movement or search extent. Whether a longer path indicates exploration,
inefficiency, uncertainty, task difficulty, or another process depends on the
task and complementary evidence.

## Pupil-response features require explicit baseline decisions

```{r simulate-boundary-pupil}
set.seed(404)

pupil <- expand.grid(
  subject = paste0(
    "S",
    sprintf("%02d", 1:8)
  ),
  trial = 1:4,
  time_ms = seq(
    -200,
    900,
    by = 50
  )
)

pupil$condition <- ifelse(
  pupil$trial %% 2 == 0,
  "specific_claim",
  "vague_claim"
)

pupil$pupil_mm <-
  3 +
  ifelse(
    pupil$condition == "vague_claim",
    0.10,
    0.04
  ) *
    exp(
      -((pupil$time_ms - 400) / 250)^2
    ) +
  rnorm(
    nrow(pupil),
    sd = 0.02
  )
```

```{r extract-boundary-pupil-features}
pupil_features <- summarize_gazepoint_pupil_response_features(
  data = pupil,
  pupil = "pupil_mm",
  time = "time_ms",
  subject = "subject",
  trial = "trial",
  baseline_window = c(
    -200,
    0
  ),
  response_window = c(
    0,
    900
  ),
  condition = "condition"
)

head(pupil_features)
```

```{r pupil-feature-association}
plot(
  pupil_features$peak_dilation,
  pupil_features$auc,
  pch = 16,
  xlab = "Peak dilation",
  ylab = "Pupil-response AUC",
  main = "Relationship between extracted pupil features"
)

abline(
  stats::lm(
    auc ~ peak_dilation,
    data = pupil_features
  ),
  lty = 2
)
```

Peak dilation and area under the curve summarize different aspects of the same
response trajectory. Their association does not make them interchangeable.
The baseline window, response window, interpolation procedure, smoothing,
luminance conditions, and missingness rules should be reported.

## Metric-to-interpretation boundary table

```{r interpretation-boundary-table}
interpretation_boundary <- data.frame(
  metric = c(
    "Fixation count",
    "Fixation duration",
    "AOI dwell time",
    "Transition count",
    "Scanpath length",
    "Pupil dilation",
    "Blink rate"
  ),
  directly_supports = c(
    "Number of detected fixations under the stated detection rule",
    "Duration of detected fixation events",
    "Time allocated to the specified AOI",
    "Observed movement between defined AOIs",
    "Spatial length of the observed gaze path",
    "Change in measured pupil size under the stated preprocessing",
    "Frequency of detected blink events per stated exposure"
  ),
  does_not_independently_establish = c(
    "Interest, comprehension, or preference",
    "Deep processing or difficulty",
    "Critical scrutiny or belief revision",
    "A cognitive strategy or causal information integration",
    "Confusion, exploration, or search inefficiency",
    "A specific emotion, cognitive load, or arousal mechanism",
    "Fatigue, stress, or disengagement"
  ),
  stringsAsFactors = FALSE
)

knitr::kable(
  interpretation_boundary,
  align = c(
    "l",
    "l",
    "l"
  )
)
```

## Model-boundary checklist

A defensible metric-to-model workflow should document:

- whether the input consists of raw samples, fixations, saccades, blinks, or
  trial-level summaries;
- the event-detection algorithm and thresholds;
- AOI definitions and overlap rules;
- sample interval or event-duration calculation;
- baseline and response windows;
- blink, artifact, interpolation, smoothing, and track-loss rules;
- denominator definitions for proportions and rates;
- the participant, item, trial, screen, and order structure;
- whether the outcome is positive continuous, count, bounded, binary, or a
  time course;
- the model family and link function;
- random effects or smooth terms;
- convergence, residual, overdispersion, and sensitivity diagnostics;
- the distinction between confirmatory, sensitivity, and exploratory results;
- the interpretation directly supported by the measurement.

## Conservative reporting examples

Prefer:

> Participants allocated more dwell time to the claim AOI in condition A than
> in condition B.

Avoid unless independently supported:

> Participants scrutinized the claim more critically in condition A.

Prefer:

> The pupil response was larger during the prespecified response window after
> baseline correction.

Avoid unless the design and complementary measures isolate the mechanism:

> Condition A produced greater emotional arousal.

Prefer:

> Scanpaths were longer and covered a larger spatial area.

Avoid without task-specific validation:

> Participants were more confused.

## Final boundary

Eye tracking and pupillometry provide valuable process measures. Their
strength lies in measuring visual allocation, temporal dynamics, event
sequences, spatial movement, and pupil change.

The most defensible workflow extracts those quantities transparently, models
their actual statistical scale, reports preprocessing and uncertainty, and
avoids converting an observed process measure into an unsupported latent
mental state.
