---
title: "Missingness and data-coverage reporting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Missingness and data-coverage reporting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  warning = FALSE,
  message = FALSE
)

library(gp3tools)
```

This article demonstrates compact helpers for summarising, plotting, and reporting missingness in Gazepoint-style data. The helpers are intended for transparent data-coverage review. They do not define exclusion rules by themselves.

## Simulate example data

```{r}
synthetic <- simulate_gazepoint_pupil_data(
  n_subjects = 4,
  n_trials = 4,
  n_time_bins = 20,
  conditions = c("control", "treatment"),
  blink_probability = 0.08,
  seed = 123
)

head(synthetic)
```

## Summarise missingness

`summarize_gazepoint_missingness()` computes missing and observed rates for selected variables, optionally within grouping columns.

```{r}
missingness <- summarize_gazepoint_missingness(
  synthetic,
  cols = c("pupil_left", "pupil_right", "pupil", "gaze_x", "gaze_y"),
  group_cols = "condition"
)

missingness
```

The British spelling alias is also available.

```{r}
summarise_gazepoint_missingness(
  synthetic,
  cols = c("pupil_left", "pupil_right", "pupil")
)
```

## Plot missingness profiles

`plot_gazepoint_missingness_profile()` can use either raw data or a missingness summary.

```{r}
plot_gazepoint_missingness_profile(
  missingness,
  plot_type = "bar",
  title = "Missingness by condition and variable"
)
```

A tile view can be useful when there are many groups or variables.

```{r}
plot_gazepoint_missingness_profile(
  missingness,
  plot_type = "tile",
  title = "Missingness profile"
)
```

## Generate cautious report text

`report_gazepoint_missingness()` returns the summary table, an overall table, a variable-level table, and compact report text.

```{r}
missingness_report <- report_gazepoint_missingness(
  missingness,
  digits = 1,
  max_variables = 3
)

missingness_report$overall
```

```{r}
missingness_report$variable_summary
```

```{r}
missingness_report$report_text
```

The generated wording is deliberately cautious. Missingness rates are data-coverage diagnostics and should be interpreted together with study design, preprocessing decisions, and planned exclusion criteria.

## Suggested workflow

A transparent reporting workflow is:

1. Summarise missingness for pupil, gaze, AOI, or trial-level variables.
2. Plot missingness by participant, trial, condition, AOI, or preprocessing stage.
3. Report overall and variable-level missingness rates.
4. Keep exclusion decisions separate from descriptive missingness summaries.
