---
title: "Model-readiness and sensitivity analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model-readiness and sensitivity analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)

library(gp3tools)
```

This article summarises model-readiness checks and sensitivity workflows.

## Model-readiness principle

Before modelling, check whether the analysis table contains the expected participant, trial, condition, outcome, predictor, and covariate columns. Also inspect missingness, group sizes, influential observations, and scale compatibility.

## Example workflow

```{r, eval=FALSE}
model_data <- prepare_gazepoint_multimodal_data(
  face_windows = face_windows,
  gaze_data = gaze_summary,
  response_data = response_data,
  by = c('participant_id', 'trial_id'),
  predictors = c('AU12_r_mean', 'claim_dwell_ms'),
  outcome = 'rating'
)

fit <- fit_gazepoint_multimodal_response_model(
  model_data,
  outcome = 'rating',
  predictors = c('AU12_r_mean', 'claim_dwell_ms'),
  covariates = c('trial_order')
)

loo <- run_gazepoint_model_leave_one_out(fit)
nested <- compare_gazepoint_nested_models(fit_reduced, fit_full)
pred <- plot_gazepoint_model_predictions(fit)
```

## Sensitivity examples

- leave-one-participant-out or leave-one-trial-out checks;
- nested model comparisons;
- robustness to covariates;
- robustness to exclusion thresholds;
- robustness to alternative AOI or baseline definitions.

## Reporting note

Sensitivity analyses should be described as robustness checks. They should not be used to selectively report only favourable specifications.
