---
title: "User-facing naming policy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{User-facing naming policy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(gp3tools)
```

## Canonical spelling

gp3tools uses British English `summarise_*` names as the canonical spelling for
new user-facing summary helpers.

```{r}
gp3tools_naming_policy()
```

Existing exported American English `summarize_*` functions remain available for
backward compatibility. The development installer creates a British alias when
an established American export does not yet have one. No existing user code is
removed or silently redirected to a different analytical implementation.

## Audit the exported API

```{r}
naming_audit <- audit_gazepoint_naming_consistency()
naming_audit
naming_audit$pairs
```

The audit distinguishes:

- `paired`: both spellings are exported;
- `canonical_only`: only the British canonical name exists;
- `missing_british_alias`: an established American export lacks its British
  counterpart.

The final development state should contain no `missing_british_alias` rows.
