---
title: "Task-phase segmentation and coverage reporting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Task-phase segmentation and coverage reporting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  warning = FALSE,
  message = FALSE
)

library(gp3tools)
```

This article demonstrates helpers for assigning Gazepoint-style samples to declared task phases and summarising phase-level data coverage. These helpers are descriptive. They do not infer experimental phases, define exclusion rules, or replace study-specific preprocessing decisions.

## Example data and phase windows

A typical eye-tracking task may include baseline, stimulus, response, and recovery periods. The phase windows should be declared explicitly by the analyst.

```{r}
gaze_data <- data.frame(
  subject = rep(c("S1", "S2"), each = 12),
  trial = rep(rep(1:2, each = 6), times = 2),
  condition = rep(c("control", "treatment"), each = 12),
  time_ms = rep(c(0, 250, 750, 1250, 1750, 2250), times = 4),
  pupil = c(
    3.1, 3.0, 3.2, 3.3, NA, 3.2,
    3.0, 3.1, 3.3, 3.4, 3.5, 3.4,
    3.2, NA, 3.4, 3.5, 3.6, 3.5,
    3.1, 3.2, 3.3, NA, 3.4, 3.3
  )
)

phase_windows <- data.frame(
  phase = c("baseline", "stimulus", "response", "recovery"),
  start = c(0, 500, 1500, 2000),
  end = c(500, 1500, 2000, 2500)
)

phase_windows
```

## Segment samples into task phases

`segment_gazepoint_task_phases()` assigns each row to the first declared phase window that contains its timestamp. Rows outside the declared windows are labelled separately.

```{r}
segmented <- segment_gazepoint_task_phases(
  gaze_data,
  time_col = "time_ms",
  phase_windows = phase_windows,
  keep_window_metadata = TRUE
)

head(segmented)
```

The helper records whether each row was assigned to a declared phase.

```{r}
table(segmented$task_phase, useNA = "ifany")
table(segmented$.gp3_phase_assigned, useNA = "ifany")
```

## Summarise phase coverage

`summarize_gazepoint_phase_coverage()` summarises row counts, timing, and optional complete-value coverage by phase and group.

```{r}
phase_coverage <- summarize_gazepoint_phase_coverage(
  segmented,
  phase_col = "task_phase",
  group_cols = c("subject", "condition"),
  time_col = "time_ms",
  value_cols = "pupil"
)

phase_coverage
```

The British spelling alias is also available.

```{r}
summarise_gazepoint_phase_coverage(
  segmented,
  phase_col = "task_phase",
  time_col = "time_ms",
  value_cols = "pupil"
)
```

## Plot phase timing

`plot_gazepoint_phase_timeline()` visualises the timing coverage of each phase. When timing information is available, it draws phase intervals by group.

```{r}
plot_gazepoint_phase_timeline(
  segmented,
  phase_col = "task_phase",
  group_cols = c("subject", "condition"),
  time_col = "time_ms",
  title = "Task-phase coverage by subject and condition"
)
```

If timing information is not supplied, the same function falls back to a descriptive count plot.

```{r}
plot_gazepoint_phase_timeline(
  segmented,
  phase_col = "task_phase",
  group_cols = "condition",
  title = "Rows by task phase and condition"
)
```

## Generate cautious report text

`report_gazepoint_phase_coverage()` returns the summary table, phase totals, overall diagnostics, and compact report text.

```{r}
phase_report <- report_gazepoint_phase_coverage(
  phase_coverage,
  digits = 1
)

phase_report$overall
```

```{r}
phase_report$phase_totals
```

```{r}
phase_report$report_text
```

The generated wording is deliberately cautious. Phase coverage summaries describe data availability within declared task windows; they do not by themselves justify exclusion, imputation, or modelling decisions.

## Suggested workflow

A transparent task-phase workflow is:

1. Declare phase windows from the experimental protocol.
2. Segment samples using the declared timing table.
3. Summarise row, timing, and value coverage by participant, trial, condition, or other grouping variables.
4. Inspect the phase timeline visually.
5. Report coverage diagnostics separately from exclusion decisions.
