---
title: "Plot gallery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plot gallery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)

library(gp3tools)
```

This article lists the main plotting workflows supported by `gp3tools` and explains what each plot is intended to communicate.

The examples are written as reusable templates. Replace the example object names with outputs from your own workflow.

## Quality-control plots

```{r, eval=FALSE}
plot_gazepoint_pupil_gaps(pupil_qc)
plot_gazepoint_pupil_baseline(baseline_qc)
plot_gazepoint_signal_activity(signal_qc)
plot_gazepoint_time_resets(time_reset_qc)
plot_gazepoint_biometric_report_dashboard(report_dashboard)
```

Use these plots to inspect missingness, timing issues, baseline stability, and signal availability before modelling.

## AOI and fixation plots

```{r, eval=FALSE}
plot_gazepoint_aoi_summary(aoi_summary)
plot_gazepoint_transition_matrix(transition_matrix)
plot_gazepoint_time_varying_transition_matrix(time_varying_transitions)
```

AOI and transition plots should be interpreted as visual-attention summaries, not as direct evidence of internal cognitive states.

## Model plots

```{r, eval=FALSE}
plot_gazepoint_model_predictions(model_predictions)
plot_gazepoint_model_leave_one_out(loo_results)
```

Model plots are useful for communicating predicted values, uncertainty, and influential cases.

## External face-data plots

```{r, eval=FALSE}
plot_gazepoint_face_quality(face_quality_qc)
```

External facial-behaviour plots should be described as quality-control and feature-summary plots. They should not be described as direct emotion detection.
