---
title: "Plot gallery"
description: "Representative gp3tools visual outputs built from safe synthetic demonstration data."
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = '#>',
  fig.width = 7,
  fig.height = 4.5,
  warning = FALSE,
  message = FALSE
)
```


This article is a website-only showcase of the main plotting outputs in `gp3tools`.
It is intended for users, reviewers, and readers who want to see the visual
diagnostics and reporting outputs produced by the package.

The gallery uses only package-safe synthetic or example data.
It should not be read as an empirical analysis of real participants.

## Plot functions covered

```{r plot-function-inventory}
plot_functions <- sort(grep(
  '^plot_',
  getNamespaceExports('gp3tools'),
  value = TRUE
))

data.frame(
  plot_function = plot_functions,
  row.names = NULL
)
```

## Planned gallery sections

- Sampling and tracking-quality plots
- Pupil preprocessing and pupil time-course plots
- AOI geometry, verification, and transition plots
- GCA and model-prediction plots
- Cluster and multiverse-result plots

We will add the rendered plots section by section, using only reproducible demonstration data.
