---
title: "Plot gallery"
description: "Representative gp3tools visual outputs built from safe synthetic demonstration data."
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = '#>',
  fig.width = 7,
  fig.height = 4.5,
  warning = FALSE,
  message = FALSE,
  echo = FALSE
)
```

This article is a website-only showcase of representative visual outputs in `gp3tools`.
It is intended for users, reviewers, and readers who want to see the visual
diagnostics and reporting plots produced by the package.

The gallery uses only package-safe synthetic or example data.
It should not be read as an empirical analysis of real participants.

## Important note

`gp3tools` can parse official Gazepoint Analysis summary exports, but metrics
recomputed from all-gaze and fixation files may not always exactly reproduce
Gazepoint's internal summary calculations. For official Gazepoint summary
values, use `read_gazepoint_summary()`. For transparent reproducible
calculations from exported rows, use the sample-level and fixation-level
summary functions.

```{r gallery-data}
library(gp3tools)

master <- gp3tools::gazepoint_example_master
fixations <- gp3tools::gazepoint_example_fixations
aoi_geometry <- gp3tools::gazepoint_example_aoi_geometry
aoi_windows <- gp3tools::gazepoint_example_aoi_windows

pupil_demo <- master
pupil_demo$pupil_clean <- ifelse(pupil_demo$artifact, NA_real_, pupil_demo$pupil)
pupil_demo$pupil_interpolated <- ifelse(
  is.na(pupil_demo$pupil_clean),
  ave(
    pupil_demo$pupil,
    pupil_demo$subject,
    pupil_demo$trial_global,
    FUN = function(z) mean(z, na.rm = TRUE)
  ),
  pupil_demo$pupil_clean
)
pupil_demo$pupil_was_interpolated <- is.na(pupil_demo$pupil_clean) &
  !is.na(pupil_demo$pupil_interpolated)
pupil_demo$pupil_interpolation_status <- ifelse(
  pupil_demo$pupil_was_interpolated,
  'interpolated',
  'observed'
)
pupil_demo$pupil_baseline_corrected <- pupil_demo$pupil_interpolated - ave(
  pupil_demo$pupil_interpolated,
  pupil_demo$subject,
  pupil_demo$trial_global,
  FUN = function(z) mean(z[seq_len(min(5, length(z)))], na.rm = TRUE)
)
pupil_demo$pupil_smoothed <- ave(
  pupil_demo$pupil_baseline_corrected,
  pupil_demo$subject,
  pupil_demo$trial_global,
  FUN = function(z) {
    zz <- as.numeric(stats::filter(z, rep(1 / 3, 3), sides = 2))
    zz[is.na(zz)] <- z[is.na(zz)]
    zz
  }
)

gca_demo <- expand.grid(
  subject = paste0('S', 1:8),
  condition = c('control', 'treatment'),
  gca_time = seq(0, 900, by = 100),
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)
gca_demo <- gca_demo[order(gca_demo$subject, gca_demo$condition, gca_demo$gca_time), ]
time_poly <- stats::poly(gca_demo$gca_time, degree = 3)
subject_shift <- stats::setNames(seq(-0.20, 0.20, length.out = 8), paste0('S', 1:8))
subject_slope <- stats::setNames(seq(-0.03, 0.03, length.out = 8), paste0('S', 1:8))
gca_demo$time_poly_1 <- as.numeric(time_poly[, 1])
gca_demo$time_poly_2 <- as.numeric(time_poly[, 2])
gca_demo$time_poly_3 <- as.numeric(time_poly[, 3])
gca_demo$gca_pupil <- 0.20 +
  ifelse(gca_demo$condition == 'treatment', 0.10, 0) +
  0.60 * gca_demo$time_poly_1 -
  0.35 * gca_demo$time_poly_2 +
  0.15 * gca_demo$time_poly_3 +
  ifelse(gca_demo$condition == 'treatment', 0.15 * gca_demo$time_poly_1, 0) +
  ifelse(gca_demo$condition == 'treatment', subject_slope[gca_demo$subject] * gca_demo$gca_time / 900, 0) +
  subject_shift[gca_demo$subject]
gca_demo$gca_weight <- 1
```

## Exported plot functions

```{r plot-function-inventory}
plot_functions <- sort(grep(
  '^plot_',
  getNamespaceExports('gp3tools'),
  value = TRUE
))

data.frame(
  plot_function = plot_functions,
  row.names = NULL
)
```

## Sampling rate

```{r sampling-rate-plot}
sampling <- check_sampling_rate(
  master,
  group_cols = c('USER_FILE', 'MEDIA_ID'),
  time_col = 'TIME'
)

plot_sampling_rate(
  sampling,
  user_col = 'USER_FILE',
  media_col = 'MEDIA_ID',
  expected_hz = 20,
  hz_tolerance = 2
)
```

## Tracking quality

```{r tracking-quality-plot}
quality <- summarise_tracking_quality(
  master,
  group_cols = c('USER_FILE', 'MEDIA_ID')
)
quality$review_required <- quality$BPOGV_valid_pct < 70

plot_tracking_quality(
  quality,
  metric_cols = 'BPOGV_valid_pct',
  user_col = 'USER_FILE',
  media_col = 'MEDIA_ID',
  review_col = 'review_required',
  min_valid_pct = 70
)
```

## Pupil status

```{r pupil-status-plot}
plot_gazepoint_pupil_status(
  pupil_demo,
  time_col = 'time',
  pupil_col = 'pupil_interpolated',
  status_col = 'pupil_interpolation_status',
  interpolated_col = 'pupil_was_interpolated',
  group_cols = c('subject', 'trial_global'),
  plot_type = 'summary'
)
```

## Pupil preprocessing

```{r pupil-preprocessing-plot}
plot_gazepoint_pupil_preprocessing(
  pupil_demo,
  subject = 'S01',
  trial_global = 'control_T1',
  time_col = 'time',
  raw_pupil_col = 'pupil',
  clean_pupil_col = 'pupil_clean',
  interpolated_pupil_col = 'pupil_interpolated',
  baseline_pupil_col = 'pupil_baseline_corrected',
  smoothed_pupil_col = 'pupil_smoothed',
  status_col = 'pupil_interpolation_status',
  plot_style = 'faceted'
)
```

## Pupil time course

```{r pupil-timecourse-plot}
plot_gazepoint_pupil_timecourse(
  master,
  pupil_col = 'pupil',
  time_col = 'time',
  condition_col = 'condition',
  bin_width_ms = 100
)
```

## AOI verification

```{r aoi-verification-plot}
plot_gazepoint_aoi_verification(
  aoi_geometry = aoi_geometry,
  gaze_data = master,
  geometry_aoi_col = 'aoi',
  geometry_stimulus_col = 'media_id',
  gaze_x_col = 'x',
  gaze_y_col = 'y',
  gaze_stimulus_col = 'MEDIA_ID',
  screen_x_range = c(0, 1),
  screen_y_range = c(0, 1),
  facet_by_stimulus = TRUE,
  show_labels = TRUE,
  show_gaze = TRUE
)
```

## AOI transition matrix

```{r aoi-transition-matrix-plot}
aoi_transitions <- compute_gazepoint_aoi_transition_matrix(
  master,
  aoi_col = 'AOI',
  time_col = 'time',
  group_cols = c('subject', 'MEDIA_ID', 'trial_global'),
  by_cols = 'condition',
  include_non_aoi = TRUE,
  include_self_transitions = FALSE
)

plot_gazepoint_aoi_transition_matrix(
  aoi_transitions,
  value = 'prob',
  by_cols = 'condition',
  facet = TRUE,
  title = 'AOI transition probabilities by condition'
)
```

## Transition heatmap

```{r transition-heatmap-plot}
legacy_transitions <- compute_transition_matrix(
  fixations,
  group_cols = 'MEDIA_ID',
  aoi_col = 'AOI',
  time_col = 'FPOGS'
)

plot_transition_heatmap(legacy_transitions)
```

## GCA model plot

```{r gca-model-plot}
gca_fit <- fit_gazepoint_gca(
  gca_demo,
  random_slopes = FALSE,
  REML = FALSE
)

plot_gazepoint_gca(
  gca_fit,
  title = 'Synthetic GCA pupil trajectories'
)
```

## Model-prediction plot

```{r model-prediction-plot}
prediction_model <- stats::lm(
  gca_pupil ~ gca_time * condition,
  data = gca_demo
)

plot_gazepoint_model_predictions(
  gca_demo,
  model = prediction_model,
  x_col = 'gca_time',
  outcome_col = 'gca_pupil',
  condition_col = 'condition'
)
```

## AOI GAMM plot

```{r aoi-gamm-plot}
aoi_gamm_demo <- data.frame(
  .gp3_aoi_gamm_subject = aoi_windows$subject,
  .gp3_aoi_gamm_condition = aoi_windows$condition,
  .gp3_aoi_gamm_time_bin = aoi_windows$window_start_ms,
  .gp3_aoi_gamm_success = aoi_windows$n_target_samples,
  .gp3_aoi_gamm_failure = pmax(
    aoi_windows$n_valid_denominator_samples - aoi_windows$n_target_samples,
    0
  ),
  .gp3_aoi_gamm_denominator = aoi_windows$n_valid_denominator_samples,
  .gp3_aoi_gamm_status = 'ok'
)

aoi_gamm_demo <- aoi_gamm_demo[
  is.finite(aoi_gamm_demo$.gp3_aoi_gamm_time_bin) &
    aoi_gamm_demo$.gp3_aoi_gamm_denominator > 0,
]

aoi_gamm_fit <- fit_gazepoint_aoi_gamm(
  aoi_gamm_demo,
  include_condition = TRUE,
  condition_smooths = TRUE,
  random_subject = TRUE,
  time_k = 4,
  method = 'REML'
)

plot_gazepoint_aoi_gamm(
  aoi_gamm_fit,
  title = 'Synthetic AOI GAMM trajectory'
)
```

## Cluster-results plot

```{r cluster-results-plot}
cluster_timecourse <- data.frame(
  .gp3_cluster_time_bin = seq(0, 900, by = 100),
  mean_difference = c(0.01, 0.02, 0.04, 0.08, 0.11, 0.10, 0.07, 0.04, 0.02, 0.01),
  statistic = c(0.3, 0.8, 1.5, 2.4, 3.1, 2.7, 2.1, 1.3, 0.7, 0.2),
  cluster_id = c(NA, NA, NA, 1, 1, 1, 1, NA, NA, NA),
  point_candidate = c(FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE, FALSE, FALSE, FALSE)
)

cluster_result <- list(
  timecourse = cluster_timecourse,
  clusters = data.frame(
    cluster_id = 1,
    cluster_direction = 'positive',
    start_time_bin = 300,
    end_time_bin = 600,
    p_value = 0.04
  ),
  settings = list(
    cluster_threshold = 2,
    tail = 'two_sided',
    difference = 'treatment - control'
  ),
  model_status = 'ok'
)

plot_gazepoint_cluster_results(
  cluster_result,
  plot_type = 'both',
  significant_only = FALSE,
  title = 'Synthetic cluster-result display'
)
```

## Multiverse-results plot

```{r multiverse-results-plot}
multiverse_summary <- list(
  branch_summary = data.frame(
    multiverse_family = rep('pupil', 4),
    branch_id = paste0('mv_pupil_', 1:4),
    branch_label = c(
      'gap75_smooth3_baseline0_200',
      'gap150_smooth3_baseline0_200',
      'gap75_smooth5_baseline0_200',
      'gap150_smooth5_baseline0_200'
    ),
    branch_status = c('completed', 'completed', 'completed', 'completed'),
    stringsAsFactors = FALSE
  )
)
class(multiverse_summary) <- c('gp3_multiverse_summary_results', 'list')

plot_gazepoint_multiverse_results(
  multiverse_summary,
  plot = 'status',
  family = 'pupil',
  title = 'Synthetic preprocessing multiverse status'
)
```
