---
title: "Plot gallery"
description: "Representative gp3tools visual outputs built from safe synthetic demonstration data."
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = '#>',
  fig.width = 7,
  fig.height = 4.5,
  warning = FALSE,
  message = FALSE
)
```

This article is a website-only showcase of representative visual outputs in `gp3tools`.
It is intended for users, reviewers, and readers who want to see the visual
diagnostics and reporting plots produced by the package.

The gallery uses only package-safe synthetic or example data.
It should not be read as an empirical analysis of real participants.

```{r gallery-data}
library(gp3tools)

master <- gp3tools::gazepoint_example_master
aoi_geometry <- gp3tools::gazepoint_example_aoi_geometry
```

## Plot functions covered

```{r plot-function-inventory}
plot_functions <- sort(grep(
  '^plot_',
  getNamespaceExports('gp3tools'),
  value = TRUE
))

data.frame(
  plot_function = plot_functions,
  row.names = NULL
)
```

## Sampling rate

The built-in demonstration data are sampled every 50 ms, corresponding to
approximately 20 Hz. This differs from a typical 60 Hz GP3 recording, but is
sufficient for a compact website example.

```{r sampling-rate-plot}
sampling <- check_sampling_rate(
  master,
  group_cols = c('USER_FILE', 'MEDIA_ID'),
  time_col = 'TIME'
)

plot_sampling_rate(
  sampling,
  user_col = 'USER_FILE',
  media_col = 'MEDIA_ID',
  expected_hz = 20,
  hz_tolerance = 2
)
```

## Tracking quality

The tracking-quality plot summarises valid gaze percentage by participant file
and stimulus. A simple review flag is created here for demonstration.

```{r tracking-quality-plot}
quality <- summarise_tracking_quality(
  master,
  group_cols = c('USER_FILE', 'MEDIA_ID')
)

quality$review_required <- quality$BPOGV_valid_pct < 70

plot_tracking_quality(
  quality,
  metric_cols = 'BPOGV_valid_pct',
  user_col = 'USER_FILE',
  media_col = 'MEDIA_ID',
  review_col = 'review_required',
  min_valid_pct = 70
)
```

## Pupil time course

The pupil time-course plot shows condition-level pupil trajectories after
binning the synthetic sample-level data.

```{r pupil-timecourse-plot}
plot_gazepoint_pupil_timecourse(
  master,
  pupil_col = 'pupil',
  time_col = 'time',
  condition_col = 'condition',
  bin_width_ms = 100
)
```

## AOI verification

The AOI verification plot overlays gaze samples on the synthetic AOI geometry.
This helps users inspect whether AOI coordinates and gaze coordinates are on
the same scale before computing AOI summaries.

```{r aoi-verification-plot}
plot_gazepoint_aoi_verification(
  aoi_geometry = aoi_geometry,
  gaze_data = master,
  geometry_aoi_col = 'aoi',
  geometry_stimulus_col = 'media_id',
  gaze_x_col = 'x',
  gaze_y_col = 'y',
  gaze_stimulus_col = 'MEDIA_ID',
  screen_x_range = c(0, 1),
  screen_y_range = c(0, 1),
  facet_by_stimulus = TRUE,
  show_labels = TRUE,
  show_gaze = TRUE
)
```

## Planned additions

Additional sections will be added for AOI transitions, pupil preprocessing,
GCA/model-prediction plots, cluster results, and multiverse summaries.
