---
title: "Pupil preprocessing and baseline audit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pupil preprocessing and baseline audit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)

library(gp3tools)
```

This article focuses on pupil preprocessing, baseline-readiness checks, and the additional sampling/estimator checks needed when pupil-response latency is reported.

## Recommended order

1. inspect raw pupil coverage;
2. flag missingness and implausible values;
3. interpolate short gaps only when justified;
4. audit baseline coverage;
5. create baseline-corrected or change-score variables;
6. if latency is a target outcome, audit estimator sensitivity and sampling resolution;
7. document preprocessing and timing choices.

## Example preprocessing workflow

```{r, eval=FALSE}
gap_qc <- audit_gazepoint_pupil_gaps(all_gaze)

pupil_flagged <- flag_gazepoint_pupil_hampel(
  all_gaze,
  pupil_col = "pupil_diameter"
)

pupil_interp <- interpolate_gazepoint_pupil_pchip(
  pupil_flagged,
  pupil_col = "pupil_diameter"
)

baseline_qc <- audit_gazepoint_pupil_baseline(
  pupil_interp,
  baseline_start = -0.500,
  baseline_end = 0
)

reliability_qc <- audit_gazepoint_pupil_reliability(pupil_interp)
```

## When latency is a target outcome

A latency estimate is partly a property of the sampling grid and the estimator, not only of the underlying physiological response. `audit_gp3_pupil_latency_resolution()` therefore compares two onset estimators and reports their spread alongside the observed sampling rate, the implied temporal resolution, robust baseline noise, signal-to-noise ratio, and a compact latency-resolvability label.

Use the diagnostic on a scientifically justified analysis trace after time alignment and preprocessing. Do not pool raw observations across independent participants or trials and treat the resulting latency as inferential evidence. The function is intended for QC and transparent reporting.

## Deterministic synthetic example

The example below creates a 60 Hz trace with a known constriction beginning after the event. The audit is then run on the resulting pupil series.

```{r latency-example}
set.seed(42)

time <- seq(-0.5, 2, by = 1 / 60)
true_onset <- 0.45
response <- ifelse(
  time > true_onset,
  -0.35 * (1 - exp(-(time - true_onset) / 0.12)),
  0
)
pupil <- 4.2 + response + rnorm(length(time), sd = 0.006)

latency_qc <- audit_gp3_pupil_latency_resolution(
  time = time,
  pupil = pupil,
  event_time = 0,
  baseline_window = c(-0.5, 0),
  search_window = c(0, 2),
  direction = "constriction",
  threshold_sigma = 3,
  sustain_ms = 50
)

latency_qc$estimates_s
round(latency_qc$estimator_spread_ms, 1)
round(latency_qc$sampling_hz, 1)
round(latency_qc$sampling_resolution_ms, 1)
latency_qc$latency_resolvability
```

The two latency estimates should be read together. A small estimator spread is reassuring, but it does not remove the sampling-resolution limit or make the estimate hardware- and algorithm-independent.

```{r latency-plot, fig.width=7, fig.height=4.5, fig.alt="Synthetic pupil trace with event time and two latency estimates"}
plot(
  time,
  pupil,
  type = "l",
  xlab = "Time from event (s)",
  ylab = "Pupil diameter (a.u.)",
  main = "Synthetic pupil trace with latency estimates"
)
abline(v = 0, lty = 2)
latency_lines <- latency_qc$estimates_s[is.finite(latency_qc$estimates_s)]
if (length(latency_lines)) {
  abline(v = latency_lines, lty = 3)
}
legend(
  "bottomleft",
  legend = c("Pupil trace", "Event time", "Latency estimates"),
  lty = c(1, 2, 3),
  bty = "n"
)
```

## Interpreting the audit

- `estimates_s` contains the sustained-threshold and maximum-slope tangent estimates.
- `estimator_spread_ms` quantifies disagreement between finite estimators.
- `sampling_hz` and `sampling_resolution_ms` make the temporal sampling limit explicit.
- `noise_mad_sigma` and `signal_to_noise` describe baseline noise and response strength used by the diagnostic.
- `latency_resolvability` is a package QC heuristic (`high`, `moderate`, or `low`), not a physiological certainty grade.

When one estimator is unavailable, report that fact rather than silently substituting the remaining estimate as if both methods agreed.

## Sensitivity checks

Threshold and sustain settings should be chosen from methodological considerations rather than tuned post hoc to obtain a preferred onset. When timing is substantively important, rerun the audit under defensible alternatives and report whether the scientific conclusion changes.

```{r, eval=FALSE}
latency_qc_lower_threshold <- audit_gp3_pupil_latency_resolution(
  time,
  pupil,
  threshold_sigma = 2.5,
  sustain_ms = 50
)

latency_qc_longer_sustain <- audit_gp3_pupil_latency_resolution(
  time,
  pupil,
  threshold_sigma = 3,
  sustain_ms = 100
)
```

## Reporting note

Baseline correction should be reported with the baseline window, minimum coverage rule, interpolation rule, excluded cases, and sensitivity checks. If latency is reported, also state the sampling rate, estimator definition, threshold/sustain settings, estimator spread, and any cases in which latency was not resolvable. Treat latency as method-dependent timing evidence rather than an instrument-independent ground truth.
