---
title: "Quality-control dashboard workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quality-control dashboard workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)

library(gp3tools)
```

This article shows how to organise reviewer-facing quality-control checks.

The focus is on transparent reporting rather than automatic exclusion.

## Core QC domains

- file structure and import success;
- missing gaze or pupil rows;
- long gaps and interpolation burden;
- baseline coverage and stability;
- calibration or recalibration diagnostics;
- timing resets and signal activity;
- exclusion recommendations and review flags.

## Example workflow

```{r, eval=FALSE}
gap_qc <- audit_gazepoint_pupil_gaps(all_gaze)
baseline_qc <- audit_gazepoint_pupil_baseline(all_gaze)
reliability_qc <- audit_gazepoint_pupil_reliability(all_gaze)
luminance_qc <- audit_gazepoint_stimulus_luminance(stimulus_data)

signal_qc <- assess_gazepoint_signal_activity(all_gaze)
reset_qc <- assess_gazepoint_time_resets(all_gaze)

exclusions <- recommend_gazepoint_exclusions(
  pupil_qc = gap_qc,
  baseline_qc = baseline_qc
)

dashboard <- plot_gazepoint_biometric_report_dashboard(
  gap_qc = gap_qc,
  baseline_qc = baseline_qc,
  signal_qc = signal_qc
)
```

## Reporting recommendation

Report QC results as descriptive diagnostics and decision support. Avoid presenting automated flags as definitive evidence that a participant or trial is invalid without a documented rule.
