---
title: "QC reporting bundle"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{QC reporting bundle}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  warning = FALSE,
  message = FALSE
)

library(gp3tools)
```

This article demonstrates lightweight helpers for collecting existing gp3tools quality-control outputs into a compact overview. The QC bundle is a reporting aid. It does not rerun audits, define exclusion rules, or replace readiness gates, reporting checklists, or study-specific decisions.

## Example QC objects

The collector works with gp3tools-style objects that contain an `overview` table, as well as standalone overview data frames.

```{r}
tracking_audit <- list(
  overview = data.frame(
    audit_status = "ok",
    message = "Tracking-quality checks passed.",
    stringsAsFactors = FALSE
  )
)

missingness_audit <- list(
  overview = data.frame(
    audit_status = "review",
    message = "Missingness should be reported by condition.",
    stringsAsFactors = FALSE
  )
)

readiness_gate <- list(
  overview = data.frame(
    readiness_status = "ready",
    decision_message = "Ready for analysis review.",
    stringsAsFactors = FALSE
  )
)

qc_objects <- list(
  tracking_audit = tracking_audit,
  missingness_audit = missingness_audit,
  readiness_gate = readiness_gate
)
```

## Collect QC summaries

`collect_gazepoint_qc_summaries()` extracts interpretable status and message information from supplied objects.

```{r}
qc_bundle <- collect_gazepoint_qc_summaries(qc_objects)

qc_bundle$overview
```

```{r}
qc_bundle$object_summary
```

```{r}
qc_bundle$status_counts
```

When requested, the function also keeps a combined long-form table of available overview rows.

```{r}
head(qc_bundle$overview_rows)
```

## Summarise QC status

`summarize_gazepoint_qc_status()` produces a compact status summary from a bundle, an object-summary table, or a raw list of objects.

```{r}
qc_status <- summarize_gazepoint_qc_status(qc_bundle)

qc_status$overview
```

The British spelling alias is also available.

```{r}
summarise_gazepoint_qc_status(qc_bundle)$status_counts
```

## Plot QC overview

`plot_gazepoint_qc_overview()` can show status counts.

```{r}
plot_gazepoint_qc_overview(
  qc_bundle,
  plot_type = "status_counts",
  title = "QC status counts"
)
```

It can also show object-level status.

```{r}
plot_gazepoint_qc_overview(
  qc_bundle,
  plot_type = "objects",
  title = "QC status by object"
)
```

## Generate cautious report text

`report_gazepoint_qc_overview()` returns compact text for reporting support.

```{r}
qc_report <- report_gazepoint_qc_overview(qc_bundle)

qc_report$report_text
```

The generated wording is deliberately cautious. The QC overview describes available outputs and interpreted status patterns; it should be read together with the underlying audit tables, readiness gates, exclusion recommendations, and reporting checklist.

## Suggested workflow

A transparent QC-reporting workflow is:

1. Run the relevant gp3tools audits, checks, summaries, and reporting helpers.
2. Collect their overview tables into a QC bundle.
3. Inspect the status summary and object-level messages.
4. Use the plot and report text as a compact review aid.
5. Keep final exclusion, imputation, modelling, and manuscript decisions separate from the overview itself.
