---
title: "Scanpath and quality-control quick wins"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scanpath and quality-control quick wins}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  warning = FALSE,
  message = FALSE
)

library(gp3tools)
```

This article demonstrates a compact set of descriptive helpers for privacy-safe examples, binocular pupil preparation, trackloss review, time-series plotting, and multi-scanpath visual inspection. These helpers are intended for transparent quality control and documentation. They do not replace study-specific preprocessing decisions or inferential modelling.

## Simulate pupil and gaze data

`simulate_gazepoint_pupil_data()` creates a small synthetic data set with participant, trial, condition, time-bin, gaze-coordinate, left-pupil, right-pupil, blink, trackloss, and combined-pupil columns.

```{r}
synthetic <- simulate_gazepoint_pupil_data(
  n_subjects = 4,
  n_trials = 4,
  n_time_bins = 20,
  conditions = c("control", "treatment"),
  blink_probability = 0.05,
  seed = 123
)

head(synthetic)
```

The generator is designed for examples and tests. It should not be used to make claims about empirical pupil physiology.

## Combine left and right pupil channels

`combine_gazepoint_eyes()` combines two eye-specific numeric columns into a single analysis column. The default method averages available left/right values. Other options can prefer one eye or choose the globally less-missing eye.

```{r}
combined <- combine_gazepoint_eyes(
  synthetic,
  left_col = "pupil_left",
  right_col = "pupil_right",
  output_col = "pupil_combined",
  method = "mean",
  valid_min = 1,
  valid_max = 9
)

head(combined[, c("pupil_left", "pupil_right", "pupil_combined")])
```

## Flag groups by trackloss

`clean_gazepoint_by_trackloss()` computes trackloss rates globally or within grouping columns. Here, participant-trial groups with more than 20% trackloss are flagged.

```{r}
trackloss_flagged <- clean_gazepoint_by_trackloss(
  synthetic,
  group_cols = c("subject", "trial"),
  tracking_col = "trackloss",
  max_trackloss = 0.20,
  action = "flag"
)

head(trackloss_flagged[, c("subject", "trial", "trackloss", ".gp3_trackloss_rate", ".gp3_trackloss_exclude")])
```

A compact group-level summary is stored as an attribute.

```{r}
head(attr(trackloss_flagged, "gp3_trackloss_summary"))
```

The same helper can filter high-trackloss groups, but filtering should normally be reported as an explicit preprocessing decision.

## Plot a descriptive time series

`plot_gazepoint_time_series()` provides a general descriptive line plot for pupil, gaze, AOI, or other time-varying measures that have already been prepared by the user.

```{r}
plot_gazepoint_time_series(
  synthetic,
  time_col = "time_bin",
  value_col = "pupil",
  group_cols = c("subject", "trial"),
  colour_col = "condition",
  title = "Synthetic pupil time series",
  x_label = "Time bin",
  y_label = "Pupil size"
)
```

This plot is descriptive. It does not smooth, model, or test condition differences.

## Plot multiple scanpaths

`plot_gazepoint_scanpaths()` supports quick visual inspection of gaze paths across participants, trials, or conditions.

```{r}
plot_gazepoint_scanpaths(
  synthetic,
  x_col = "gaze_x",
  y_col = "gaze_y",
  order_col = "time_bin",
  group_cols = c("subject", "trial"),
  colour_col = "condition",
  screen_width = 1920,
  screen_height = 1080,
  title = "Synthetic multi-scanpath inspection"
)
```

For crowded data, faceting can make trial or condition-level review easier.

```{r}
plot_gazepoint_scanpaths(
  synthetic,
  x_col = "gaze_x",
  y_col = "gaze_y",
  order_col = "time_bin",
  group_cols = c("subject", "trial"),
  facet_col = "condition",
  show_points = FALSE,
  title = "Faceted synthetic scanpaths"
)
```


## Audit screen bounds

`audit_gazepoint_screen_bounds()` checks whether gaze coordinates are missing, equal to `(0, 0)`, or outside expected screen or stimulus bounds. This is useful before heatmaps, AOI checks, or scanpath visualisation.

```{r}
screen_audit <- audit_gazepoint_screen_bounds(
  synthetic,
  x_col = "gaze_x",
  y_col = "gaze_y",
  screen_width = 1920,
  screen_height = 1080,
  group_cols = c("subject", "trial")
)

screen_audit$overall_summary
```

The row-level and group-level outputs make the diagnostic transparent without automatically changing the data.

```{r}
head(screen_audit$group_summary)
```

## Harmonize screen coordinates

`harmonize_gazepoint_screen_coordinates()` rescales gaze coordinates from one screen or stimulus resolution to another. This is a deterministic transformation for harmonising exports before plotting or descriptive summaries. It is not a recalibration method.

```{r}
harmonized <- harmonize_gazepoint_screen_coordinates(
  synthetic,
  x_col = "gaze_x",
  y_col = "gaze_y",
  from_width = 1920,
  from_height = 1080,
  to_width = 1280,
  to_height = 720
)

head(harmonized[, c("gaze_x", "gaze_y", "gaze_x_harmonized", "gaze_y_harmonized")])
```
These visualisations are intended for quality review and documentation, not as inferential scanpath-comparison methods.
