---
title: "Scanpath-cluster stability and sensitivity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scanpath-cluster stability and sensitivity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Scope

Cluster assignments can change when scanpaths, linkage methods, or distance
definitions change. This article demonstrates a lightweight subsampling
workflow for evaluating that sensitivity.

The workflow records:

- full-data reference solutions;
- iteration-level adjusted Rand agreement;
- pairwise co-clustering probabilities;
- pair-coverage rates;
- sequence-level within-cluster stability;
- representative-scanpath selection rates;
- sensitivity across hierarchical linkage methods.

These diagnostics quantify reproducibility under the specified resampling
scheme. They do not prove that a cluster solution represents a true latent
cognitive strategy.

```{r load-package}
library(gp3tools)
```

## Synthetic distance structure

The example contains nine scanpaths arranged around three synthetic sequence
patterns. Within-pattern distances are small and between-pattern distances are
larger, with modest deterministic variation.

```{r synthetic-distance}
sequence_ids <- paste0("scanpath_", sprintf("%02d", 1:9))
latent_cluster <- rep(1:3, each = 3)

distance_matrix <- outer(
  seq_along(sequence_ids),
  seq_along(sequence_ids),
  FUN = function(i, j) {
    ifelse(
      latent_cluster[i] == latent_cluster[j],
      0.08 + 0.01 * abs(i - j),
      0.85 + 0.02 * abs(i - j)
    )
  }
)

diag(distance_matrix) <- 0

dimnames(distance_matrix) <- list(
  sequence_ids,
  sequence_ids
)

round(distance_matrix, 2)
```

## Reference clustering

```{r reference-fit}
reference_fit <- cluster_gazepoint_scanpaths(
  distance_matrix,
  k = 3,
  method = "hierarchical",
  linkage = "average"
)

reference_fit$assignments
```

```{r reference-mds}
plot_gazepoint_scanpath_clusters(
  reference_fit,
  plot = "mds",
  main = "Synthetic scanpath-distance structure"
)
```

## Subsampling stability

Each iteration retains 80% of scanpaths without replacement, refits the
three-cluster solution, and compares it with the corresponding full-data
reference partition.

```{r bootstrap-stability}
stability <- bootstrap_gazepoint_scanpath_clusters(
  distance_matrix,
  k = 3,
  n_boot = 100,
  sample_fraction = 0.8,
  method = "hierarchical",
  linkages = c("average", "complete", "ward.D2"),
  seed = 20260715
)

utils::head(stability$iteration_summary)
```

The seed is local to the helper: the caller's random-number state is restored
after the routine finishes.

## Adjusted Rand agreement

```{r ari-distribution}
plot_gazepoint_scanpath_cluster_stability(
  stability,
  plot = "ari",
  specification = "hierarchical_average"
)
```

The adjusted Rand index compares a subsampled partition with the full-data
reference partition after restricting both to the sampled scanpaths. Values
near one indicate strong agreement; lower values identify sensitivity to the
particular scanpaths retained.

## Co-clustering probabilities

```{r coclustering-plot}
plot_gazepoint_scanpath_cluster_stability(
  stability,
  plot = "coclustering",
  specification = "hierarchical_average"
)
```

Each cell reports the proportion of iterations in which two observed
scanpaths were assigned to the same cluster, conditional on both being
sampled. Pair coverage is retained separately because some pairs may co-occur
less often than others.

```{r pair-coverage}
round(
  stability$pair_coverage$hierarchical_average,
  2
)
```

## Compact stability summaries

```{r stability-summary}
stability_summary <-
  summarise_gazepoint_scanpath_cluster_stability(
    stability,
    min_pair_coverage = 0.50,
    stable_threshold = 0.75
  )

stability_summary$overview
```

```{r sequence-summary}
stability_summary$sequence_summary
```

Within-cluster stability is the mean co-clustering probability between a
scanpath and the other members of its full-data reference cluster. Between-
cluster co-clustering provides a complementary separation diagnostic.

```{r sequence-plot}
plot_gazepoint_scanpath_cluster_stability(
  stability,
  plot = "sequence",
  specification = "hierarchical_average",
  min_pair_coverage = 0.50,
  stable_threshold = 0.75
)
```

The dashed line marks the reporting threshold supplied to the plot. It is a
review threshold, not a universal inferential cutoff.

## Representative-scanpath stability

```{r representative-stability}
representative_table <-
  stability_summary$representative_stability

representative_table[
  order(
    representative_table$specification,
    representative_table$reference_cluster,
    -representative_table$
      representative_rate_when_included
  ),
]
```

`representative_rate_when_included` reports how often an observed scanpath was
selected as the central representative for its mapped reference cluster,
conditional on that scanpath being included in the iteration.

## Linkage sensitivity

```{r linkage-comparison}
stability_summary$overview[
  ,
  c(
    "specification",
    "mean_adjusted_rand_index",
    "mean_within_cluster_coclustering",
    "mean_between_cluster_coclustering",
    "min_sequence_stability",
    "stability_status"
  )
]
```

Agreement across linkage methods supports robustness to that analytical
choice. Divergence indicates that the cluster geometry or distance structure
requires closer inspection.

## PAM sensitivity

PAM stability can be evaluated with the same interface when the optional
`cluster` package is available.

```{r pam-stability}
if (requireNamespace("cluster", quietly = TRUE)) {
  pam_stability <-
    bootstrap_gazepoint_scanpath_clusters(
      distance_matrix,
      k = 3,
      n_boot = 50,
      sample_fraction = 0.8,
      method = "pam",
      seed = 20260715
    )

  summarise_gazepoint_scanpath_cluster_stability(
    pam_stability
  )$overview
}
```

## Recommended reporting

Report:

- distance definition and sequence preprocessing;
- clustering method and linkage;
- full-data cluster count and sizes;
- number of resamples;
- retained sample fraction and resolved sample size;
- random seed;
- mean, variability, and minimum adjusted Rand agreement;
- within- and between-cluster co-clustering;
- pair-coverage threshold;
- sequence-level stability;
- representative-scanpath stability;
- sensitivity across linkage or clustering methods.

The resampling design assesses procedural stability under the observed
distance structure. It does not replace external validation or substantive
review of the representative scanpaths.
