---
title: "Integrated signal-preprocessing workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integrated signal-preprocessing workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5
)
```

## Scope

`preprocess_gazepoint_signals()` coordinates the lightweight signal-processing
helpers in a fixed, reviewable order. The workflow preserves original columns,
adds processed columns, records each operation, and returns blink and fixation
event tables separately.

The function is intended for transparent preprocessing and methodological
sensitivity analysis. It does not imply that one detector or preprocessing
configuration recovers a uniquely true sequence of eye events.

```{r load-package}
library(gp3tools)
```

## Synthetic sample-level data

The example is synthetic and contains no participant records.

```{r synthetic-data}
set.seed(1)

n <- 180L
time <- seq(0, by = 1 / 60, length.out = n)

signal <- data.frame(
  USER_ID = rep("P01", n),
  trial = rep(c("T01", "T02"), each = n / 2),
  TIME = time,
  FPOGX = c(
    rep(0.25, 55),
    seq(0.25, 0.75, length.out = 10),
    rep(0.75, 55),
    seq(0.75, 0.35, length.out = 10),
    rep(0.35, 50)
  ),
  FPOGY = 0.50 + stats::rnorm(n, 0, 0.002),
  LPupil = 3.2 + stats::rnorm(n, 0, 0.03),
  RPupil = 3.1 + stats::rnorm(n, 0, 0.03),
  stringsAsFactors = FALSE
)

signal$LPupil[35:40] <- NA_real_
signal$RPupil[35:40] <- NA_real_
signal$LPupil[130:134] <- NA_real_
signal$RPupil[130:134] <- NA_real_

utils::head(signal)
```

## Run the workflow

```{r run-workflow}
result <- preprocess_gazepoint_signals(
  data = signal,
  id_col = "USER_ID",
  group_cols = "trial",
  time_col = "TIME",
  x_col = "FPOGX",
  y_col = "FPOGY",
  left_pupil_col = "LPupil",
  right_pupil_col = "RPupil",
  pupil_mode = "mean",
  detect_blinks = TRUE,
  interpolate_blinks = TRUE,
  smooth_pupil = TRUE,
  smooth_coordinates = TRUE,
  downsample_factor = 2,
  detect_fixations = TRUE,
  blink_args = list(
    min_duration = 30,
    include_rapid_changes = FALSE
  ),
  fixation_args = list(
    vmax = 5,
    min_duration = 60
  )
)
```

## Decision log

Every requested operation is represented explicitly.

```{r decision-log}
result$decision_log
```

The log records whether an operation was applied or skipped, its input and
output row counts, and a compact description of the resolved action.

## Processed data

```{r processed-data}
names(result$data)
utils::head(
  result$data[
    c(
      "USER_ID",
      "trial",
      "TIME",
      "LPupil",
      "RPupil",
      "gp3_pupil_fused",
      "gp3_pupil_fused_blink_interp",
      "pupil_smoothed",
      "FPOGX",
      "FPOGX_smooth"
    )
  ]
)
```

Original columns remain available. Downsampling reduces rows only when
`downsample_factor` is larger than one.

## Blink intervals

```{r blink-events}
result$blinks
result$diagnostics$blink_summary
```

Blink detection is heuristic and should be reviewed against the recording
context, sampling rate, pupil scale, and missing-data pattern.

## Velocity-based fixation events

```{r fixation-events}
utils::head(result$fixations)
result$diagnostics$fixation_summary
```

The fixation table is computed from the full-resolution coordinate series
before optional downsampling.

## Diagnostic overview

```{r diagnostics}
result$diagnostics$overview
result$diagnostics$signal_summary
```

## Visual audit

```{r pupil-plot}
plot(
  signal$TIME,
  signal$LPupil,
  type = "l",
  xlab = "Time",
  ylab = "Pupil",
  main = "Synthetic pupil preprocessing"
)

lines(
  result$data$TIME,
  result$data$pupil_smoothed,
  lwd = 2
)

legend(
  "topright",
  legend = c("Original left pupil", "Processed pupil"),
  lty = 1,
  lwd = c(1, 2),
  bty = "n"
)
```

```{r coordinate-plot}
plot(
  signal$TIME,
  signal$FPOGX,
  type = "l",
  xlab = "Time",
  ylab = "Horizontal gaze coordinate",
  main = "Synthetic coordinate smoothing"
)

lines(
  result$data$TIME,
  result$data$FPOGX_smooth,
  lwd = 2
)

legend(
  "topright",
  legend = c("Original", "Smoothed"),
  lty = 1,
  lwd = c(1, 2),
  bty = "n"
)
```

## Alternative specifications

Use `pupil_mode = "regression"` for the cross-eye regression helper, or
`pupil_mode = "none"` with an explicit `pupil_col` when the input already
contains a selected pupil trace. Override lists expose the underlying helper
settings without changing workflow-managed data, identifier, or output-column
arguments.

Report the operation order, blink criteria, interpolation method and maximum
gap, smoothing windows, downsampling factor, coordinate scale, velocity
threshold, minimum fixation duration, and any sensitivity specifications.
