---
title: "Statistical extensions and plots"
description: "Plot-first examples for the statistical, sequence, scanpath, and model-reporting extension layer."
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = '#>',
  fig.width = 7,
  fig.height = 4.5,
  warning = FALSE,
  message = FALSE
)
```

This article demonstrates the post-1.0.2 statistical-extension layer in `gp3tools`.
The examples use small synthetic data created inside the article.
They are software demonstrations only, not empirical findings.

```{r setup}
library(gp3tools)
```

## Example data

```{r example-data}
set.seed(2026)

aoi_demo <- expand.grid(
  subject_id = paste0('S', sprintf('%02d', 1:6)),
  trial_id = paste0('T', 1:3),
  time_ms = seq(0, 1200, by = 100),
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)

aoi_demo$condition <- ifelse(aoi_demo$subject_id %in% paste0('S', sprintf('%02d', 1:3)), 'control', 'treatment')
aoi_demo$aoi <- sample(c('Claim', 'Evidence', 'CTA', 'Navigation'), nrow(aoi_demo), replace = TRUE)

fix_demo <- data.frame(
  subject_id = 'S01',
  trial_id = 'T1',
  fixation_index = 1:8,
  start_time_ms = seq(0, 1050, by = 150),
  end_time_ms = seq(120, 1170, by = 150),
  x = c(220, 260, 310, 470, 520, 610, 700, 760),
  y = c(180, 210, 240, 260, 330, 360, 390, 420),
  aoi = c('Claim', 'Claim', 'Evidence', 'Evidence', 'CTA', 'CTA', 'Navigation', 'Navigation'),
  stringsAsFactors = FALSE
)

effect_demo <- data.frame(
  time_ms = seq(0, 1200, by = 100),
  estimate = seq(0.18, 0.42, length.out = 13) + sin(seq(0, pi, length.out = 13)) * 0.05
)
effect_demo$lower <- pmax(0, effect_demo$estimate - 0.07)
effect_demo$upper <- pmin(1, effect_demo$estimate + 0.07)

model_demo <- data.frame(
  target_prop = c(0.20, 0.24, 0.28, 0.31, 0.35, 0.39, 0.43, 0.48),
  condition = rep(c('control', 'treatment'), each = 4)
)
model_fit <- stats::lm(target_prop ~ condition, data = model_demo)
```

## AOI timeline

`plot_gazepoint_aoi_timeline()` provides a compact scarf-style view of AOI state over time.

```{r aoi-timeline, fig.alt='Synthetic AOI timeline showing AOI labels across time by subject and trial.'}
plot_gazepoint_aoi_timeline(
  aoi_demo,
  aoi_col = 'aoi',
  time_col = 'time_ms',
  subject_col = 'subject_id',
  trial_col = 'trial_id',
  title = 'Synthetic AOI timeline',
  x_label = 'Time (ms)'
)
```

## Scanpath plot

`plot_gazepoint_scanpath()` connects fixation coordinates in temporal order.

```{r scanpath-plot, fig.alt='Synthetic scanpath plot connecting fixation points in sequence.'}
plot_gazepoint_scanpath(
  fix_demo,
  x_col = 'x',
  y_col = 'y',
  group_cols = c('subject_id', 'trial_id'),
  time_col = 'start_time_ms',
  fixation_index_col = 'fixation_index',
  reverse_y = TRUE,
  title = 'Synthetic scanpath example'
)
```

## Time-varying effect plot

`plot_gazepoint_time_varying_effect()` visualises a time-varying estimate and interval.

```{r time-varying-effect, fig.alt='Synthetic time-varying estimate with lower and upper interval limits.'}
plot_gazepoint_time_varying_effect(
  effect_demo,
  time_col = 'time_ms',
  estimate_col = 'estimate',
  lower_col = 'lower',
  upper_col = 'upper',
  title = 'Synthetic time-varying effect',
  x_label = 'Time (ms)',
  y_label = 'Estimate'
)
```

## Model residual diagnostics

`plot_gazepoint_model_residuals()` gives a compact residual diagnostic for fitted models.

```{r residual-plot, fig.alt='Synthetic residual diagnostic plot for a simple fitted model.'}
plot_gazepoint_model_residuals(
  model = model_fit,
  title = 'Synthetic residual diagnostic'
)
```

## Sequence and scanpath summaries

The extension layer also includes lightweight tabular sequence summaries.

```{r sequence-summaries}
compute_gazepoint_aoi_entropy(
  aoi_demo,
  aoi_col = 'aoi',
  group_cols = c('subject_id', 'trial_id'),
  time_col = 'time_ms',
  collapse_repeats = TRUE
) |> head()

compute_gazepoint_aoi_sequence_metrics(
  aoi_demo,
  aoi_col = 'aoi',
  group_cols = c('subject_id', 'trial_id'),
  time_col = 'time_ms'
) |> head()

compute_gazepoint_sequence_distance(
  c('Claim', 'Evidence', 'CTA'),
  c('Claim', 'CTA', 'Evidence')
)

compute_gazepoint_scanpath_similarity(
  aoi_demo[aoi_demo$subject_id %in% c('S01', 'S02') & aoi_demo$trial_id == 'T1', ],
  aoi_col = 'aoi',
  group_cols = 'subject_id',
  time_col = 'time_ms',
  collapse_repeats = TRUE
)
```

## Other extension helpers

Additional helpers support bootstrap time-course summaries, saccade metrics, Markov and semi-Markov summaries, simulation, BIDS-style export, Bayesian-model dry runs, recurrence summaries, transition-network metrics, and optional dashboard preparation.
