---
title: "Validating Pupil Reconstruction with Artificial Monocular Loss"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validating Pupil Reconstruction with Artificial Monocular Loss}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4.5
)
library(gp3tools)
```

## Purpose

Cross-eye pupil reconstruction has a useful property that ordinary missing-data problems
often lack: bilateral samples provide observable targets that can be hidden deliberately.
This article uses that fact to test a reconstruction policy before it is applied to
naturally missing eye measurements.

Artificial monocular loss does not prove that naturally missing values behave in the same
way. It does, however, directly answer a narrower and important question: **when one eye
is known, how accurately does the declared calibration reconstruct the other eye in this
dataset?**

## Build a deterministic synthetic benchmark

```{r data}
dat <- simulate_gazepoint_pupil_data(
  n_subjects = 8,
  n_trials = 4,
  n_time_bins = 120,
  blink_probability = 0.01,
  seed = 4102
)

# Give the right eye a reproducible affine offset.
dat$pupil_right <- 0.06 + 0.985 * dat$pupil_right
```

## Random held-out samples

Random masking tests local prediction when individual bilateral observations are hidden.

```{r random-validation}
random_validation <- validate_gazepoint_binocular_reconstruction(
  dat,
  left_col = "pupil_left",
  right_col = "pupil_right",
  time_col = "timestamp_ms",
  group_cols = "subject",
  gap_group_cols = c("subject", "trial"),
  mask_prop = 0.10,
  mask_mode = "random",
  repeats = 5,
  seed = 2026,
  min_pairs = 80
)

random_validation$summary
```

```{r random-plot}
plot_gazepoint_binocular_diagnostics(random_validation, "validation")
```

## Contiguous held-out runs

Natural eye loss is often temporally clustered. Contiguous masking therefore asks a more
demanding question than hiding independent samples.

```{r contiguous-validation}
contiguous_validation <- validate_gazepoint_binocular_reconstruction(
  dat,
  left_col = "pupil_left",
  right_col = "pupil_right",
  time_col = "timestamp_ms",
  group_cols = "subject",
  gap_group_cols = c("subject", "trial"),
  mask_prop = 0.10,
  mask_mode = "contiguous",
  block_size = 8,
  repeats = 5,
  seed = 2027,
  min_pairs = 80
)

contiguous_validation$summary
```

The row-level prediction table retains direction, model ID, calibration level, gap
duration, extrapolation status, and reconstruction outcome.

```{r prediction-table}
head(contiguous_validation$predictions)
```

```{r residuals}
plot_gazepoint_binocular_diagnostics(contiguous_validation, "residuals")
```

## Stress-test increasing monocular loss

A single masking percentage can conceal a deterioration in calibration support. The
stress-test helper repeats validation across a declared range of missingness.

```{r stress}
stress <- stress_test_gazepoint_binocular_reconstruction(
  dat,
  left_col = "pupil_left",
  right_col = "pupil_right",
  time_col = "timestamp_ms",
  group_cols = "subject",
  gap_group_cols = c("subject", "trial"),
  missingness = c(0.05, 0.10, 0.20, 0.30),
  mask_modes = c("random", "contiguous"),
  block_size = 8,
  repeats = 3,
  seed = 2028,
  min_pairs = 80
)

stress$results
```

Useful quantities include both error and **prediction rate**. A low RMSE among the few
rows that remain eligible is not enough if the declared gates prevent reconstruction of
most requested samples.

## Evaluate a gap-duration policy directly

The cross-eye model uses the contralateral eye at the same time point, so a gap-duration
restriction is a governance choice rather than a mathematical requirement of linear
regression. If a study chooses such a restriction, it can be tested explicitly.

```{r gap-policy}
short_gap_validation <- validate_gazepoint_binocular_reconstruction(
  dat,
  left_col = "pupil_left",
  right_col = "pupil_right",
  time_col = "timestamp_ms",
  group_cols = "subject",
  gap_group_cols = c("subject", "trial"),
  mask_prop = 0.10,
  mask_mode = "contiguous",
  block_size = 8,
  repeats = 3,
  seed = 2029,
  min_pairs = 80,
  max_gap_ms = 100
)

short_gap_validation$summary
```

Compare the prediction rate and error with the unrestricted run. This separates the cost
of a strict eligibility policy from the error of the regression model itself.

## Compare directions

Separate left-from-right and right-from-left models are retained because one direction
can be better calibrated than the other.

```{r directions}
left_only_validation <- validate_gazepoint_binocular_reconstruction(
  dat,
  "pupil_left", "pupil_right",
  time_col = "timestamp_ms",
  group_cols = "subject",
  gap_group_cols = c("subject", "trial"),
  direction = "left_from_right",
  mask_prop = 0.10,
  repeats = 3,
  seed = 2030,
  min_pairs = 80
)

right_only_validation <- validate_gazepoint_binocular_reconstruction(
  dat,
  "pupil_left", "pupil_right",
  time_col = "timestamp_ms",
  group_cols = "subject",
  gap_group_cols = c("subject", "trial"),
  direction = "right_from_left",
  mask_prop = 0.10,
  repeats = 3,
  seed = 2031,
  min_pairs = 80
)

rbind(left_only_validation$summary, right_only_validation$summary)
```

## What to report

At minimum report:

- calibration level (for example participant, participant-session, or pooled);
- minimum paired observations and any model-quality gate;
- whether a pooled fallback was allowed;
- whether extrapolation was permitted;
- maximum reconstructed missing-eye run, if restricted;
- artificial-missingness design (random or contiguous, proportion, repeats, seed);
- prediction rate as well as RMSE/MAE/bias;
- final reconstruction burden in the analysed data;
- whether reconstruction rates differed across experimental groups;
- whether downstream summaries were stable under alternative pupil-construction
  policies.

The reporting helper can combine validation with the realised reconstruction audit.

```{r reporting, eval=FALSE}
report <- summarise_gazepoint_binocular_reporting(
  reconstructed,
  audit = audit,
  validation = contiguous_validation
)
cat(report$text)
```

## Interpretation boundary

Artificial monocular loss is a validation device, not a missingness model. It evaluates
prediction where ground truth is observed and then hidden. It cannot establish what an eye
would have measured during naturally occurring loss, nor can it justify reconstruction
through blinks, long bilateral loss, or other intervals where the contralateral signal is
also unavailable or contaminated.
